---
title: "n2 AB008: DEG and gene analysis"
author: "Elisa Balmas PhD"
date: "2025-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment

```{r}
library(tidyr)
library(stringr)
library(grid)
library(viridis)
library(ggthemes)
library(dplyr)
library(tidyverse)
library(ggsignif)
library(umap)
library(heatmap3)
library(plyr)
library(edgeR)
library(compareGroups)
library(dbscan)
library(reshape2)
library(scran)
library(Matrix)
library(fgsea)
library(msigdbr)
library(RNAseqQC)
library(DESeq2)
library(ensembldb)
library(tibble)
library(purrr)
library(magrittr)
library(ggplotify)
library(limma)
library(Glimma)
library(topGO)
library(org.Hs.eg.db)
#library(org.Mm.eg.db)
library(ggrepel)
library(igraph)
library(readxl)
library(clusterProfiler)
library(enrichplot)
library(SeuratObject)
library(Seurat)
library(pathview)
library(pheatmap)
library(ggplot2)
library(reshape2)
library(WGCNA)
library(TCseq)
library(readr)
library(ADImpute)


library(ggplot2); library(reshape2); theme_set(theme_bw(26) + theme(panel.grid.major = element_blank(), 
                                                                    panel.grid.minor = element_blank()) +
                                                 theme(legend.key = element_blank()))
update_geom_defaults("point", aes(size = 4))

#setwd("/home/ebalmas/scpHUB/projects/H001AS7_iKDseq_10X_4tp_monocle/analysis_EB")

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))


# Set up filename prefixes and output folders with the data
dir.create("Output")
#dir.create(file.path("Output","Fig5F_input_folder"))


dir.create(file.path("Output","deseq2"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "dotplot"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "gene_expression"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "gene_expression","GO"))
dir.create(file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "gvolcano"))


fname_Fig5F_input_folder <- file.path("Fig5F_input_folder")

#set up folder path QC

fname_prefix_dotplot_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "dotplot", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_volcano_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "volcano", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_R_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_csv_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_gene_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "gene_expression", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_gene_GO_deseq2 <- file.path("Output","deseq2", format(Sys.Date(), "%y%m%d"), "gene_expression", "GO",
                               format(Sys.Date(), "%y%m%d"))


#okabe_tab <- read_excel("Fig5F_input_folder/color_scale.xlsx")
#okabe_pal=okabe_tab$hex

okabe_pal=c("#E69F00","#56B4E9","#009E73","#f0E442","#0072B2","#D55E00","#CC79A7","#000000")


pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
         "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
         "#920000","#924900","#db6d00","#24ff24","#ffff6d")
```

########load data

```{r}
#load(paste0(getwd(),"/Fig5F_input_folder/dds_PCA.RData"))
load(paste0(getwd(),"/Fig5F_input_folder/260213_dds_PCA.RData"))

#dds <- dds[ !rownames(dds) %in% c("COM_K", "dCAS9", "MCP_PH_TagBFP", "OPTetR"), ]
#rownames(dds)[rownames(dds) %in% c("COM_K", "dCAS9", "MCP_PH_TagBFP", "OPTetR")]
```

####change PCA appeareance

```{r}

okabe_named <- c(
  "NTC"    = "#000000",
  "hiPSC"  = "#56B4E9",
  "CIRI"   = "#CC79A7",
  "OPTIOX" = "#E69F00"
)

p=plot_pca_scatters(vsd, n_PCs = 2,  color_by = "name", shape_by = "replicate",show_var_exp=T)#batch 

p2 <- p +
  scale_shape_manual(
    name = "Replicate",
    values = c(
      "R1" = 16,  # filled circle
      "R2" = 17,  # filled triangle (up)
      "R3" = 15,  # filled square
      "R4" = 25   # filled triangle (down)
    )
  ) +
  scale_color_manual(
    name = "Group",
    values = okabe_named
  ) +
  labs(
    x = paste0("PC1","(", pca_variability[1,1], "%)"),
    y = paste0("PC2","(", pca_variability[2,1], "%)"),
    title = "PCA plot"
  )+
  geom_point(size = 3)

p2

ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_PCA.pdf"),
       plot = p2, width = 5, height = 4)
```



####process PCA to add gene names to loadings and scale arrows

```{r}

# Ensure gene names exist
if (!"gene_name" %in% colnames(PCA_loadings)) {
  PCA_loadings$gene_name <- rownames(PCA_loadings)
}

# Select top genes by contribution
n_arrows <- 25
loadings2 <- PCA_loadings %>%
  mutate(contrib = PC1^2 + PC2^2) %>%
  slice_max(contrib, n = n_arrows)

# Scale loadings so arrows fit in score space
arrow_scale <- max(abs(c(PCAdata$PC1, PCAdata$PC2)), na.rm = TRUE) * 0.9
loadings2 <- loadings2 %>%
  mutate(PC1s = PC1 * arrow_scale,
         PC2s = PC2 * arrow_scale)

# Axis labels with explained variance
pc1_lab <- paste0("PC1 (", scales::percent(pca_res$var_exp[1], accuracy = 0.1), ")")
pc2_lab <- paste0("PC2 (", scales::percent(pca_res$var_exp[2], accuracy = 0.1), ")")

# === Biplot ===
p_biplot <- ggplot() +
  geom_point(data = PCAdata,
             aes(x = PC1, y = PC2, color = name, shape = replicate),
             size = 3, alpha = 0.9) +
  geom_segment(data = loadings2,
               aes(x = 0, y = 0, xend = PC1s, yend = PC2s),
               arrow = arrow(length = unit(0.02, "npc")), linewidth = 0.3) +
  geom_text_repel(data = loadings2,
                  aes(x = PC1s, y = PC2s, label = gene_name),
                  size = 3, max.overlaps = Inf, min.segment.length = 0) +
  labs(x = pc1_lab, y = pc2_lab, title = "PCA biplot (VST, DESeq2)") +
  theme_classic() +
  theme(plot.title = element_text(face = "bold"))

print(p_biplot)

ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_PCA_biplot_loadings_DESeq2.pdf"),
       plot = p_biplot, width = 6.5, height = 5)

ggplot(PCA_loadings, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.4) +
  geom_text_repel(data = PCA_loadings %>% slice_max(PC1^2 + PC2^2, n = 25),
                  aes(label = gene_name), size = 3) +
  theme_classic() +
  labs(title = "Gene loadings on PC1 and PC2",
       x = "PC1 loading", y = "PC2 loading")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_PCA_loadings_dotplot.pdf"),
       width = 6.5, height = 5)


```
####PCA loading analysis top 50 per PC

```{r}
top_pc1_pos <- PCA_loadings %>% arrange(desc(PC1)) %>% slice_head(n = 50)
top_pc1_neg <- PCA_loadings %>% arrange(PC1) %>% slice_head(n = 50)

top_pc2_pos <- PCA_loadings %>% arrange(desc(PC2)) %>% slice_head(n = 50)
top_pc2_neg <- PCA_loadings %>% arrange(PC2) %>% slice_head(n = 50)

gene_ranks_pc1 <- PCA_loadings$PC1###pluripotency
names(gene_ranks_pc1) <- PCA_loadings$gene_name

gene_ranks_pc2 <- PCA_loadings$PC2###differentiation in muscle
names(gene_ranks_pc2) <- PCA_loadings$gene_name

###unranked and on top 50
top_pc1_pos_entrez <- bitr(top_pc1_pos$gene_name, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
top_pc1_neg_entrez <- bitr(top_pc1_neg$gene_name, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
top_pc2_pos_entrez <- bitr(top_pc2_pos$gene_name, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)
top_pc2_neg_entrez <- bitr(top_pc2_neg$gene_name, fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db)

###entrez lists on ranked lists

gene_ranks_pc1_entrez <- bitr(names(gene_ranks_pc1), fromType="SYMBOL",
                              toType="ENTREZID", OrgDb=org.Hs.eg.db)
gene_ranks_pc1 <- gene_ranks_pc1[match(gene_ranks_pc1_entrez$SYMBOL, names(gene_ranks_pc1))]
names(gene_ranks_pc1) <- gene_ranks_pc1_entrez$ENTREZID

gene_ranks_pc2_entrez <- bitr(names(gene_ranks_pc2), fromType="SYMBOL",
                              toType="ENTREZID", OrgDb=org.Hs.eg.db)
gene_ranks_pc2 <- gene_ranks_pc2[match(gene_ranks_pc2_entrez$SYMBOL, names(gene_ranks_pc2))]
names(gene_ranks_pc2) <- gene_ranks_pc2_entrez$ENTREZID

```

####run GO enrichment test on top 50 ###PC1 #pliripotency

```{r}
# Run enrichment tests (GO or KEGG)
ego_pc1_pos <- enrichGO(gene = top_pc1_pos_entrez$ENTREZID,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        qvalueCutoff = 0.05,
                        readable = TRUE)

ego_pc1_neg <- enrichGO(gene = top_pc1_neg_entrez$ENTREZID,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        qvalueCutoff = 0.05,
                        readable = TRUE)

#  Visualize results
library(enrichplot)

dotplot(ego_pc1_pos,showCategory = 15, title = "Top PC1 Positive Loadings (iPSC) – GO BP")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GO_PC1pos.pdf"),
     width = 10, height = 12)
dotplot(ego_pc1_neg,showCategory = 15, title = "Top PC1 Negative Loadings (muscle) – GO BP")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GO_PC1neg.pdf"),
     width = 10, height = 12)

# For KEGG instead:
ekegg_pc1_pos <- enrichKEGG(gene = top_pc1_pos_entrez$ENTREZID, organism = "hsa")
ekegg_pc1_neg <- enrichKEGG(gene = top_pc1_neg_entrez$ENTREZID, organism = "hsa")

dotplot(ekegg_pc1_pos, showCategory = 15, title = "KEGG enrichment (PC1 positive)")
```
####run GO enrichment test on top 50 ###PC2 #muscle, but less clear

```{r}
# Run enrichment tests (GO or KEGG)
ego_pc2_pos <- enrichGO(gene = top_pc2_pos_entrez$ENTREZID,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        qvalueCutoff = 0.05,
                        readable = TRUE)

ego_pc2_neg <- enrichGO(gene = top_pc1_neg_entrez$ENTREZID,
                        OrgDb = org.Hs.eg.db,
                        ont = "BP",
                        pAdjustMethod = "BH",
                        qvalueCutoff = 0.05,
                        readable = TRUE)

#  Visualize results
library(enrichplot)

dotplot(ego_pc2_pos, showCategory = 15, title = "Top PC2 Positive Loadings (OPTIOX)– GO BP")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GO_PC2pos.pdf"),
     width = 10, height = 12)
dotplot(ego_pc2_neg, showCategory = 15, title = "Top PC2 Negative Loadings (CIRI-hiPSC)– GO BP")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GO_PC2neg.pdf"),
     width = 10, height = 12)
# For KEGG instead:
ekegg_pc2_pos <- enrichKEGG(gene = top_pc2_pos_entrez$ENTREZID, organism = "hsa")
ekegg_pc2_neg <- enrichKEGG(gene = top_pc2_neg_entrez$ENTREZID, organism = "hsa")

dotplot(ekegg_pc2_pos, showCategory = 15, title = "KEGG enrichment (PC2 positive)")
```
####GSEA on PC1

```{r}

# Rank all genes by PC2 loading
gene_ranks_pc1 <- PCA_loadings$PC1
names(gene_ranks_pc1) <- PCA_loadings$gene_name

# Convert gene symbols to ENTREZ IDs
gene_ranks_pc1_entrez <- bitr(names(gene_ranks_pc1), fromType="SYMBOL",
                              toType="ENTREZID", OrgDb=org.Hs.eg.db)
gene_ranks_pc1 <- gene_ranks_pc1[match(gene_ranks_pc1_entrez$SYMBOL, names(gene_ranks_pc1))]
names(gene_ranks_pc1) <- gene_ranks_pc1_entrez$ENTREZID

# Run GSEA
gsea_pc1 <- gseGO(geneList = sort(gene_ranks_pc1, decreasing = TRUE),
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  minGSSize = 10,
                  pvalueCutoff = 0.05,
                  verbose = FALSE)

#ridgeplot(gsea_pc2, showCategory = 15)
dotplot(gsea_pc1, color = "NES",split =".sign",showCategory = 20, decreasing = T,title = "GSEA PC1")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC1.pdf"),
     width = 8, height = 8)

#term_to_plot=c("GO:0030239","GO:0030216")  # if you need to plot multiple terms
#p_top <- gseaplot2(gsea_pc2, geneSetID = 1,pvalue_table = TRUE, base_size = 12)   # 1 = first row in gsea_pc1@result after default ordering
p_top <-gseaplot2(
    gsea_pc1,
    geneSetID = 1,
    #geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = "firebrick",
    pvalue_table = TRUE
  )
p_top
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC1_top_enrichment.pdf"),
     width = 6, height = 4)

gsea_pal=c("#E69F00","blue","black")
term_to_plot=c("GO:0030239","GO:0061061","GO:0055001")
gseaplot2(
    gsea_pc1,
    #geneSetID = 1,
    geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = gsea_pal[seq_along(term_to_plot)],
    pvalue_table = TRUE
  )

ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC1_iPSC_muscle.pdf"),
     width = 7, height = 5)

gsea_pal=c("#E69F00","blue","black")
term_to_plot=c("GO:0055002"#top
               ,"GO:0060537"#striated muscle tissue development
               ,"GO:0007369"#bottom 
               )
gseaplot2(
    gsea_pc1,
    #geneSetID = 1,
    geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = gsea_pal[seq_along(term_to_plot)],
    pvalue_table = TRUE
  )
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC1_top_bottom_enrichment.pdf"),
     width = 7, height = 5)
```

####GSEA on PC2

```{r}

# Rank all genes by PC2 loading
gene_ranks_pc2 <- PCA_loadings$PC2
names(gene_ranks_pc2) <- PCA_loadings$gene_name

# Convert gene symbols to ENTREZ IDs
gene_ranks_pc2_entrez <- bitr(names(gene_ranks_pc2), fromType="SYMBOL",
                              toType="ENTREZID", OrgDb=org.Hs.eg.db)
gene_ranks_pc2 <- gene_ranks_pc1[match(gene_ranks_pc2_entrez$SYMBOL, names(gene_ranks_pc2))]
names(gene_ranks_pc2) <- gene_ranks_pc2_entrez$ENTREZID

# Run GSEA
gsea_pc2 <- gseGO(geneList = sort(gene_ranks_pc2, decreasing = TRUE),
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  minGSSize = 10,
                  pvalueCutoff = 0.05,
                  verbose = FALSE)

#ridgeplot(gsea_pc2, showCategory = 15)
dotplot(gsea_pc2, color = "NES",split =".sign",showCategory = 20, decreasing = T,title = "GSEA PC2")
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC2.pdf"),
     width = 8, height = 8)

#term_to_plot=c("GO:0030239","GO:0030216")  # if you need to plot multiple terms
#p_top <- gseaplot2(gsea_pc2, geneSetID = 1,pvalue_table = TRUE, base_size = 12)   # 1 = first row in gsea_pc1@result after default ordering
p_top <-gseaplot2(
    gsea_pc2,
    geneSetID = 1,
    #geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = "firebrick",
    pvalue_table = TRUE
  )
p_top
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC2_top_enrichment.pdf"),
     width = 6, height = 4)

gsea_pal=c("#E69F00","blue","black")
term_to_plot=c("GO:0030239","GO:0061061","GO:0055001")
gseaplot2(
    gsea_pc2,
    #geneSetID = 1,
    geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = gsea_pal[seq_along(term_to_plot)],
    pvalue_table = TRUE
  )

ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC2_muscle.pdf"),
     width = 7, height = 5)

gsea_pal=c("#E69F00","blue","black")
term_to_plot=c("GO:0003012"#top
               ,"GO:0006936"#
               ,"GO:0003015"#bottom 
               )
gseaplot2(
    gsea_pc2,
    #geneSetID = 1,
    geneSetID = term_to_plot,
    #title = gsea_pc2@result$Description[gsea_pc2@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = gsea_pal[seq_along(term_to_plot)],
    pvalue_table = TRUE
  )
ggsave(filename = paste0(fname_prefix_dotplot_deseq2, "_", "GSEA_PC2_top_bottom_enrichment.pdf"),
     width = 7, height = 5)
```


####make plots for specific genes by Deseq2 data normalized

```{r}
okabe_named <- c(
  "NTC"    = "#000000",
  "hiPSC"  = "#56B4E9",
  "CIRI"   = "#CC79A7",
  "OPTIOX" = "#E69F00"
)

geneInfo_f <- geneInfo_f %>%
  mutate(gene_name = ifelse(
    gene_ID == "MYOD1",          # the custom one you added
    "MYOD1exo",                  # rename it
    gene_name                    # keep all others unchanged
  ))

genes_to_plot=c("MYOG","TTN","IGF2","DLK1","MYL2","POU5F1","NANOG","MYOD1","MYOC","NR3C1","NR3C2","dCAS9","COM_K", "PAX7","DLK1","IGF2","MYH3","MYL4","DES","HOXB1","HOXB9","NOXO1","RYR1","CACNA1S","STAC3","SCN4A","PAX3","MYF5","MEF2C","MEF2D")
g_info_plot=geneInfo_f[ geneInfo_f$gene_name %in% genes_to_plot , c("gene_ID","gene_name") ]
genes_to_plot_ensb=g_info_plot$gene_ID

#dd <- plotCounts(dds, gene=c("ENSG00000122180"), intgroup=c("condition","name","replicate","name_R"), returnData=TRUE)
#gene_symbol <- geneInfo_f$gene_name[ geneInfo_f$gene_ID == "ENSG00000122180" ]
#colnames(dd)[ colnames(dd) == "count" ] <- gene_symbol

intgroup      <- c("condition","name","replicate","name_R","cell_type")

# helper: match Ensembl (with or w/o version) to row index in dds
ens_dds_raw    <- rownames(dds)
ens_dds_novers <- sub("\\..*$","", ens_dds_raw)
get_idx <- function(ens_id) {
  key <- sub("\\..*$","", ens_id)
  idx <- which(ens_dds_novers == key)
  if (length(idx) == 0) return(integer(0))
  if (length(idx) == 1) return(idx)
  # if multiple rows, pick the most expressed by total normalized counts
  nc <- counts(dds, normalized=TRUE)[idx, , drop=FALSE]
  idx[ which.max(rowSums(nc, na.rm=TRUE)) ]
}

# build one small dd per gene, keep only keys + this gene's counts (renamed)
dd_list <- lapply(seq_len(nrow(g_info_plot)), function(i) {
  ens  <- g_info_plot$gene_ID[i]
  sym  <- g_info_plot$gene_name[i]
  idx  <- get_idx(ens)
  if (length(idx) == 0) return(NULL)  # silently skip if not found in dds
  dd_i <- plotCounts(dds, gene = idx, intgroup = intgroup, normalized=TRUE
,returnData = TRUE)
  colnames(dd_i)[ colnames(dd_i) == "count" ] <- sym
  # keep only join keys + this gene column
  dd_i[, c(intgroup, sym), drop=FALSE]
})

# drop missing
dd_list <- Filter(Negate(is.null), dd_list)
if (length(dd_list) == 0) stop("None of the requested genes were found in 'dds'.")

# merge all by the requested columns
dd_merged <- Reduce(function(x, y) merge(x, y, by = intgroup, all = TRUE), dd_list)

#dd_merged=dd_merged[dd_merged$name!="hiPSC", ]


ggplot(dd_merged, aes(x = name, y = MYOG)) +
  # Boxplot: same box per cell type (no grouping by condition)
  geom_boxplot(outlier.shape = NA, alpha = 0.3, color = "gray40") +
  # Jitter: condition affects shape and color
  geom_jitter(aes(color = name), width = 0.2, size = 2) +
  scale_color_manual(name = "Group", values = okabe_named, drop = FALSE) +
  geom_signif(comparisons = list(c("CIRI", "OPTIOX")),
              map_signif_level = TRUE, test = "wilcox.test",
  test.args = list(exact = FALSE, correct = FALSE)) +
  #geom_signif(comparisons = list(c("hiPSC", "NTC")),  
   #           map_signif_level = TRUE, test = "wilcox.test",
  #test.args = list(exact = FALSE, correct = FALSE)) +
  scale_y_continuous(trans = "log1p") +
  labs(
    title = "MYOG expression by cell type and condition",
    y = "DESeq2 normalized counts (log1p)",
    x = ""
  ) +
  theme_bw()
ggsave(filename = paste0(fname_prefix_gene_deseq2, "_MYOG.pdf"),
       width = 6.5, height = 5)

# Loop through each gene
for (gene in genes_to_plot) {

  # Skip if gene column not present
  if (!(gene %in% colnames(dd_merged))) {
    message("Skipping ", gene, " (not found in dd_merged)")
    next
  }

  # Build the plot
  p <- ggplot(dd_merged, aes(x = name, y = .data[[gene]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3, color = "gray40") +
    geom_jitter(aes(color = name), width = 0.2, size = 2) +
    scale_color_manual(name = "Group", values = okabe_named, drop = FALSE) +
      geom_signif(comparisons = list(c("CIRI", "OPTIOX")),
              map_signif_level = TRUE, test = "wilcox.test",
  test.args = list(exact = FALSE, correct = FALSE)) +
  #geom_signif(comparisons = list(c("hiPSC", "NTC")),
   #           map_signif_level = TRUE, test = "wilcox.test",
  #test.args = list(exact = FALSE, correct = FALSE)) +
    scale_y_continuous(trans = "log1p") +
    labs(
      title = paste0(gene, " expression by cell type and condition"),
      y = "DESeq2 normalized counts (log1p)",
      x = ""
    ) +
    theme_bw()

  # Save each plot
  ggsave(
    filename = paste0(fname_prefix_gene_deseq2, "_", gene, ".pdf"),
    plot = p,
    width = 6.5, height = 5
  )

  message("Saved plot for ", gene)
}

```

####make plots for specific genes by TPM data normalized

```{r}

geneInfo_f <- geneInfo_f %>%
  mutate(gene_name = ifelse(
    gene_ID == "MYOD1",          # the custom one you added
    "MYOD1exo",                  # rename it
    gene_name                    # keep all others unchanged
  ))

genes_to_plot=c("MYOG","TTN","IGF2","DLK1","MYL2","POU5F1","NANOG","MYOD1","MYOC","NR3C1","NR3C2","dCAS9","COM_K", "PAX7","DLK1","IGF2","MYH3","MYL4","DES","HOXB1","HOXB9","NOXO1","RYR1","CACNA1S","STAC3","SCN4A","PAX3","MYF5","MEF2C","MEF2D")
g_info_plot=geneInfo_f[ geneInfo_f$gene_name %in% genes_to_plot , c("gene_ID","gene_name") ]
genes_to_plot_ensb=g_info_plot$gene_ID

filter_tpm=counts_tpm[ rownames(counts_tpm) %in% genes_to_plot_ensb , ]
filter_tpm$ensb=rownames(filter_tpm)
filter_tpm=filter_tpm%>%
  left_join(g_info_plot, by=c("ensb"="gene_ID"))
rownames(filter_tpm)=filter_tpm$gene_name
filter_tpm=filter_tpm%>%
  select(-ensb, -gene_name)

m=as.data.frame(t(filter_tpm))
m$sample_ID=rownames(m)
m=m%>%
  left_join(meta, by="sample_ID")
rownames(m)=m$sample_ID

#m=m[ m$name!="hiPSC", ]#take off the hiPSC

# Loop through each gene
for (gene in genes_to_plot) {

  # Skip if gene column not present
  if (!(gene %in% colnames(m))) {
    message("Skipping ", gene, " (not found in m)")
    next
  }

  # Build the plot
  p <- ggplot(m, aes(x = cell_type, y = .data[[gene]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3, color = "gray40") +
    geom_jitter(aes(shape = condition, color = name), width = 0.2, size = 2) +
      scale_color_manual(name = "Group", values = okabe_named, drop = FALSE) +
  geom_signif(comparisons = list(c("CIRI", "OPTIOX")),
              map_signif_level = TRUE, test = "wilcox.test",
  test.args = list(exact = FALSE, correct = FALSE)) +
 # geom_signif(comparisons = list(c("hiPSC", "NTC")),  
  #            map_signif_level = TRUE, test = "wilcox.test") +
    scale_y_continuous(trans = "log1p") +
    labs(
      title = paste0(gene, " expression by cell type and condition"),
      y = "TPM normalized counts (log1p)",
      x = ""
    ) +
    theme_bw()

  # Save each plot
  ggsave(
    filename = paste0(fname_prefix_gene_deseq2, "_", gene, "_TPM.pdf"),
    plot = p,
    width = 6.5, height = 5
  )

  message("Saved plot for ", gene)
}

external_exp=m

#external_exp=external_exp%>%
 # filter(cell_type=="GERALT")

```

####make plots for specific genes by TPM data normalized
####polot genes from DEG analysis

```{r}

geneInfo_f <- geneInfo_f %>%
  mutate(gene_name = ifelse(
    gene_ID == "MYOD1",          # the custom one you added
    "MYOD1exo",                  # rename it
    gene_name                    # keep all others unchanged
  ))

genes_to_plot=c("MYOG","TTN","IGF2","DLK1","MYL2","POU5F1","NANOG","MYOD1","MYOC","NR3C1","NR3C2","dCAS9","COM_K", "PAX7","DLK1","IGF2","MYH3","MYL4","DES","HOXB1","HOXB9","NOXO1","RYR1","CACNA1S","STAC3","SCN4A","PAX3","MYF5","MEF2C","MEF2D")
g_info_plot=geneInfo_f[ geneInfo_f$gene_name %in% genes_to_plot , c("gene_ID","gene_name") ]
genes_to_plot_ensb=g_info_plot$gene_ID

filter_tpm=counts_tpm[ rownames(counts_tpm) %in% genes_to_plot_ensb , ]
filter_tpm$ensb=rownames(filter_tpm)
filter_tpm=filter_tpm%>%
  left_join(g_info_plot, by=c("ensb"="gene_ID"))
rownames(filter_tpm)=filter_tpm$gene_name
filter_tpm=filter_tpm%>%
  select(-ensb, -gene_name)

m=as.data.frame(t(filter_tpm))
m$sample_ID=rownames(m)
m=m%>%
  left_join(meta, by="sample_ID")
rownames(m)=m$sample_ID

# keep only CIRI + OPTIOX
m <- m %>%
  dplyr::filter(name %in% c("CIRI", "OPTIOX")) %>%
  droplevels()

rownames(m) <- m$sample_ID

#m=m[ m$name!="hiPSC", ]#take off the hiPSC

# Loop through each gene
for (gene in genes_to_plot) {

  # Skip if gene column not present
  if (!(gene %in% colnames(m))) {
    message("Skipping ", gene, " (not found in m)")
    next
  }

  # Build the plot
  p <- ggplot(m, aes(x = name, y = .data[[gene]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3, color = "gray40") +
    geom_jitter(aes(shape = condition, color = name), width = 0.2, size = 2) +
      scale_color_manual(name = "Group", values = okabe_named, drop = FALSE) +
  geom_signif(comparisons = list(c("CIRI", "OPTIOX")),
              map_signif_level = TRUE, test = "wilcox.test",
  test.args = list(exact = FALSE, correct = FALSE)) +
    scale_y_continuous(trans = "log1p") +
    labs(
      title = paste0(gene, " expression by cell type and condition"),
      y = "TPM normalized counts (ln(TPM + 1))",
      x = ""
    ) +
    theme_bw()

  # Save each plot
  ggsave(
    filename = paste0(fname_prefix_gene_deseq2, "_", gene, "_spec_TPM.pdf"),
    plot = p,
    width = 6.5, height = 5
  )

  message("Saved plot for ", gene)
}

```


####make plot of NANOG expression comparing the 3 configurations

```{r}

genes_to_plot=c("MYOG","TTN","IGF2","DLK1","MYL2","POU5F1","NANOG","NANOGxt","MYOD1","MYOC","MYOD1exo","NR3C1","NR3C2")

# Loop through each gene
for (gene in genes_to_plot) {

  # Skip if gene column not present
  if (!(gene %in% colnames(external_exp))) {
    message("Skipping ", gene, " (not found in external_exp)")
    next
  }

  # Build the plot
  p <- ggplot(external_exp, aes(x = name, y = .data[[gene]])) +
    geom_boxplot(outlier.shape = NA, alpha = 0.3, color = "gray40") +
    geom_jitter(aes(shape = name, color = name), width = 0.2, size = 2) +
      geom_signif(comparisons = list(c("TT2MNi", "TT2MNb")),  # replace with your levels
              map_signif_level = TRUE,
              test = "t.test")+
    scale_y_continuous(trans = "log1p") +
    labs(
      title = paste0(gene, " expression by cell type and condition"),
      y = "TPM normalized counts (ln(TPM + 1))",
      x = ""
    ) +
    theme_bw()

  # Save each plot
  ggsave(
    filename = paste0(fname_prefix_gene_deseq2, "_", gene, "_TPM_GERALT_ivsb.pdf"),
    plot = p,
    width = 6.5, height = 5
  )

  message("Saved plot for ", gene)
}
```


####create Meta data with contrasts for DEG analysis
####already present so silenced the pipeline

```{r}
# Create contrast columns for each condition comparison
#meta <- meta %>%
 # mutate(cfr_GERALT_vs_OPTIOX = ifelse(cell_type == "GERALT", "GERALT",
  #                                   ifelse(cell_type == "OPTIOX"&condition !="hiPSCs", "OPTIOX", "no")),
       #  cfr_TT2MN_vs_TMM = ifelse(name == "TT2MN", "TT2MN",
        #                             ifelse(name == "TMM", "TMM", "no")),
         #cfr_TT2MNb_vs_TMM = ifelse(name == "TT2MNb", "TT2MNb",
        #                             ifelse(name == "TMM", "TMM", "no")),
        # cfr_TT2MNi_vs_TMM = ifelse(name == "TT2MNi", "TT2MNi",
        #                             ifelse(name == "TMM", "TMM", "no")),
        # cfr_TT2MN_vs_TT2MNb = ifelse(name == "TT2MN", "TT2MN",
        #                             ifelse(name == "TT2MNb", "TT2MNb", "no")),
        # cfr_TT2MN_vs_TT2MNi = ifelse(name == "TT2MN", "TT2MN",
         #                            ifelse(name == "TT2MNi", "TT2MNi", "no")),
         #cfr_TT2MNi_vs_TT2MNb = ifelse(name == "TT2MNi", "TT2MNi",
          #                           ifelse(name == "TT2MNb", "TT2MNb", "no")))
```


####DEG analysis between CIRI and OPTIOX and hiPSC vs NTC for each cell type one by one then do a loop

```{r}
#In meta I already placed my contrast columns starting with "cfr_"
#in case cfr_columns do not exist just create them. Place in the comparison the condition name (meta$condition relative to the sample) and the samples not analyzed will have "no" instead.
# Identify contrast columns (those starting with "cfr_")
contrast_cols <- grep("^cfr_", colnames(meta), value = TRUE)
contrast_cols
# Pick one contrast manually for now (first one)
contrast_col <- contrast_cols[1]#cfr_CIRIvsOPTIOX column (need to check the contrast col)
contrast_col

# Prepare metadata for this contrast
meta_cfr <- meta %>%
  dplyr::select(sample_ID, condition, replicate, !!sym(contrast_col)) %>%
  dplyr::rename(cfr = !!sym(contrast_col)) %>%
  dplyr::filter(cfr != "no")

# Subset DESeq2 object to relevant samples
dds_sub <- dds[, colnames(dds) %in% meta_cfr$sample_ID]
dds_sub$group <- factor(meta_cfr$cfr)
#levels(dds_sub$group) #check level of contrast
# (Optional — set whichever condition you want as baseline)
dds_sub$group <- relevel(dds_sub$group, ref = levels(dds_sub$group)[1])
# --- Define new design formula ---
design(dds_sub) <- ~ group
# Run DESeq2
dds_sub <- DESeq(dds_sub)

# Define contrast (2 levels only)
cfr_levels <- levels(dds_sub$group)
res <- results(dds_sub, contrast = c("group", cfr_levels[1], cfr_levels[2]))#CIRI is 1 and OPTIOX is 2

# Clean result table
res_df <- as.data.frame(res)
res_df <- res_df %>%
  rownames_to_column("gene_ID") %>%
  left_join(geneInfo_f, by = "gene_ID") %>%
  arrange(padj)

#res_df=res_df%>%dplyr::filter(!gene_ID %in% c("COM_K", "dCAS9","MCP_PH_TagBFP","OPTetR"))#eliminate gene names not needed
#gene_ID=counts_df$gene_ID

```
####create a table to store all the data on specific comparisons

```{r}
###set the comparisons you want to do

comparisons=c("cfr_CIRIvsOPTIOX",
              "cfr_NTCvshiPSCs")

first_to_compare=c("CIRI","NTC")

###create an empty list to store all the DEG results
deg_results_list <- list()

for (i in seq_along(comparisons)) {
  contrast_col <- comparisons[i]
  ref_level    <- first_to_compare[i]
  message("Processing contrast: ", contrast_col, "  (reference = ", ref_level, ")")

  meta_cfr <- meta %>%
    dplyr::select(sample_ID, condition, replicate, !!sym(contrast_col)) %>%
    dplyr::rename(cfr = !!sym(contrast_col)) %>%
    dplyr::filter(cfr != "no")

  dds_sub <- dds[, colnames(dds) %in% meta_cfr$sample_ID]
  dds_sub$group <- factor(meta_cfr$cfr)

  # relevel to make sure first_to_compare is the reference
  if (ref_level %in% levels(dds_sub$group)) {
    dds_sub$group <- relevel(dds_sub$group, ref = ref_level)
  } else {
    warning("Reference ", ref_level, " not found in ", contrast_col)
  }

  design(dds_sub) <- ~ group
  dds_sub <- DESeq(dds_sub)

  # contrast = "comparison vs baseline"
  cfr_levels <- levels(dds_sub$group)
  res <- results(dds_sub, contrast = c("group", cfr_levels[1], cfr_levels[2]))

  res_df <- as.data.frame(res) %>%
    tibble::rownames_to_column("gene_ID") %>%
    left_join(geneInfo_f, by = "gene_ID") %>%
    arrange(padj)

  deg_results_list[[contrast_col]] <- res_df
}

# --- user-defined thresholds --- add directionality
fc_cutoff <- log2(2)   # change as needed, e.g. log2(2) for 2× fold change
padj_cutoff <- 0.05      # adjust significance threshold

# --- refine and label each DEG table ---
for (contrast_col in comparisons) {
  message("Processing contrast for final table: ", contrast_col)
  
  deg_results_list[[contrast_col]] <- deg_results_list[[contrast_col]] %>%
   # dplyr::filter(!gene_ID %in% c("COM_K", "dCAS9","MCP_PH_TagBFP","OPTetR"))%>%#eliminate gene names not needed
    dplyr::select(gene_ID, gene_name, log2FoldChange, pvalue, padj) %>%
    mutate(
      direction = case_when(
        padj < padj_cutoff & log2FoldChange >  fc_cutoff ~ "Up",
        padj < padj_cutoff & log2FoldChange < -fc_cutoff ~ "Down",
        TRUE ~ "NS"
      )
    ) 
}

deg_results_list_preprocessed=deg_results_list
#make a dataframe with all the results

# --- user-defined thresholds ---
fc_cutoff <- log2(2)   # change as needed, e.g. log2(2) for 2× fold change
padj_cutoff <- 0.05      # adjust significance threshold

# --- refine and label each DEG table ---
for (contrast_col in comparisons) {
  message("Processing contrast for final table: ", contrast_col)
  
  deg_results_list[[contrast_col]] <- deg_results_list[[contrast_col]] %>%
    dplyr::select(gene_ID, gene_name, log2FoldChange, pvalue, padj, direction) %>%
    dplyr::rename(
      !!paste0(contrast_col, "_log2FC") := log2FoldChange,
      !!paste0(contrast_col, "_pvalue") := pvalue,
      !!paste0(contrast_col, "_padj")   := padj,
      !!paste0(contrast_col, "_direction") := direction
    )
}

library(purrr)
# --- merge all contrasts into one big table ---
deg_merged <- purrr::reduce(deg_results_list, full_join, by = c("gene_ID", "gene_name"))
de_stats <- deg_merged

#save as txt and RDS
filename = paste0(fname_prefix_csv_deseq2, "_DEG_all_comparisons_cfr.txt")
write.table(deg_merged, file=filename, sep="\t", quote=F, row.names=F)

Rdata_filename = paste0(fname_prefix_R_deseq2, "_DEG_analysis.RData")
save(deseq_counts,geneInfo_f,counts_f,meta,dds,counts_tpm,deg_results_list,deg_merged,deg_results_list_preprocessed,
     file = Rdata_filename)

```

####make volcano plots in general
#make them custom and one by one to make sure they are nice

```{r}
library(EnhancedVolcano)

###this is from the merged table
#enhancedvolcano is quite nice
#EnhancedVolcano(deg_merged,
              #  lab = deg_merged$gene_name,
               # x = 'cfr_GERALT_vs_OPTIOX_log2FC',
              #  y = 'cfr_GERALT_vs_OPTIOX_padj',
               # pCutoff = 0.05,
              #  FCcutoff = 1.5,
               # title = 'GERALT vs OPTIOX',
              #  subtitle = 'Volcano Plot',
               # caption = 'Cutoff: padj < 0.05 & |log2FC| > 1.5')

#create a name vector for colors
ok_pal <- c("Up" = "#CC79A7", "Down" = "#009E73", "NS" = "grey70")

EnhancedVolcano(deg_results_list_preprocessed$cfr_CIRIvsOPTIOX,
                lab = deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$gene_name,
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = log2(2),
                title = 'CIRI vs OPTIOX',
                subtitle = 'Volcano Plot',
                caption = 'Cutoff: padj < 0.05 & |log2FC| > 1',
                pointSize = 2.0,
                #labSize = 3, 
  ###add colors
  colCustom = ok_pal[deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$direction],
  colAlpha=0.5,
  # --- labeling options ---
  drawConnectors = T,            # add arrow lines to labels
  widthConnectors = 0.1,            # arrow thickness
  colConnectors = 'black',
  # --- label appearance ---
  labSize = 2.5,                    # font size for labels
  boxedLabels = F,               # optional: draw boxes around labels
  max.overlaps = 15,                # control overlap density
  # --- control which genes to label ---
  #selectLab = head(
   # deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$gene_name[
    #  order(deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$padj)
    #],
    #60                             # label top 20 most significant genes
 # ),
  #legendPosition = 'right',
  legendLabSize = 10,
  legendIconSize = 3.5,
  borderWidth = 0.6,
  axisLabSize = 12
  )+
 annotate("text",
           x = -10,  # adjust based on your FC range
           y = max(-log10(deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$padj), na.rm = TRUE) * 0.9,
           label = "OPTIOX",
           hjust = 0, size = 4, fontface = "bold", colour = "black") +
  annotate("text",
           x = 10,   # adjust based on your FC range
           y = max(-log10(deg_results_list_preprocessed$cfr_CIRIvsOPTIOX$padj), na.rm = TRUE) * 0.9,
           label = "CIRI",
           hjust = 1, size = 4, fontface = "bold", colour = "black")+
theme(
  panel.background = element_rect(fill = "white", colour = "white"),
  plot.background  = element_rect(fill = "white", colour = "white"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.line        = element_line(colour = "black"),
  legend.background = element_rect(fill = "white", colour = "white")
)

ggsave(filename = paste0(fname_prefix_volcano_deseq2, "_", "CIRIvsOPTIOX.pdf"),
     width = 6, height = 7)

EnhancedVolcano(deg_results_list_preprocessed$cfr_NTCvshiPSCs,
                lab = deg_results_list_preprocessed$cfr_NTCvshiPSCs$gene_name,
                x = 'log2FoldChange',
                y = 'padj',
                pCutoff = 0.05,
                FCcutoff = log2(2),
                title = 'NTC vs hiPSCs',
                subtitle = 'Volcano Plot',
                caption = 'Cutoff: padj < 0.05 & |log2FC| > 1',
                pointSize = 2.0,
                #labSize = 3, 
  ###add colors
  colCustom = ok_pal[deg_results_list_preprocessed$cfr_NTCvshiPSCs$direction],
  colAlpha=0.5,
  # --- labeling options ---
  drawConnectors = TRUE,            # add arrow lines to labels
  widthConnectors = 0.1,            # arrow thickness
  colConnectors = 'black',
  # --- label appearance ---
  labSize = 2.5,                    # font size for labels
  boxedLabels = F,               # optional: draw boxes around labels
  max.overlaps = 15,                # control overlap density
  # --- control which genes to label ---
  #selectLab = head(
   # deg_results_list_preprocessed$cfr_TT2MN_vs_TMM$gene_name[
    #  order(deg_results_list_preprocessed$cfr_TT2MN_vs_TMM$padj)
    #],
    #60                             # label top 20 most significant genes
 # ),
  #legendPosition = 'right',
  legendLabSize = 10,
  legendIconSize = 3.5,
  borderWidth = 0.6,
  axisLabSize = 12
  )+
 annotate("text",
           x = -5,  # adjust based on your FC range
           y = max(-log10(deg_results_list_preprocessed$cfr_NTCvshiPSCs$padj), na.rm = TRUE) * 0.9,
           label = "iPSC",
           hjust = 0, size = 4, fontface = "bold", colour = "black") +
  annotate("text",
           x = 5,   # adjust based on your FC range
           y = max(-log10(deg_results_list_preprocessed$cfr_NTCvshiPSCs$padj), na.rm = TRUE) * 0.9,
           label = "NTC",
           hjust = 1, size = 4, fontface = "bold", colour = "black")+
theme(
  panel.background = element_rect(fill = "white", colour = "white"),
  plot.background  = element_rect(fill = "white", colour = "white"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.line        = element_line(colour = "black"),
  legend.background = element_rect(fill = "white", colour = "white")
)
ggsave(filename = paste0(fname_prefix_volcano_deseq2, "_", "NTCvshiPSCs.pdf"),
     width = 6, height = 7)



```


####calculate nb of DEGs per comparison

```{r}
# Extract only the *_direction columns
direction_cols <- grep("_direction$", colnames(deg_merged), value = TRUE)

# Build a long-format table (gene_ID, gene_name, comparison, direction)
deg_long <- deg_merged %>%
  select(gene_ID, gene_name, all_of(direction_cols)) %>%
  pivot_longer(
    cols = all_of(direction_cols),
    names_to = "comparison",
    values_to = "direction"
  ) %>%
  mutate(
    # clean comparison name (remove "_direction" suffix)
    comparison = str_remove(comparison, "_direction$")
  )

# Count Up / Down / NS per comparison
deg_summary <- deg_long %>%
  group_by(comparison, direction) %>%
  summarise(n_genes = n(), .groups = "drop") %>%
  # optional: pivot wider to have separate columns for Up/Down/NS
  pivot_wider(names_from = direction, values_from = n_genes, values_fill = 0) %>%
  arrange(comparison)

print(deg_summary)

write.table(
  deg_summary,
  file = paste0(fname_prefix_csv_deseq2, "_DEG_summary_counts.txt"),
  sep = "\t", quote = FALSE, row.names = FALSE
)

g=as.data.frame(t(deg_summary))
colnames(g)=g[1,]
g=g[-1,]
g$direction=rownames(g)

deg_summary_long <- deg_summary %>%
  pivot_longer(cols = c(Up, Down, NS),
               names_to = "direction", values_to = "count")

ggplot(deg_summary_long, aes(x = comparison, y = count, fill = direction)) +
  geom_bar(stat = "identity", position = "stack") +
  coord_flip() +
  labs(
    title = "DEG summary per comparison",
    x = "",
    y = "Number of genes"
  ) +
  scale_fill_manual(values = c("Up" = "#CC79A7", "Down" = "#009E73", "NS" = "grey70")) +
  theme_bw()
ggsave(filename = paste0(fname_prefix_volcano_deseq2, "_DEG_summary_barplot.pdf"),
     width = 8, height = 6)


```


###now run GSEA analisis in general on general signatures

```{r}

####make gene list to study

library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(enrichplot)
library(DOSE)

set_min_n_genes=5

comparisons=c("cfr_CIRIvsOPTIOX",
              "cfr_NTCvshiPSCs")

genes_cfr_CIRIvsOPTIOX=subset(de_stats,de_stats$cfr_CIRIvsOPTIOX_direction!="NS")
genes_cfr_NTCvshiPSCs=subset(de_stats,de_stats$cfr_NTCvshiPSCs_direction!="NS")

genes <- as.numeric(genes_cfr_NTCvshiPSCs$cfr_NTCvshiPSCs_log2FC)
names(genes) <- genes_cfr_NTCvshiPSCs$gene_name
genes <- sort(genes, decreasing = TRUE)
genes_cfr_NTCvshiPSCs=genes

genes <- as.numeric(genes_cfr_CIRIvsOPTIOX$cfr_CIRIvsOPTIOX_log2FC)
names(genes) <- genes_cfr_CIRIvsOPTIOX$gene_name
genes <- sort(genes, decreasing = TRUE)
genes_cfr_CIRIvsOPTIOX=genes



gene_sets_GSEA <- list(
  "cfr_CIRIvsOPTIOX" = genes_cfr_CIRIvsOPTIOX,
  "cfr_NTCvshiPSCs"    = genes_cfr_NTCvshiPSCs
)

# Ensure each element is a named numeric vector (values = logFC, names = SYMBOL)
stopifnot(all(sapply(gene_sets_GSEA, function(x) is.numeric(x) && !is.null(names(x)))))

# Map SYMBOL -> ENTREZ, then rebuild ranked vectors in ENTREZ space
symbol_to_entrez <- suppressMessages(bitr(
  unique(unlist(lapply(gene_sets_GSEA, names))),
  fromType="SYMBOL", toType="ENTREZID", OrgDb=org.Hs.eg.db
))

gsea_list   <- list()
GSEA_tables <- list()

for (nm in names(gene_sets_GSEA)) {
  ranks_sym <- gene_sets_GSEA[[nm]]

  # join to map to ENTREZ
  df <- data.frame(SYMBOL = names(ranks_sym), rank = as.numeric(ranks_sym))
  df2 <- merge(df, symbol_to_entrez, by="SYMBOL")
  if (!nrow(df2)) next

  # deduplicate ENTREZ if necessary (keep max |rank|)
  df2 <- df2[order(-abs(df2$rank)), ]
  df2 <- df2[!duplicated(df2$ENTREZID), ]

  ranks_entrez <- df2$rank
  names(ranks_entrez) <- df2$ENTREZID
  ranks_entrez <- sort(ranks_entrez, decreasing = TRUE)

  gsea <- gseGO(geneList      = ranks_entrez,
                OrgDb         = org.Hs.eg.db,
                keyType       = "ENTREZID",
                ont           = "BP",
                minGSSize     = set_min_n_genes,#place this automatic
                maxGSSize     = 500,
                pAdjustMethod = "BH",
                pvalueCutoff  = 1,
                nPermSimple = 10000,# readable=TRUE,
                verbose       = FALSE, by = "fgsea")
  
gsea_readable <- setReadable(gsea, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

  gsea_list[[nm]]   <- gsea
  GSEA_tables[[nm]] <- as.data.frame(gsea_readable)
}

# filter significant terms
# Make a new list of GSEA results filtered by significance
gsea_sig_list <- lapply(gsea_list, function(gsea_obj) {
  if (is.null(gsea_obj)) return(NULL)
  df <- gsea_obj@result
  if (is.null(df) || !nrow(df)) return(NULL)
  
  # filter by adjusted p-value < 0.05
  df_sig <- dplyr::filter(df, p.adjust < 0.05)
  
  # replace the result slot with filtered df
  gsea_obj@result <- df_sig
  gsea_obj
})

# 1) Write every GO table to CSV: GO_<name>.csv
if (length(GSEA_tables)) {
  for (nm in names(GSEA_tables)) {
    if (is.null(GSEA_tables[[nm]]) || !nrow(GSEA_tables[[nm]])) next
    # file <- file.path(out_dir, paste0(fname_prefix_GO, "_GO_", nm, ".csv"))
    file <- paste0(fname_prefix_csv_deseq2, "_GSEA_DEG_", nm, ".csv")
    write.csv(GSEA_tables[[nm]], file, row.names = FALSE)
    message("Wrote: ", file)
  }
}

# Helper: save ggplot if it exists
.save_plot <- function(p, file, w=8, h=10) {
  if (!is.null(p) && inherits(p, "ggplot")) {
    ggsave(filename = file, plot = p, width = w, height = h)
    message("Saved: ", file)
  }
}

# Make a dotplot per GSEA result in gsea_list
plot_gsea_dotplots <- function(gsea_list, fname_prefix, show_k = 10, width=8, height=6) {
  if (!length(gsea_list)) {
    message("gsea_list is empty.")
    return(invisible(NULL))
  }

  for (nm in names(gsea_list)) {
    g <- gsea_list[[nm]]
    # Skip NULL or empty results
    df <- tryCatch(as.data.frame(g), error = function(e) NULL)
    if (is.null(df) || !nrow(df)) {
      message("Skipping ", nm, " (no terms).")
      next
    }

    k <- min(show_k, nrow(df))

    # GSEA-friendly dotplot: split by .sign for up/down, order by NES
    p <- tryCatch(
      dotplot(g, showCategory = k, split = ".sign", orderBy = "NES") +
        facet_grid(. ~ .sign),
      error = function(e) { message("Plot error in ", nm, ": ", e$message); NULL }
    )

    out_pdf <- paste0(fname_prefix, "_GSEA_", nm, "_dotplot.pdf")
    .save_plot(p, out_pdf, w = width, h = height)
  }
}

plot_gsea_dotplots(gsea_list, fname_prefix = fname_prefix_gene_GO_deseq2, show_k = 10)


# Vector of target GO IDs ###ID studied before
go_ids <- c("GO:0007059","GO:0030855","GO:002240","GO:0070252","GO:0006936") # 

# Function: loop through GO IDs for a given contrast
save_gsea_plots <- function(gsea_res, contrast_name, go_ids, prefix, width=8, height=6) {
  for (go_id in go_ids) {
    if (go_id %in% gsea_res@result$ID) {
      p <- gseaplot2(gsea_res, geneSetID = go_id, title = go_id)
      out_file <- paste0(prefix, "_", contrast_name, "_", gsub(":", "-", go_id), ".pdf")
      ggsave(out_file, plot = p, width = width, height = height)
      message("Saved: ", out_file)
    } else {
      message("GO term ", go_id, " not found in ", contrast_name)
    }
  }
}

# Example usage:
save_gsea_plots(gsea_list[["cfr_CIRIvsOPTIOX"]], "cfr_CIRIvsOPTIOX", go_ids, fname_prefix_gene_GO_deseq2)
save_gsea_plots(gsea_list[["cfr_NTCvshiPSCs"]], "cfr_NTCvshiPSCs", go_ids, fname_prefix_gene_GO_deseq2)

# Vector of target GO IDs for plotting
#go_ids <- c("GO:0042551","GO:0007492","GO:0007498")#neuron migration

# Loop over and plot#
#for (go_id in go_ids) {
 # if (go_id %in% gsea_list[["3vsall"]]@result$ID) {
  #  print(gseaplot2(gsea_list[["3vsall"]], geneSetID = go_id, title = go_id))
  #} else {
   # message(paste("GO term", go_id, "not found in results."))
  #}
#}


#dotplot(gsea, showCategory = 10, split = ".sign",orderBy = "enrichmentScore") + facet_grid(.~.sign)
#ggsave(filename = paste0(fname_prefix_ATAC_GO, "_", "_GSEA.pdf"),
 #    width = 8, height = 8)

### Plotting GSEA results
#dotplot(gsea_list[["3vsall"]], showCategory = 10, split = ".sign",orderBy = "enrichmentScore") + facet_grid(.~.sign)
#ggsave(filename = paste0(fname_prefix_GO, "_", "_GSEA.pdf"),
 #    width = 8, height = 8)


Rdata_filename = paste0(fname_prefix_gene_GO_deseq2, "_GSEA_analysis_DEG.RData")
save(gene_sets_GSEA,gsea_list,GSEA_tables,de_stats,gsea_readable,
     file = Rdata_filename)

gsea_res <- gsea_list[["cfr_CIRIvsOPTIOX"]]

# Pick a GO ID manually or find it by keyword
term_to_plot <- c("GO:0061061")  # or
#term_to_plot <- grep("muscle cell differentiation", gsea_res@result$Description, value = TRUE)[13] #NB this is not the ID but Description

###this is top for optiox

if (term_to_plot %in% gsea_res@result$ID) {
  p <- gseaplot2(
    gsea_res,
    geneSetID = term_to_plot,
    title = gsea_res@result$Description[gsea_res@result$ID == term_to_plot],
    subplots = 1:3,  # Enrichment curve + NES info
    color = "firebrick",
    pvalue_table = TRUE
  )
  print(p)
} else {
  message("GO term ", term_to_plot, " not found in GSEA results.")
}

ggsave(filename = paste0(fname_prefix_gene_GO_deseq2, "_", "GSEA_muscle_cell_differentiation_CIRIvsOPTIOX.pdf"),
    width = 8,
    height = 4)

###now try place multiple courves in one plot
gsea_res <- gsea_list[["cfr_CIRIvsOPTIOX"]]

# Pick a GO ID manually or find it by keyword
#term_to_plot <- c("GO:0042692","GO:0007059","GO:0030855")  # or
term_to_plot=c("GO:1903047","GO:0014706","GO:0008544")
valid_terms <- intersect(term_to_plot, gsea_res@result$ID)

if (length(valid_terms) == 0) {
  stop("None of the requested GO terms are present in GSEA results.")
}

# Define colors (must match number of terms)
palette <- c("#E69F00","#000000","#009292")
#palette <- c("#000000","#009292")


# Plot multiple enrichment curves
p <- gseaplot2(
  gsea_res,
  geneSetID   = valid_terms,
  subplots    = 1:2,
  color       = palette[seq_along(valid_terms)],
  base_size   = 14,
  pvalue_table = T   # not available with multiple curves
)

print(p)

ggsave(filename = paste0(fname_prefix_gene_GO_deseq2, "_", "GSEA_CIRIvsOPTIOX.pdf"),
    width = 8,
    height = 5)
```

####Running GSEA on a custom gene list

###create gene list

```{r}
# ============================================================
# HUMAN SKELETAL MUSCLE & iPSC-MYOGENIC SIGNATURES (v2025)
# Integrated from Chemello et al. 2025, De Micheli et al. 2020,
# Murgia et al. 2021, Hicks et al. 2025, Front Cell Dev Biol 2025,
# and classic literature (Schiaffino & Reggiani 2011; Bodine 2001).
# ============================================================

gs <- list(

  # ===============================
  # CELL CYCLE / RIBOSOME TONE
  # ===============================
  cellcycle_s = c("MCM5","PCNA","TYMS","FEN1","MCM2","MCM4","RRM1","UNG","GINS2",
                  "MCM6","CDCA7","DTL","PRIM1","UHRF1","HELLS","RFC2","RPA2","NASP",
                  "RAD51","CHEK1","ORC6","MCM3","MCM7","MLF1IP","BRIP1","E2F8"),
  # Ref: Seurat S.Score; Whitfield et al., Cell 2002; Trapnell et al., Nat Biotech 2014

  cellcycle_g2m = c("HMGB2","CDK1","NUSAP1","UBE2C","BIRC5","TPX2","TOP2A","NDC80",
                    "CKS2","NUF2","CKS1B","MKI67","TMPO","CENPF","TACC3","SMC4",
                    "CCNB2","CKAP2L","CKAP2","AURKB","BUB1","KIF11","ANP32E","TUBB4B",
                    "GTSE1","KIF20B","HJURP","CDCA3","CDC20","TTK","CDC45","CDC6",
                    "EXO1","TIPIN","DSCC1","BLM","CASP8AP2","USP1","CLSPN","POLA1",
                    "CHAF1B","BRCA1","EXOSC2","USP16","POLD3","MSH2","ATAD2","RAD51AP1","RRM2"),
  # Ref: Seurat G2M.Score; Whitfield et al., Cell 2002; Trapnell 2014
  
  ribosome_tone = c("RPLP0","RPL3","RPL5","RPL7","RPL11","RPL13A","RPL23A",
                    "RPS3","RPS6","RPS8","RPS12","RPS18","RPS27","NCL","FBL",
                    "NPM1","EIF4EBP1","MYC","DDIT4"),
  # Ref: Murgia et al. 2021; Chemello et al. 2025 (fiber proteome core)

  # ===============================
  # FIBER-TYPE / METABOLIC PROGRAMS
  # ===============================
  fiber_type_I_slow = c("MYH7","TNNI1","TNNT1","MYL2","ATP2A2","CKMT2","MB",
                        "COX7A1","COX6A2","PPARGC1A","SDHA","CS","SLC25A4",
                        "MYBPC1","MYL3","PDK4","ESRRA","NDUFA10"),
  # Ref: Schiaffino & Reggiani 2011; Chemello et al. 2025; Murgia et al. 2021

  fiber_type_IIa_fast = c("MYH2","TNNI2","TNNT3","MYL1","ATP2A1","ACTN3","TPM2",
                          "PVALB","LDHA","PKM","PFKM","ENO1","PGAM2","ALDOA",
                          "CKM","DMD","CASQ1"),
  # Ref: Murgia et al. 2021; Bloemberg & Quadrilatero 2012; Chemello 2025

  fiber_type_IIx_fast = c("MYH1","ACTN3","MYLK2","TNNT3","TNNI2","ATP2A1","TPM2",
                          "CKM","LDHA","ENO1","PFKM","MYH4"),
  # Ref: Murgia 2021; Uhlén et al., HPA 2021

  developmental_embryonic_perinatal = c("MYH3","MYH8","MYL4","TNNI1","TNNT1",
                                        "MYOG","MYOD1","MEF2C","DES","MYMK",
                                        "MYMX","CHRNG","ACTA1","CKM"),
  # Ref: Chal et al. 2015; Hicks et al. 2025; Sampath et al. 2018

  sarcomere_core = c("TTN","NEB","ACTN2","TCAP","MYBPC1","MYBPC2","MYL1","MYL2",
                     "TNNI1","TNNI2","TNNT1","TNNT3","TPM1","TPM2","MYOM1","MYOM2","CSRP3"),
  # Ref: Lundberg et al. 2019; GO:0030016 "myofibril"

  calcium_excitation_contraction = c("RYR1","CACNA1S","ATP2A1","ATP2A2","CASQ1",
                                     "CASQ2","SLN","PLN","TRDN","STAC3","JPH1",
                                     "CAMK2A","CAMK2B","CAV3"),
  # Ref: Protasi et al. 2021; Jorgensen et al. 2023 (hiPSC EC coupling)

  oxidative_mitochondrial = c("SDHA","SDHB","UQCRC1","NDUFA9","NDUFS1","COX4I1",
                              "COX6A2","ATP5F1A","ATP5MC1","CKMT2","SLC25A4",
                              "PPARGC1A","TFAM","CS","ACADM","NDUFB8","CYCS","PDHA1"),
  # Ref: Lin et al. Nature 2002; Chemello et al. 2025; Murgia 2021

  glycolytic_fast = c("HK2","PFKM","ALDOA","TPI1","GAPDH","PGK1","PGAM2","ENO1",
                      "PKM","LDHA","PDK1","SLC2A4","PFKFB3","ENO3","MYL1"),
  # Ref: Schiaffino & Reggiani 2011; Chemello 2025

  # ===============================
  # STEM / MYOGENIC LINEAGES
  # ===============================
  satellite_cell_core = c("PAX7","VCAM1","ITGA7","CXCR4","MYF5","CDH15","SDC4",
                          "EGFR","NCAM1","SPRY1","MUSTN1","NOS1"),
  # Ref: De Micheli et al. 2020; Yin et al. 2013; GSE143437 atlas

  myoblast_proliferation_cycle = c("MYOD1","MYF5","MKI67","TOP2A","PCNA","CCNB1",
                                   "CDK1","E2F1","NUSAP1","HMGB2","UBE2C","AURKB","RRM2"),
  # Ref: Hicks et al. 2025; Trapnell et al. 2014 pseudotime

  myotube_differentiation_fusion = c("MYOG","MEF2C","MYMK","MYMX","MYH3","MYH8",
                                     "CKM","DES","DNM2","ACTA1"),
  # Ref: Chal et al. 2015; Sampath et al. 2018; Hicks 2025

  myogenic_regulators_axis = c("PAX3","PAX7","MYF5","MYOD1","MYOG","MEF2C","MYF6"),
  # Ref: Weintraub 1991; Rudnicki et al. 1993

  # ===============================
  # NMJ / ECM / NICHE
  # ===============================
  neuromuscular_junction = c("AGRN","LRP4","MUSK","DOK7","RAPSN","COLQ","CHRNA1",
                             "CHRNB1","CHRND","CHRNE","CHRNG","ACHE","UTRN"),
  # Ref: Burden et al. 2018; Cantor et al. 2022 (hiPSC NMJ coculture)

  muscle_ecm_basement = c("LAMA2","LAMC1","COL6A1","COL6A2","COL6A3","COL1A1",
                          "COL1A2","COL3A1","FN1","SPP1","LUM","DCN","DAG1",
                          "ITGA7","THY1","COL4A1","COL4A2","COL5A1"),
  # Ref: De Micheli et al. 2020; Saini et al. 2021

  fibro_adipogenic_progenitors = c("PDGFRA","COL1A1","COL3A1","DCN","LUM","CXCL14",
                                   "THY1","GPC6","FBN1","DPT"),
  # Ref: De Micheli 2020; Front Cell Dev Biol 2025

  atrophy_stress = c("FBXO32","TRIM63","DDIT4","GADD45A","ATF4","FOXO3","PSMA1",
                     "PSMB1","SESN1","SESN3","EIF4EBP1","PPP1R15A"),
  # Ref: Bodine et al. 2001; Gomes et al. 2001; Milan et al. 2021

  # ===============================
  # PLURIPOTENCY → MYOGENIC (hiPSC/hESC)
  # ===============================
  pluripotency_core = c("POU5F1","NANOG","SOX2","PRDM14","DPPA2","DPPA4","ZFP42",
                        "DNMT3A","DNMT3B"),
  # Ref: Theunissen et al. 2016; Nichols & Smith 2012

  naive_pluripotency = c("KLF17","KLF4","TFCP2L1","DNMT3L","DPPA3","TBX3","ESRRB","SUSD2"),
  # Ref: Collier et al. 2022; Pastor et al. 2018

  mesoderm_specification = c("MESP1","PDGFRA","KDR","ISL1","HAND1","HAND2","GATA4",
                             "TBX5","NKX2-5","MEF2C","FGF8","EOMES","TBXT"),
  # Ref: Chan et al. 2013; Loh et al. 2016; Hicks 2025

  early_myogenic_commitment = c("PAX3","PAX7","MYF5","MYOD1","MET","ITGA7","FGFR4","LBX1","MYOG")
  # Ref: Chal et al. 2015; Hicks et al. 2025
)
```

####Running GSEA on the custom gene list

```{r}

###one by one GSEA
#gene_ranks <- genes_cfr_GERALT_vs_OPTIOX
#gene_ranks <- sort(gene_ranks, decreasing = TRUE)

#term2gene <- stack(gs)
#colnames(term2gene) <- c("gene", "term")
#term2gene <- term2gene[, c("term", "gene")]

# Run GSEA
#gsea_custom <- GSEA(
 # geneList = gene_ranks,
  #TERM2GENE = term2gene,
  #minGSSize = 5,
  #maxGSSize = 500,
  #pvalueCutoff = 1,
  #verbose = FALSE,
  #seed = TRUE
#)

#as.data.frame(gsea_custom)[, c("ID", "Description", "NES", "pvalue", "p.adjust")]

# Parameters
set_min_n_genes <- 3
p_cutoff <- 1

# --- Prepare TERM2GENE mapping from your gene sets ---
term2gene <- stack(gs)
colnames(term2gene) <- c("gene", "term")
term2gene <- term2gene[, c("term", "gene")]

# --- Initialize result containers ---
gsea_list_custom   <- list()
GSEA_tables_custom <- list()

# --- Loop over each comparison in your ranked gene lists ---
for (nm in names(gene_sets_GSEA)) {
  message("Running custom GSEA for: ", nm)
  
  # Extract ranked genes (log2FC values)
  gene_ranks <- gene_sets_GSEA[[nm]]
  gene_ranks <- sort(gene_ranks, decreasing = TRUE)
  
  # Run clusterProfiler::GSEA on your custom gene sets
  gsea_custom <- tryCatch({
    GSEA(
      geneList      = gene_ranks,
      TERM2GENE     = term2gene,
      minGSSize     = set_min_n_genes,
      maxGSSize     = 500,
      pAdjustMethod = "BH",
      pvalueCutoff  = p_cutoff,
      verbose       = FALSE,
      seed          = TRUE
    )
  }, error = function(e) {
    message("Error in ", nm, ": ", e$message)
    NULL
  })
  
  if (is.null(gsea_custom) || nrow(as.data.frame(gsea_custom)) == 0) {
    message("No enrichment found for ", nm)
    next
  }
  
  # Save objects
  gsea_list_custom[[nm]]   <- gsea_custom
  GSEA_tables_custom[[nm]] <- as.data.frame(gsea_custom)
}

# --- Filter significant results (adjusted p < 0.05) ---
gsea_sig_list_custom <- lapply(gsea_list_custom, function(gsea_obj) {
  if (is.null(gsea_obj)) return(NULL)
  df <- gsea_obj@result
  if (is.null(df) || !nrow(df)) return(NULL)
  df_sig <- dplyr::filter(df, p.adjust < 0.05)
  gsea_obj@result <- df_sig
  gsea_obj
})

# --- Write GSEA tables to CSV ---
if (length(GSEA_tables_custom)) {
  for (nm in names(GSEA_tables_custom)) {
    df <- GSEA_tables_custom[[nm]]
    if (is.null(df) || !nrow(df)) next
    
    file <- paste0(fname_prefix_csv_deseq2, "_CustomGSEA_", nm, ".csv")
    write.csv(df, file, row.names = FALSE)
    message("Wrote: ", file)
  }
}

# --- Helper function to save plots ---
.save_plot <- function(p, file, w = 8, h = 10) {
  if (!is.null(p) && inherits(p, "ggplot")) {
    ggsave(filename = file, plot = p, width = w, height = h)
    message("Saved: ", file)
  }
}

# --- Generate dotplots for each comparison ---
plot_gsea_dotplots_custom <- function(gsea_list, fname_prefix, show_k = 10, width = 8, height = 6) {
  if (!length(gsea_list)) {
    message("gsea_list is empty.")
    return(invisible(NULL))
  }
  
  for (nm in names(gsea_list)) {
    g <- gsea_list[[nm]]
    
    # Safely convert to data frame
    df <- tryCatch(as.data.frame(g), error = function(e) NULL)
    if (is.null(df) || !nrow(df)) {
      message("Skipping ", nm, " (no terms).")
      next
    }
    
    k <- min(show_k, nrow(df))
    
    # --- FIXED: added missing closing parenthesis and split handling ---
    p <- tryCatch({
      dotplot(
        g,
        showCategory = k,
        split = ".sign",
        orderBy = "NES",
        title = paste0("Custom GSEA: ", nm)
      ) + facet_grid(. ~ .sign)
    },
    error = function(e) {
      message("Plot error in ", nm, ": ", e$message)
      NULL
    })
    
    # Save PDF
    out_pdf <- paste0(fname_prefix, "_CustomGSEA_", nm, "_dotplot.pdf")
    .save_plot(p, out_pdf, w = width, h = height)
  }
}

plot_gsea_dotplots_custom(gsea_list_custom, fname_prefix = fname_prefix_gene_GO_deseq2, show_k = 10)

# --- Save the workspace for reproducibility ---
Rdata_filename <- paste0(fname_prefix_gene_GO_deseq2, "_CustomGSEA_analysis.RData")
save(gsea_list_custom, GSEA_tables_custom, gsea_sig_list_custom, gs, term2gene, file = Rdata_filename)
message("Saved RData: ", Rdata_filename)

```
###make enrichment plots for specific gene sets

```{r}
###now try place multiple courves in one plot
gsea_res <- gsea_list_custom[["cfr_CIRIvsOPTIOX"]]

# Pick a GO ID manually or find it by keyword
term_to_plot <- c("sarcomere_core","fiber_type_IIa_fast","early_myogenic_commitment")  # or
valid_terms <- intersect(term_to_plot, gsea_res@result$ID)

if (length(valid_terms) == 0) {
  stop("None of the requested GO terms are present in GSEA results.")
}

# Define colors (must match number of terms)
palette <- c("#E69F00","#000000","#009292")

# Plot multiple enrichment curves
p <- gseaplot2(
  gsea_res,
  geneSetID   = valid_terms,
  subplots    = 1:2,
  color       = palette[seq_along(valid_terms)],
  base_size   = 14,
  pvalue_table = T   # not available with multiple curves
)

print(p)

ggsave(filename = paste0(fname_prefix_gene_GO_deseq2, "_", "GSEA_CIRIvsOPTIOX_custom.pdf"),
    width = 8,
    height = 5)
```


###make volcano plot by showing a gene signature
#CIRI vs OPTIOX

```{r}
# define the signature genes to label
my_signature <- gs$sarcomere_core  # example gene set

# prepare the data
deg_df <- deg_results_list_preprocessed$cfr_CIRIvsOPTIOX

# define thresholds
fc_cutoff <- log2(2)
padj_cutoff <- 0.05

# filter to keep only genes that are significant and in the signature
sig_in_data <- deg_df %>%
  filter(
    gene_name %in% my_signature &
      padj < padj_cutoff &
      abs(log2FoldChange) > fc_cutoff
  ) %>%
  pull(gene_name)

# color palette (same as before)
ok_pal <- c("Up" = "#CC79A7", "Down" = "#009E73", "NS" = "grey70")

# volcano plot
EnhancedVolcano(
  deg_df,
  lab = deg_df$gene_name,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = padj_cutoff,
  FCcutoff = fc_cutoff,
  title = "CIRI vs OPTIOX",
  subtitle = "Volcano Plot: sarcomere core signature genes",
  caption = "Cutoff: padj < 0.05 & |log2FC| > 1",
  pointSize = 2.0,
  colCustom = ok_pal[deg_df$direction],
  colAlpha = 0.5,
  drawConnectors = TRUE,
  widthConnectors = 0.3,
  colConnectors = "black",
  boxedLabels = TRUE,
  labSize = 2.5,
  max.overlaps = 50,
  # --- highlight only your signature *and significant* genes ---
  selectLab = sig_in_data,
  legendLabSize = 10,
  legendIconSize = 3.5,
  borderWidth = 0.6,
  axisLabSize = 12
) +
  annotate(
    "text",
    x = -10,
    y = max(-log10(deg_df$padj), na.rm = TRUE) * 0.9,
    label = "OPTIOX",
    hjust = 0, size = 4, fontface = "bold", colour = "black"
  ) +
  annotate(
    "text",
    x = 10,
    y = max(-log10(deg_df$padj), na.rm = TRUE) * 0.9,
    label = "CIRI",
    hjust = 1, size = 4, fontface = "bold", colour = "black"
  ) +
  theme(
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.background = element_rect(fill = "white", colour = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.background = element_rect(fill = "white", colour = "white")
  )

ggsave(
  filename = paste0(fname_prefix_volcano_deseq2, "_CIRI_OPTIOX_sarcomere_core_sigOnly.pdf"),
  width = 6, height = 7
)
```

###make volcano plot by showing a gene signature
#GERALT vs OPTIOX gliycolitic (OPTIOX) and type 1 fibers (GERALT)

```{r}
# define the signature genes to label
my_signature <- c(gs$sarcomere_core, gs$fiber_type_I_slow, gs$early_myogenic_commitment) # example gene set

# prepare the data
deg_df <- deg_results_list_preprocessed$cfr_CIRIvsOPTIOX

# define thresholds
fc_cutoff <- log2(2)
padj_cutoff <- 0.05

# filter to keep only genes that are significant and in the signature
sig_in_data <- deg_df %>%
  filter(
    gene_name %in% my_signature &
      padj < padj_cutoff &
      abs(log2FoldChange) > fc_cutoff
  ) %>%
  pull(gene_name)

# color palette (same as before)
ok_pal <- c("Up" = "#CC79A7", "Down" = "#009E73", "NS" = "grey70")

# volcano plot
EnhancedVolcano(
  deg_df,
  lab = deg_df$gene_name,
  x = "log2FoldChange",
  y = "padj",
  pCutoff = padj_cutoff,
  FCcutoff = fc_cutoff,
  title = "CIRI vs OPTIOX",
  subtitle = "Volcano Plot: early myogenic,sarcomere core and type I slow fibers signature genes",
  caption = "Cutoff: padj < 0.05 & |log2FC| > 1",
  pointSize = 2.0,
  colCustom = ok_pal[deg_df$direction],
  colAlpha = 0.5,
  drawConnectors = TRUE,
  widthConnectors = 0.3,
  colConnectors = "black",
  boxedLabels = F,
  labSize = 2.5,
  max.overlaps = 50,
  # --- highlight only your signature *and significant* genes ---
  selectLab = sig_in_data,
  legendLabSize = 10,
  legendIconSize = 3.5,
  borderWidth = 0.6,
  axisLabSize = 12
) +
  annotate(
    "text",
    x = -10,
    y = max(-log10(deg_df$padj), na.rm = TRUE) * 0.9,
    label = "OPTIOX",
    hjust = 0, size = 4, fontface = "bold", colour = "black"
  ) +
  annotate(
    "text",
    x = 10,
    y = max(-log10(deg_df$padj), na.rm = TRUE) * 0.9,
    label = "CIRI",
    hjust = 1, size = 4, fontface = "bold", colour = "black"
  ) +
  theme(
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.background = element_rect(fill = "white", colour = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(colour = "black"),
    legend.background = element_rect(fill = "white", colour = "white")
  )

ggsave(
  filename = paste0(fname_prefix_volcano_deseq2, "_CIRI_OPTIOX_early_sarcomeric_and _fibers_sigOnly.pdf"),
  width = 6, height = 7
)
```

####compute signature scores

# ============================================================
# Signature scores (module scores) + comparisons by cell_type
# Uses vsd (VST), geneInfo_f (ENSG->symbol), meta (sample annotations)
# ============================================================


```{r}

library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)

# ---------- 1) Build SYMBOL-labeled expression matrix from vsd ----------
expr <- assay(vsd)  # rows ~ ENSG + custom, cols = samples

# ENSG -> SYMBOL map (strip version suffix if present)
map <- geneInfo_f %>%
  transmute(
    ens = sub("\\..*$", "", gene_ID),
    sym = gene_name
  ) %>%
  distinct() %>%
  filter(!is.na(sym), sym != "")

expr_df <- as.data.frame(expr) %>%
  rownames_to_column("rid_raw") %>%
  mutate(ens = sub("\\..*$", "", rid_raw)) %>%
  left_join(map, by = "ens") %>%
  filter(!is.na(sym), sym != "")

# Collapse duplicate symbols (mean across duplicates)
expr_sym <- expr_df %>%
  select(-rid_raw, -ens) %>%
  group_by(sym) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE), .groups = "drop")

expr_sym_mat <- as.matrix(expr_sym[, -1])
rownames(expr_sym_mat) <- expr_sym$sym

# ---------- 2) Module score helper ----------
score_module_z <- function(expr_mat, genes, min_genes = 3) {
  genes <- intersect(genes, rownames(expr_mat))
  if (length(genes) < min_genes) return(rep(NA_real_, ncol(expr_mat)))
  m  <- expr_mat[genes, , drop = FALSE]
  mz <- t(scale(t(m)))            # z per gene across samples
  colMeans(mz, na.rm = TRUE)      # mean z per sample
}

# ---------- 3) Choose signatures to score ----------

sig_list <- list(
  # induction axis
  pluripotency_core   = gs$pluripotency_core,
  naive_pluripotency  = gs$naive_pluripotency,
  mesoderm_spec       = gs$mesoderm_specification,
  early_myogenic      = gs$early_myogenic_commitment,

  # myogenic state / maturation
  satellite_core      = gs$satellite_cell_core,
  myotube_fusion      = gs$myotube_differentiation_fusion,
  dev_embryo_peri     = gs$developmental_embryonic_perinatal,
  sarcomere_core      = gs$sarcomere_core,
  ec_coupling         = gs$calcium_excitation_contraction,

  # metabolism / fiber programs (optional)
  oxidative_mito      = gs$oxidative_mitochondrial,
  glycolytic_fast     = gs$glycolytic_fast,
  fiber_I_slow        = gs$fiber_type_I_slow,
  fiber_IIa_fast      = gs$fiber_type_IIa_fast
)

# Quick overlap check (useful for debugging)
overlap_n <- sapply(sig_list, \(g) length(intersect(g, rownames(expr_sym_mat))))
print(overlap_n)

# ---------- 4) Compute per-sample scores + combined indices ----------
scores <- lapply(sig_list, \(g) score_module_z(expr_sym_mat, g)) %>% as.data.frame()
scores$sample_ID <- colnames(expr_sym_mat)

scores <- scores %>%
  mutate(
    pluripotency_off      = -pluripotency_core,
    early_induction_on    = mesoderm_spec + early_myogenic,
    early_induction_index = early_induction_on - pluripotency_core
  ) %>%
  left_join(meta, by = "sample_ID")

# Optional: ensure name order & stable colors
okabe_named <- c(
  "NTC"    = "#000000",
  "hiPSC"  = "#56B4E9",
  "CIRI"   = "#CC79A7",
  "OPTIOX" = "#E69F00"
)

order_cell_type=c("hiPSC","NTC","CIRI","OPTIOX")

# If your meta uses different strings (e.g. "hiPSCs"), check and adjust:
# unique(scores$name)

# ---------- 5) Compare scores by cell_type (faceted), colored by name ----------
# Choose which score(s) you want to visualize:
score_vars <- c(
  "early_induction_index",
  "early_induction_on",
  "pluripotency_core",
  "pluripotency_off",
  "naive_pluripotency",
  "mesoderm_spec",
  "early_myogenic",
  "satellite_core",
  "myotube_fusion",
  "dev_embryo_peri",
  "sarcomere_core",
  "ec_coupling",
  "oxidative_mito",
  "glycolytic_fast",
  "fiber_I_slow",
  "fiber_IIa_fast"
)

scores_long <- scores %>%
  select(sample_ID, name, replicate, condition, cell_type, all_of(score_vars)) %>%
  pivot_longer(cols = all_of(score_vars), names_to = "signature", values_to = "score")

p_scores <- ggplot(scores_long, aes(x = factor(name,levels=order_cell_type), y = score, color = name)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  geom_jitter( width = 0.15, size = 3) +
  facet_wrap(~signature) +
  scale_color_manual(values = okabe_named, drop = FALSE) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Signature scores by cell type",
    x = "", y = "Module score (mean z across genes)"
  )

print(p_scores)

ggsave(
  filename = paste0(fname_prefix_gene_deseq2, "_SignatureScores_byCellType.pdf"),
  plot = p_scores, width = 18, height = 15
)

# ---------- 6) Stats: CIRI vs OPTIOX within each cell_type for each signature ----------
# (Uses Wilcoxon; no continuity correction; tiny n -> interpret carefully)
tests_CIRI_OPTIOX <- scores_long %>%
  filter(name %in% c("CIRI","OPTIOX")) %>%
  group_by(signature) %>%
  summarise(
    n_CIRI   = sum(name=="CIRI"   & !is.na(score)),
    n_OPTIOX = sum(name=="OPTIOX" & !is.na(score)),
    p = ifelse(min(n_CIRI, n_OPTIOX) >= 2,
               wilcox.test(score ~ name, exact=FALSE, correct=FALSE)$p.value,
               NA_real_),
    .groups="drop"
  ) %>%
  arrange(p)

print(tests_CIRI_OPTIOX)

write.table(
  tests_CIRI_OPTIOX,
  file = paste0(fname_prefix_csv_deseq2, "_SignatureScores_stats_CIRIvsOPTIOX_byCellType.txt"),
  sep = "\t", quote = FALSE, row.names = FALSE
)

# ---------- 7) Stats: hiPSC vs NTC within each cell_type for each signature ----------
tests_hiPSC_NTC <- scores_long %>%
  filter(name %in% c("hiPSC", "NTC")) %>%
  group_by(signature) %>%
  summarise(
    n_hiPSC = sum(name == "hiPSC" & !is.na(score)),
    n_NTC   = sum(name == "NTC"   & !is.na(score)),
    p = ifelse(min(n_hiPSC, n_NTC) >= 2,
               wilcox.test(score ~ name, exact = FALSE, correct = FALSE)$p.value,
               NA_real_),
    .groups = "drop"
  ) %>%
  arrange(p)

print(tests_hiPSC_NTC)

write.table(
  tests_hiPSC_NTC,
  file = paste0(fname_prefix_csv_deseq2, "_SignatureScores_stats_hiPSCvsNTC_byCellType.txt"),
  sep = "\t", quote = FALSE, row.names = FALSE
)
```









