---
title: "n1 AB017: QC and monocle Seurat objects creation"
author: "Elisa Balmas PhD"
date: "2026-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment

```{r}
#library(monocle3)
library(mixtools)
library(quantmod)
library(tidyr)
library(stringr)
library(grid)
library(viridis)
library(viridisLite)
library(ggthemes)
library(dplyr)
library(tidyverse)
library(ggsignif)
library(umap)
library(heatmap3)
library(plyr)
library(edgeR)
library(compareGroups)
library(dbscan)
library(MAST)
library(geosphere)
library(RColorBrewer)
library(htmlwidgets)
library(Matrix)
library(tidyseurat)
library(SeuratWrappers)
#library(scMAGeCK)
library(readxl)
library(SparseArray)
library(Seurat)
#library(rCASC)
#library(deepToolsUtils)
#library(R.utils)
library(sceptre)
library(sceptredata)
library(monocle3)


library(ggplot2); library(reshape2); theme_set(theme_bw(26) + theme(panel.grid.major = element_blank(), 
                                                                    panel.grid.minor = element_blank()) +
                                                 theme(legend.key = element_blank()))
update_geom_defaults("point", aes(size = 4))

#setwd("/home/ebalmas/scpHUB/projects/H001AS7_iKDseq_10X_4tp_monocle/analysis_EB")

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))


# Set up filename prefixes and output folders with the data
dir.create("Output")
#dir.create(file.path("Output","scratch"))


dir.create(file.path("Output","QC"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "filtering"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "umap"))
dir.create(file.path("Output","monocle"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplot"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "umap"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "heatmap"))


fname_scratch <- file.path("scratch")

#set up folder path QC
fname_prefix_R_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_csv_QC<-file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_ribomito_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_filtering_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "filtering", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_umap_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "umap", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_UMAP <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "umap", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplot", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_R <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_csv <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_heatmap <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "heatmap", 
                               format(Sys.Date(), "%y%m%d"))


#okabe_tab <- read_excel("scratch/color_scale.xlsx")
#okabe_pal=okabe_tab$hex

okabe_pal=c("#E69F00","#56B4E9","#009E73","#f0E442","#0072B2","#D55E00","#CC79A7","#000000")


pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
         "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
         "#920000","#924900","#db6d00","#24ff24","#ffff6d")
```

###load data
```{r}
load(paste0(getwd(),"/scratch/scratch_protein_coding_genes.RData"))
aggr_info <- read.csv(paste0(getwd(),"/scratch/aggregation.csv"))
aggr_info$orig.ident=rownames(aggr_info)

#load the RNA data in Seurat v5
data <- Seurat::Read10X_h5(paste0(getwd(),"/scratch/filtered_feature_bc_matrix.h5"),unique.features = T)
counts=data
counts=counts[row.names(counts) %in% protein_coding_genes,]#filter for protein coding genes straight away


```
#create Seurat object

```{r}
# create a Seurat object containing the RNA data
data_seurat <- CreateSeuratObject(
  counts = counts,
  assay = "RNA"
)
#sample_info$orig.ident=c("1","2","3")
meta_original=data_seurat[[]]
meta_original$libID=rownames(meta_original)

meta_original=meta_original%>%dplyr::mutate(orig.ident=sapply(strsplit(as.character(libID), "-"), "[[", 2),
lib_ID=sapply(strsplit(as.character(libID), "-"), "[[", 1))

meta_original=meta_original%>%dplyr::right_join(aggr_info, by="orig.ident")
rownames(meta_original)=meta_original$libID


#update metadata 
data_seurat<- AddMetaData(data_seurat, meta_original, col.name = NULL)
#data_seurat <- subset(data_seurat, subset =libID%in%select)

###select cells annotated with sgRNA from cellranger
#select=rownames(cell_info)
#update metadata and select for the cells annotated with sgRNA
#data_seurat <- subset(data_seurat, subset =libID%in%select)
#data_seurat<- AddMetaData(data_seurat, meta_original, col.name = NULL)

pre_QCcounts=meta_original %>%dplyr::group_by(sample_id)%>%dplyr::summarise(count = n())
write.csv(pre_QCcounts, file= paste0(fname_prefix_csv_QC, "_", "pre_QCcounts.csv"), row.names=T)

print(pre_QCcounts)
```

#QC library
####test QC cuts

```{r}

#explore variables:
mito_set1=1
mito_set2=15
ribo_set=3
nGene1=300
nGene2=7000
nUMIset=100

#########calculate % of mitocondiral reads and ribosomal reads
DefaultAssay(data_seurat) <- 'RNA'
data_seurat[["percent.mt"]] <- PercentageFeatureSet(data_seurat, pattern = "^MT-")
data_seurat[["percent.ribo"]] <- PercentageFeatureSet(data_seurat, pattern = "RPS")

QC=data_seurat[[c("nFeature_RNA","percent.mt","nCount_RNA","percent.ribo","sample_id")]]

#make a QC dataframe to plot the QC metrics and then use it as meta table for monocle
QC <- QC %>%
        dplyr::rename(nUMI = nCount_RNA,
                      nGene = nFeature_RNA)
QC$info=rownames(QC)
QC$log10GenesPerUMI <- log10(QC$nGene) / log10(QC$nUMI)


plot1 <- ggplot(QC, aes(x = sample_id, y = nGene, fill = sample_id)) +
  geom_point(position = position_jitter(width = 0.2), size = 0.5, alpha = 0.5) +
  geom_violin(scale = "width", alpha = 0.8) +
  scale_fill_manual(values = okabe_pal) +
  geom_hline(yintercept = nGene1, linetype = "dotted") +
  geom_hline(yintercept = nGene2, linetype = "dotted")

print(plot1)#nGene

plot2 <- ggplot(QC, aes(x = sample_id, y = nUMI, fill = sample_id)) +
  geom_point(position = position_jitter(width = 0.2), size = 0.5, alpha = 0.5) +
  geom_violin(scale = "width", alpha = 0.8) +
  geom_hline(yintercept = nUMIset, linetype = "dotted") +
  scale_fill_manual(values = okabe_pal)

print(plot2) #nUMI 

#not cleaned for nb of UMI yet
plot1 + plot2
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "feature_QC_old.pdf"),
     width = 10, height = 8)

###low UMI cut doesn't matter because gene cut is already cleaning those cells (as well as mito and ribo)
QC %>% 
  	ggplot(aes(color=sample_id, x=nUMI, fill= sample_id)) + 
  	geom_density(alpha = 0.2) +scale_fill_manual(values =c(okabe_pal)) + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nUMIset)
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "UMI_QC.pdf"),
     width = 10, height = 8)
###n genes detected
QC %>% 
  	ggplot(aes(color=sample_id, x=nGene, fill= sample_id)) + 
  	geom_density(alpha = 0.2) +scale_fill_manual(values =c(okabe_pal)) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = nGene1,linetype="dotted")+geom_vline(xintercept = nGene2,linetype="dotted") ###on the lower end cut was 200 but 00 or 350 is better
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_det_QC.pdf"),
     width = 10, height = 8)
QC %>% 
  	ggplot(aes(x=sample_id, y=log10(nGene), fill=sample_id)) + 
  	geom_boxplot() +scale_fill_manual(values =c(okabe_pal)) + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  geom_hline(yintercept = log10(nGene1),linetype="dotted")+geom_hline(yintercept = log10(nGene2),linetype="dotted")+
  	ggtitle("NCells vs NGenes")
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_QC.pdf"),
     width = 10, height = 8)
QC %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point(size=1.5,alpha = 0.5) + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = nUMIset) +
  	geom_hline(yintercept = nGene1,linetype="dotted")+geom_hline(yintercept = nGene2,linetype="dotted")
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_UMI_QC.pdf"),
     width = 10, height = 8)

QC %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=sample_id)) + 
  	geom_point(size=1.5,alpha = 0.5) + scale_colour_manual(values =c(okabe_pal)) + 
	#scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = nUMIset) +
  	geom_hline(yintercept = nGene1,linetype="dotted")+geom_hline(yintercept = nGene2,linetype="dotted")
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_UMI_QC_sample_id.pdf"),
     width = 10, height = 8)

QC %>% 
  	ggplot(aes(x=percent.ribo, y=percent.mt, color=sample_id)) + 
  	geom_point(size=1.5,alpha = 0.5) + scale_colour_manual(values =c(okabe_pal)) +
  	#stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = ribo_set,linetype="dotted")+#ribo
  geom_hline(yintercept = mito_set2,linetype="dotted")+#mito
   geom_hline(yintercept = mito_set1,linetype="dotted")#mito
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "percent_MT_RIBO_QC.pdf"),
     width = 10, height = 8)

QC %>%
  	ggplot(aes(x=log10GenesPerUMI, color = sample_id, fill=sample_id)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 0.85)
#rownames(QC)=QC$info
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "log10GenesPerUMI_QC.pdf"),
     width = 10, height = 8)

plot1 <- FeatureScatter(data_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt",group.by = "sample_id")+
  geom_hline(yintercept = mito_set2,linetype="dotted")+geom_hline(yintercept = 1,linetype="dotted")+ scale_colour_manual(values =c(okabe_pal))
plot2 <- FeatureScatter(data_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by = "sample_id")+
  geom_hline(yintercept = nGene1,linetype="dotted")+geom_hline(yintercept = nGene2,linetype="dotted")+ scale_colour_manual(values =c(okabe_pal))
plot3 <- FeatureScatter(data_seurat, feature1 = "percent.ribo", feature2 = "percent.mt",group.by = "sample_id")+geom_vline(xintercept = ribo_set,linetype="dotted")+
  geom_hline(yintercept = mito_set2,linetype="dotted")+geom_hline(yintercept = mito_set1,linetype="dotted")+ scale_colour_manual(values =c(okabe_pal))
plot1 + plot2+ plot3
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "all_QC_final.pdf"),
     width = 20, height = 8)

```
#QC library
####do filter cuts

```{r}
#SET variables:
mito_set1=0
mito_set2=10
ribo_set=1
nGene1=300
nGene2=7000
nUMIset=100
######################################
#FILTER CELLS by QC metrics 
filtered_seurat <- subset(data_seurat, subset = nFeature_RNA>nGene1 & nFeature_RNA<nGene2 & nCount_RNA>nUMIset & percent.mt>=mito_set1 & percent.mt<mito_set2& percent.ribo>=ribo_set)

metadata=filtered_seurat[[]]
counts <- GetAssayData(object = filtered_seurat, slot = "counts")

write.csv(counts, paste0(fname_prefix_csv_QC,"_silencing_matrix_all_samples_AB017.csv",sep=""))
write.csv(metadata, paste0(fname_prefix_csv_QC,"_cell_metadata_filtered_AB017.csv",sep=""))

post_QCcounts=metadata%>%dplyr::group_by(sample_id)%>%dplyr::summarise(count = n())
post_QCcounts$tr_nGene=paste0(nGene1," genes per cell and less than ",nGene2)
post_QCcounts$tr_n_count_UMI=paste0(nUMIset," UMI gene")
post_QCcounts$tr_mito=paste0("more than or equal",mito_set1,"% and less than ",mito_set2,"%")
post_QCcounts$tr_ribo=paste0("more than or equal",ribo_set,"%")
colnames(pre_QCcounts)=c("sample_id","pre_QC_count")
post_QCcounts=post_QCcounts%>%left_join(pre_QCcounts,by=c("sample_id"))
post_QCcounts$lost=post_QCcounts$pre_QC_count-post_QCcounts$count
write.csv(post_QCcounts, file= paste0(fname_prefix_csv_QC, "_", "post_QCcounts.csv"), row.names=T)

#QCcounts=pre_QCcounts%>%right_join(post_QCcounts,by=c("sample_id"))
#QCcounts=QCcounts%>%dplyr::rename(Pre_QC="count.x",Post_QC="count.y")
#QCcounts$lost=QCcounts$Pre_QC-QCcounts$Post_QC
#write.csv(QCcounts, file= paste0(fname_prefix_csv_QC, "_", "QCcounts.csv"), row.names=T)


Rdata_filename = paste0(fname_prefix_R_QC, "_seurat_data_filtered.RData")
save(meta_original, counts, metadata,filtered_seurat,aggr_info,
     file = Rdata_filename)


plot1 <- FeatureScatter(filtered_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt",group.by = "sample_id")+
  geom_vline(xintercept = nUMIset,linetype="dotted")+
  geom_hline(yintercept = mito_set1,linetype="dotted")+
  geom_hline(yintercept = mito_set2,linetype="dotted")+ scale_colour_manual(values =c(pal))

plot2 <- FeatureScatter(filtered_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by = "sample_id")+
  geom_vline(xintercept = nUMIset,linetype="dotted")+
  geom_hline(yintercept = nGene1,linetype="dotted")+
  geom_hline(yintercept = nGene2,linetype="dotted")+ scale_colour_manual(values =c(pal))

plot3 <- FeatureScatter(filtered_seurat, feature1 = "percent.ribo", feature2 = "percent.mt",group.by = "sample_id")+geom_vline(xintercept = 4,linetype="dotted")+
  #geom_hline(yintercept = 1,linetype="dotted")+
  geom_hline(yintercept = mito_set1,linetype="dotted")+
  geom_hline(yintercept = mito_set2,linetype="dotted")+
  scale_colour_manual(values =c(pal))

plot1 + plot2+ plot3
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "all_QC_after_FILTER.pdf"),
     width = 20, height = 8)
```


#add prepare data for Monocle

```{r}
####gather info from Seurat
cell_metadata=metadata
expression_matrix <- filtered_seurat@assays[["RNA"]]@layers[["counts"]]
counts <- GetAssayData(object = filtered_seurat, slot = "counts")

# get gene annotations
gene_annotation=data.frame(rownames(counts))
colnames(gene_annotation) <- "gene_short_name"
rownames(gene_annotation)=gene_annotation$gene_short_name

```

####preprocess monocle data

```{r}

###create cds file
cds <- new_cell_data_set(counts,cell_metadata = cell_metadata,
                                      gene_metadata = gene_annotation)
cds <- preprocess_cds(cds, num_dim = 30)
#cds <- align_cds(cds, num_dim = 100, alignment_group = "sample")
plot_pc_variance_explained(cds)

ggsave(filename = paste0(fname_prefix_dotplot, "_", "variance.png"),
     width = 12, height = 8)

#cds <- align_cds(cds, alignment_group = "experiment")
set.seed(678686)
cds = reduce_dimension(cds,reduction_method="tSNE")
#set.seed(79876858)
cds = reduce_dimension(cds,umap.min_dist = 0.2,reduction_method="UMAP")
#cds = reduce_dimension(cds,umap.min_dist = 0.5,reduction_method="UMAP")

#monocle.object = order_cells(monocle.object, reduction_method = "UMAP")
#set.seed(12345)
cds_1 = cluster_cells(cds, resolution = 1e-2)
colData(cds_1)$monocle_clusters = as.character(monocle3::clusters(cds_1))
#set.seed(123479870)
cds_2 = cluster_cells(cds, resolution = 1e-4)
colData(cds_2)$monocle_clusters = as.character(monocle3::clusters(cds_2))
#set.seed(123457768)
#cds_3 = cluster_cells(cds, resolution = 5e-5)
cds_3 = cluster_cells(cds, k=15, resolution = 2e-4)
colData(cds_3)$monocle_clusters = as.character(monocle3::clusters(cds_3))
#set.seed(123458980)
cds_4 = cluster_cells(cds, resolution = 2e-4)
colData(cds_4)$monocle_clusters = as.character(monocle3::clusters(cds_4))
#set.seed(1234597698)
cds_5 = cluster_cells(cds, resolution = 5e-4)
colData(cds_5)$monocle_clusters = as.character(monocle3::clusters(cds_5))


plot_cells(cds_1, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_1_monocle_clusters_resolution_1e_2.pdf"),
     width = 20, height = 4)
plot_cells(cds_2, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_2_monocle_clusters_resolution_1e_4.pdf"),
     width = 20, height = 4)
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_3_monocle_clusters_resolution_2e_4_K15.pdf"),
     width = 20, height = 4)
plot_cells(cds_4, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_4_monocle_clusters_resolution_2e_4.pdf"),
     width = 20, height = 4)
plot_cells(cds_5, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_5_monocle_clusters_resolution_5e_4.pdf"),
     width = 20, height = 4)

plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_4_monocle_clusters_resolution_2e_4_K15_single.pdf"),
     width = 4, height = 4)
plot_cells(cds_4, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_4_monocle_clusters_resolution_2e_4_single.pdf"),
     width = 4, height = 4)
plot_cells(cds_5, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
ggsave(filename = paste0(fname_prefix_UMAP, "_cds_5_monocle_clusters_resolution_5e_4_single.pdf"),
     width = 4, height = 4)


Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered.RData")
save(cell_metadata, counts,cds, cds_1,
     cds_2, cds_3, cds_4, cds_5,
     file = Rdata_filename)
```

###chose the cluster iteration with better results and clean environment
#either 3 or 4 

```{r}
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8,show_trajectory_graph=F)
#plot_cells(cds, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8,show_trajectory_graph=F,label_cell_groups=F)
#last_plot()+facet_grid(~sample_id)
last_plot()+facet_grid(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP.pdf"),
     width = 10, height = 5)

#map by seurat clusters generated
cds_3 <- learn_graph(cds_3)
cds=cds_3#this is the iteration we decided to use

#cds_cardio= cds[,cds@colData@listData[["monocle_clusters"]]=="5"|cds@colData@listData[["monocle_clusters"]]=="4"|cds@colData@listData[["monocle_clusters"]]=="3"|cds@colData@listData[["monocle_clusters"]]=="2"]

cds@colData$samp=paste0(cds@colData$type,"_",cds@colData$media)

Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered_FINAL.RData")
save(cell_metadata, counts,cds,
     file = Rdata_filename)

#clean environment
rm(combined_exp,data_exp5,expression_data,gene_annotation,expression_matrix,cds_1,cds_2,cds_3,cds_4,cds_5)
```

###select cds with protein coding genes
#add umap data on cds meta
#add pseudotime info

```{r}
#add UMAP data to cell_metadata
pbuild <- plot_cells(cds, color_cells_by = "monocle_clusters", label_cell_groups=T, show_trajectory_graph = F, label_leaves=F, label_branch_points=F, graph_label_size=3, group_label_size = 4, cell_size = 1) 
UmapCDS <- data.frame(pbuild$data$sample_name,UMAP1_monocle = pbuild$data$data_dim_1, UMAP2_monocle = pbuild$data$data_dim_2, row.names = 1) 
colData(cds)$UMAP_1_monocle=pbuild$data$data_dim_1 #add umap data for further analysis outside monocle
colData(cds)$UMAP_2_monocle=pbuild$data$data_dim_2 #add umap data for further analysis outside monocle

original_counts=counts
#counts=as.data.frame(exprs(cds))
counts=as.data.frame(normalized_counts(cds))
meta=as.data.frame(cds@colData@listData)

#cds<- order_cells(cds,reduction_method = "UMAP")#selected around cluster 3

#plot_cells(cds,
 #          color_cells_by = "pseudotime",
    #       label_cell_groups=FALSE,
   #        label_leaves=FALSE,
  #         label_branch_points=FALSE,
 #          graph_label_size=1.5)
#last_plot()+facet_grid(~sample)
#ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_pseudotime.pdf"),
 #    width = 18, height = 8)

Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered_FINAL.RData")
save(cell_metadata, counts,cds,meta,original_counts,
     file = Rdata_filename)


```

###make graphs for clusters and features
##clean up from a cluster before doing the last saving and Seurat?

```{r}

#eliminated cl7 since it is mainly cells with lower gene numbers

#select all clusters except 7 from cds

#cds= cds[,cds@colData@listData[["monocle_clusters"]]!="7"]

original_counts=as.data.frame(exprs(cds))
counts=as.data.frame(normalized_counts(cds))
meta=as.data.frame(cds@colData@listData)


#cl_pal=c("#000000","#004949", "#009E73", "#CC79A7", "#D55E00", "#f0E442", "#0072B2")

cl_pal=c("#000000","#56B4E9", "#009E73","#004949","#24ff24", "#CC79A7", "#D55E00", "#f0E442", "#0072B2","#b6dbff","#ff6db6","#924900","#920000","#ffb6db")


#cl_pal=c("#000000","#E69F00", "#009E73", "#CC79A7", "#D55E00", "#f0E442", "#0072B2")

ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = monocle_clusters),
               size = 0.8,alpha=0.7)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clusters")+
  scale_color_manual(name = "monocle",
                        values=c(cl_pal))
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP.pdf"),
     width = 5, height = 5)

ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = monocle_clusters),
               size = 0.8,alpha=0.7)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clusters")+
  scale_color_manual(name = "monocle",
                        values=c(cl_pal))+facet_wrap(~factor(samp,levels = c("WT_E8","WT_E6_FGF","NANOG_OVER_E8","NANOG_OVER_E6_FGF")))
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_monocle_clusters.pdf"),
     width = 6, height = 6)


ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = monocle_clusters),
               size = 0.8,alpha=0.7)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "replicate")+
  scale_color_manual(name = "monocle",
                        values=c(cl_pal))+facet_wrap(~sample_id)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_samp.pdf"),
     width = 12, height = 10)
  
plot_feature=c("percent.mt", "percent.ribo","nFeature_RNA","nCount_RNA")

for (feature in plot_feature) {
  
  p <- ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes_string(color = feature), size = 0.8) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = feature
    ) +
    scale_color_viridis(option = "E", discrete = FALSE) #+
    #facet_grid(~sample_id)
  
  ggsave(
    filename = paste0(fname_prefix_umap_QC, "_", "UMAP_", feature, ".pdf"),
    plot = p,
    width = 5,
    height = 5
  )
}

####cluster 5 may be due to quality more than biology
#cds= cds[,cds@colData@listData[["monocle_clusters"]]!="5"]
#counts=as.data.frame(normalized_counts(cds))
#meta=as.data.frame(cds@colData@listData)
```


###get top genes in a loop 

```{r}
marker_test_res = top_markers(cds, group_cells_by="monocle_clusters", reference_cells=NULL, cores=16)
marker_test_res <- top_markers(cds, group_cells_by="monocle_clusters", 
                               reference_cells=10000, cores=8)

# List of top_n values to iterate through
top_n_values <- c(2, 5, 10, 25)

# Loop through each top_n value
for (n in top_n_values) {
  
  # Filter markers and select the top_n genes
  top_specific_markers <- marker_test_res %>%
    filter(fraction_expressing >= 0.10) %>%
    group_by(cell_group) %>%
    top_n(n, pseudo_R2)
  
  # Extract unique gene IDs
  top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
  
  # Create a data frame for the top genes
  t_data <- data.frame(
    ids = top_specific_markers$gene_id, 
    cluster = top_specific_markers$cell_group, 
    stringsAsFactors = FALSE
  )
  
  # Write the data to a CSV file
  write.csv(
    t_data, 
    file = paste0(fname_prefix_csv, "_", "top_", n, "genes_per_cluster.csv"), 
    row.names = FALSE
  )
  
  # Generate a heatmap plot
  p <- plot_genes_by_group(
    cds,
    top_specific_marker_ids,
    group_cells_by = "monocle_clusters",
    ordering_type = "maximal_on_diag",
    max.size = 3
  ) + scale_color_viridis(option = "E", discrete = FALSE)
  
  # Save the plot
  ggsave(
    filename = paste0(fname_prefix_heatmap, "_", "top", n, "_heatmap.pdf"),
    plot = p,
    width = 8, 
    height = ifelse(n == 5, 10, 25) # Adjust the height for larger plots
  )
}

```


##get top 10 genes per cluster

```{r}

marker_test_res = top_markers(cds, group_cells_by="monocle_clusters", reference_cells=NULL, cores=16)
marker_test_res <- top_markers(cds, group_cells_by="monocle_clusters", 
                               reference_cells=10000, cores=8)

top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(10, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t10=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t10)=c("ids", "cluster")

write.csv(t10, file= paste0(fname_prefix_csv, "_", "top_10genes_per_cluster.CSV"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="monocle_clusters",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_heatmap, "_", "top10_heatmap.pdf"),
     width = 8, height = 15)

```

##make graphs to visualize genes

```{r}

#load("scratch/monocle_AB04.RData") #it was 251124_monocle_cds_clustered_FINAL.RData

plot_cells(cds,
           genes = c("MB","MYOD1","MYOG","CDX2","PAX7","MYH7","MYH5","MYH4","DMD","SOX2","MKI67","AURKB"),
            label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
 
 ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_my.pdf"),
     width = 8, height = 6)

plot_cells(cds,
           genes = c("MKI67","PCNA","TOP2A","MCM5","CCNB1","CCNB2","CCNA2","CDC20","BIRC5","UBE2C","AURKA","AURKB"),
            label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
 
 ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_proliferation.pdf"),
     width = 8, height = 6)
 

plot_cells(cds,
           genes = c("SOX2","NODAL","SOX4","SOX3","ZIC1","BMP6","BMP4","BMP2","CDH1","EGFR","GATA6","FOX17","FOXA2","VIM","COL1A1"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_transition.pdf"),
     width = 8, height = 6)
   
plot_cells(cds,
           genes = c("EOMES", "LEFTY1", "LEFY2", "NANOG", "GDF3", "MIXL1", "NODAL", "CER1", "FGF8", "HEY2", "SOX17", "GSC"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_TGFbeta_targets.pdf"),
     width = 8, height = 6)
   
plot_cells(cds,
           genes = c("CDX2", "SMAD6", "WNT8A", "EGFLAM", "MEIS2", "CITED2", "SOX2", "GDF11", "NSUN7", "NDGR3", "HOXB2"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_TGFbeta_neg.pdf"),
     width = 8, height = 6)
   
plot_cells(cds,
           genes = c("POU5F1", "WNT3", "PRDM14", "MYC", "POL3G", "DNM3B", "ETV5", "EPCAM", "DPPA4", "LZTS1", "FOXA2", "PCH1", "GLIPR1L1"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_TGFbeta_nontargets.pdf"),
     width = 8, height = 6)
   
plot_cells(cds,
           genes = c("ZIC1", "KHDRBS2", "UNC5D", "SIX3", "ZNF528", "TFAP2C"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_iPSCneuro.pdf"),
     width = 8, height = 6)
   
plot_cells(cds,
           genes = c("NODAL","ZIC1","LEFTY1","LEFTY2","CER1","EOMES","MIXL1","GSC","ZIC2","ZIC3","OTX2","SOX11","LHX2","SOX1","PAX6", "UNC5D","LRRC75A","ZNF138"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_SMAD2_epi_meso_neuro.pdf"),
     width = 8, height = 6)

plot_cells(cds,
           genes = c("SMARCD3"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
  labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = "SMARCD3 expression"
    )+
     scale_color_viridis(option="E", discrete=F)+facet_wrap(~sample_id)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_SMARCD3.pdf"),
     width = 8, height = 6)
   
   plot_cells(cds,
           genes = c("KDM6B"),
           label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
  labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = "JMJD3 expression"
    )+
     scale_color_viridis(option="E", discrete=F)+facet_wrap(~sample_id)
   ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_JMJ.pdf"),
     width = 8, height = 6)
   
   

```
#####make gene lists for muscle cell types


```{r}
gs <- list(

  # ===============================
  # CELL CYCLE / RIBOSOME TONE
  # ===============================
  cellcycle_s = c("MCM5","PCNA","TYMS","FEN1","MCM2","MCM4","RRM1","UNG","GINS2",
                  "MCM6","CDCA7","DTL","PRIM1","UHRF1","HELLS","RFC2","RPA2","NASP",
                  "RAD51","CHEK1","ORC6","MCM3","MCM7","MLF1IP","BRIP1","E2F8"),
  # Ref: Seurat S.Score; Whitfield et al., Cell 2002; Trapnell et al., Nat Biotech 2014

  cellcycle_g2m = c("HMGB2","CDK1","NUSAP1","UBE2C","BIRC5","TPX2","TOP2A","NDC80",
                    "CKS2","NUF2","CKS1B","MKI67","TMPO","CENPF","TACC3","SMC4",
                    "CCNB2","CKAP2L","CKAP2","AURKB","BUB1","KIF11","ANP32E","TUBB4B",
                    "GTSE1","KIF20B","HJURP","CDCA3","CDC20","TTK","CDC45","CDC6",
                    "EXO1","TIPIN","DSCC1","BLM","CASP8AP2","USP1","CLSPN","POLA1",
                    "CHAF1B","BRCA1","EXOSC2","USP16","POLD3","MSH2","ATAD2","RAD51AP1","RRM2"),
  # Ref: Seurat G2M.Score; Whitfield et al., Cell 2002; Trapnell 2014
  
  ribosome_tone = c("RPLP0","RPL3","RPL5","RPL7","RPL11","RPL13A","RPL23A",
                    "RPS3","RPS6","RPS8","RPS12","RPS18","RPS27","NCL","FBL",
                    "NPM1","EIF4EBP1","MYC","DDIT4"),
  # Ref: Murgia et al. 2021; Chemello et al. 2025 (fiber proteome core)

  # ===============================
  # FIBER-TYPE / METABOLIC PROGRAMS
  # ===============================
  fiber_type_I_slow = c("MYH7","TNNI1","TNNT1","MYL2","ATP2A2","CKMT2","MB",
                        "COX7A1","COX6A2","PPARGC1A","SDHA","CS","SLC25A4",
                        "MYBPC1","MYL3","PDK4","ESRRA","NDUFA10"),
  # Ref: Schiaffino & Reggiani 2011; Chemello et al. 2025; Murgia et al. 2021

  fiber_type_IIa_fast = c("MYH2","TNNI2","TNNT3","MYL1","ATP2A1","ACTN3","TPM2",
                          "PVALB","LDHA","PKM","PFKM","ENO1","PGAM2","ALDOA",
                          "CKM","DMD","CASQ1"),
  # Ref: Murgia et al. 2021; Bloemberg & Quadrilatero 2012; Chemello 2025

  fiber_type_IIx_fast = c("MYH1","ACTN3","MYLK2","TNNT3","TNNI2","ATP2A1","TPM2",
                          "CKM","LDHA","ENO1","PFKM","MYH4"),
  # Ref: Murgia 2021; Uhlén et al., HPA 2021

  developmental_embryonic_perinatal = c("MYH3","MYH8","MYL4","TNNI1","TNNT1",
                                        "MYOG","MYOD1","MEF2C","DES","MYMK",
                                        "MYMX","CHRNG","ACTA1","CKM"),
  # Ref: Chal et al. 2015; Hicks et al. 2025; Sampath et al. 2018

  sarcomere_core = c("TTN","NEB","ACTN2","TCAP","MYBPC1","MYBPC2","MYL1","MYL2",
                     "TNNI1","TNNI2","TNNT1","TNNT3","TPM1","TPM2","MYOM1","MYOM2","CSRP3"),
  # Ref: Lundberg et al. 2019; GO:0030016 "myofibril"

  calcium_excitation_contraction = c("RYR1","CACNA1S","ATP2A1","ATP2A2","CASQ1",
                                     "CASQ2","SLN","PLN","TRDN","STAC3","JPH1",
                                     "CAMK2A","CAMK2B","CAV3"),
  # Ref: Protasi et al. 2021; Jorgensen et al. 2023 (hiPSC EC coupling)

  oxidative_mitochondrial = c("SDHA","SDHB","UQCRC1","NDUFA9","NDUFS1","COX4I1",
                              "COX6A2","ATP5F1A","ATP5MC1","CKMT2","SLC25A4",
                              "PPARGC1A","TFAM","CS","ACADM","NDUFB8","CYCS","PDHA1"),
  # Ref: Lin et al. Nature 2002; Chemello et al. 2025; Murgia 2021

  glycolytic_fast = c("HK2","PFKM","ALDOA","TPI1","GAPDH","PGK1","PGAM2","ENO1",
                      "PKM","LDHA","PDK1","SLC2A4","PFKFB3","ENO3","MYL1"),
  # Ref: Schiaffino & Reggiani 2011; Chemello 2025

  # ===============================
  # STEM / MYOGENIC LINEAGES
  # ===============================
  satellite_cell_core = c("PAX7","VCAM1","ITGA7","CXCR4","MYF5","CDH15","SDC4",
                          "EGFR","NCAM1","SPRY1","MUSTN1","NOS1"),
  # Ref: De Micheli et al. 2020; Yin et al. 2013; GSE143437 atlas

  myoblast_proliferation_cycle = c("MYOD1","MYF5","MKI67","TOP2A","PCNA","CCNB1",
                                   "CDK1","E2F1","NUSAP1","HMGB2","UBE2C","AURKB","RRM2"),
  # Ref: Hicks et al. 2025; Trapnell et al. 2014 pseudotime

  myotube_differentiation_fusion = c("MYOG","MEF2C","MYMK","MYMX","MYH3","MYH8",
                                     "CKM","DES","DNM2","ACTA1"),
  # Ref: Chal et al. 2015; Sampath et al. 2018; Hicks 2025

  myogenic_regulators_axis = c("PAX3","PAX7","MYF5","MYOD1","MYOG","MEF2C","MYF6"),
  # Ref: Weintraub 1991; Rudnicki et al. 1993

  # ===============================
  # NMJ / ECM / NICHE
  # ===============================
  neuromuscular_junction = c("AGRN","LRP4","MUSK","DOK7","RAPSN","COLQ","CHRNA1",
                             "CHRNB1","CHRND","CHRNE","CHRNG","ACHE","UTRN"),
  # Ref: Burden et al. 2018; Cantor et al. 2022 (hiPSC NMJ coculture)

  muscle_ecm_basement = c("LAMA2","LAMC1","COL6A1","COL6A2","COL6A3","COL1A1",
                          "COL1A2","COL3A1","FN1","SPP1","LUM","DCN","DAG1",
                          "ITGA7","THY1","COL4A1","COL4A2","COL5A1"),
  # Ref: De Micheli et al. 2020; Saini et al. 2021

  fibro_adipogenic_progenitors = c("PDGFRA","COL1A1","COL3A1","DCN","LUM","CXCL14",
                                   "THY1","GPC6","FBN1","DPT"),
  # Ref: De Micheli 2020; Front Cell Dev Biol 2025

  atrophy_stress = c("FBXO32","TRIM63","DDIT4","GADD45A","ATF4","FOXO3","PSMA1",
                     "PSMB1","SESN1","SESN3","EIF4EBP1","PPP1R15A"),
  # Ref: Bodine et al. 2001; Gomes et al. 2001; Milan et al. 2021

  # ===============================
  # PLURIPOTENCY → MYOGENIC (hiPSC/hESC)
  # ===============================
  pluripotency_core = c("POU5F1","NANOG","SOX2","PRDM14","DPPA2","DPPA4","ZFP42",
                        "DNMT3A","DNMT3B"),
  # Ref: Theunissen et al. 2016; Nichols & Smith 2012

  naive_pluripotency = c("KLF17","KLF4","TFCP2L1","DNMT3L","DPPA3","TBX3","ESRRB","SUSD2"),
  # Ref: Collier et al. 2022; Pastor et al. 2018

  mesoderm_specification = c("MESP1","PDGFRA","KDR","ISL1","HAND1","HAND2","GATA4",
                             "TBX5","NKX2-5","MEF2C","FGF8","EOMES","TBXT"),
  # Ref: Chan et al. 2013; Loh et al. 2016; Hicks 2025

  early_myogenic_commitment = c("PAX3","PAX7","MYF5","MYOD1","MET","ITGA7","FGFR4","LBX1","MYOG")
  # Ref: Chal et al. 2015; Hicks et al. 2025
)
```

####make plots

```{r}
for (sig_name in names(gs)) {
  
  message("Plotting gene set: ", sig_name)
  
  p <- plot_cells(
    cds,
    genes = gs[[sig_name]],
    label_cell_groups = FALSE,
    show_trajectory_graph = FALSE
  ) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = sig_name
    ) +
    scale_color_viridis(option = "E", discrete = FALSE)
  
  print(p)
  
  ggsave(
    filename = paste0(fname_prefix_dotplot, "_UMAP_", sig_name, ".pdf"),
    plot = p,
    width = 8,
    height = 6
  )
}
```


###convert to Seurat and add cell cycle feature

```{r}
#load("scratch/monocle_AB04.RData") #it was 251203_monocle_cds_clustered_FINAL.RData

cell_metadata=as.data.frame(cds@colData@listData)
seurat_AB017<-as.Seurat(cds, assay = NULL)
###add cell cycle scores
exp.mat <- read.table(file = "scratch/cell_cicle_reg_genes/nestorawa_forcellcycle_expressionMatrix.txt", header = TRUE, 
    as.is = TRUE, row.names = 1)
# We can# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes #s phase genes
g2m.genes <- cc.genes$g2m.genes #G2phase genes

# Use fewer bins than 24 (e.g. 10 or 5)
seurat_AB017 <- CellCycleScoring(
  seurat_AB017,
  s.features   = s.genes,
  g2m.features = g2m.genes,
  set.ident    = TRUE,
  nbin         = 20   # or 5 if your dataset is very small
)

colData(cds)$S_score=seurat_AB017@meta.data[["S.Score"]]
colData(cds)$G2M_score=seurat_AB017@meta.data[["G2M.Score"]]
colData(cds)$Phase=seurat_AB017@meta.data[["Phase"]]

meta=as.data.frame(cds@colData@listData)###7603
cell_metadata=meta

meta_AB017=meta

ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = Phase),
               size=0.8)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "cell cycle phase")+
  scale_color_manual(name = "cell cycle phase",
                        values=c("#56B4E9","#009E73","#D55E00"))#+facet_grid(~factor(sample_id,levels = c("E8_R2","E8_R3","M8_R2","M8_R3","B8_R2","B8_R3")))
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_cellcycle_phase.pdf"),
     width = 5, height = 5)
  

plot_feature=c("S_score", "G2M_score")

for (feature in plot_feature) {
  
  p <- ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes_string(color = feature), size = 0.8) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = feature
    ) +
    scale_color_viridis(option = "E", discrete = FALSE) #+
    #facet_grid(~sample_id)
  
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", feature, ".pdf"),
    plot = p,
    width = 5,
    height = 5
  )
}

```

###add features
```{r}
fix_features <- function(object, genes) {
  present <- intersect(genes, rownames(object))
  if (length(present) == 0L) warning("No genes found from set: ", paste(head(genes, 5), collapse=", "))
  present
}

# Compute module scores for every set in `gs`
for (nm in names(gs)) {
  feats <- fix_features(seurat_AB017, gs[[nm]])
  # skip empty sets
  if (length(feats) == 0L) next
  seurat_AB017 <- AddModuleScore(
    object   = seurat_AB017,
    features = list(feats),        # one set at a time
    nbin = 20,                    # default is 24; adjust if needed
    name     = paste0(nm, "_score")# Seurat will append "1"
  )
  # Move the created column <name>1 into a clean column <name>
  raw_col <- paste0(nm, "_score1")
  if (raw_col %in% colnames(seurat_AB017@meta.data)) {
    seurat_AB017@meta.data[[nm]] <- seurat_AB017@meta.data[[raw_col]]
    # optional: drop the generated column
    seurat_AB017@meta.data[[raw_col]] <- NULL
  }
}

common_cells <- intersect(colnames(seurat_AB017), colnames(cds))
if (length(common_cells) == 0L) stop("No overlapping cells between Seurat and cds.")

for (nm in names(gs)) {
  if (nm %in% colnames(seurat_AB017@meta.data)) {
    vec <- seurat_AB017@meta.data[common_cells, nm, drop=TRUE]
    colData(cds)[common_cells, nm] <- vec
  }
}

feat_to_plot <- c(names(gs))

for (feat in feat_to_plot) {
  if (!feat %in% colnames(colData(cds))) next
  p <- plot_cells(
    cds,
    reduction_method   = "UMAP",
    color_cells_by     = feat,
    label_cell_groups  = TRUE,
    show_trajectory_graph = FALSE
  ) +
    labs(title = feat, x = "UMAP 1", y = "UMAP 2") +
    ggplot2::scale_color_viridis_c(option = "E")
  ggsave(paste0(fname_prefix_dotplot, "_UMAP_sign_", feat, ".pdf"), p, width = 4, height = 3)
}
```

#########################################################
######################SAVE DATA##########################
#########################################################

```{r}

Rdata_filename = paste0(fname_prefix_R, "_monocle_seurat_cds_clustered_FINAL_AB017.RData")#latest is of 20/01/2026
save(cell_metadata, counts,cds,meta,original_counts,seurat_AB017,gs,
     file = Rdata_filename)
Rdata_filename = paste0(fname_prefix_R, "_seurat_AB017.RData")
save(meta,seurat_AB017,gs,
     file = Rdata_filename)

####place them into the scratch folder, outputs from 09/04/2026 (eliminated date)
```



