---
title: "n2 AB017: DEG and cluster ED"
author: "Elisa Balmas PhD"
date: "2025-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment

```{r}
library(EnhancedVolcano)
library(monocle3)
library(tidyr)
library(stringr)
library(grid)
library(viridis)
library(viridisLite)
library(ggthemes)
library(dplyr)
library(tidyverse)
library(ggsignif)
library(umap)
library(heatmap3)
library(plyr)
library(edgeR)
library(compareGroups)
library(dbscan)
library(MAST)
library(geosphere)
library(RColorBrewer)
library(Seurat)
library(htmlwidgets)
library(Matrix)
library(tidyseurat)
library(SeuratWrappers)
library(scMAGeCK)
library(readxl)
library(rCASC)
library(deepToolsUtils)
library(R.utils)
library(plotly)


library(ggplot2); library(reshape2); theme_set(theme_bw(26) + theme(panel.grid.major = element_blank(), 
                                                                    panel.grid.minor = element_blank()) +
                                                 theme(legend.key = element_blank()))
update_geom_defaults("point", aes(size = 4))

#setwd("/home/ebalmas/scpHUB/projects/H001AS7_iKDseq_10X_4tp_monocle/analysis_EB")

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))


# Set up filename prefixes and output folders with the data

dir.create(file.path("Output","DEG"))#create forders MONOCLE
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "dotplots"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "dotplots","single"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "stat"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "UMAP"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "mosaic"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "heatmap"))
dir.create(file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "GO"))


#set up forder path MONOCLE
fname_prefix_R <- file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot <- file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "dotplots", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot_single <- file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "dotplots", "single", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_csv<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_stat<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "stat", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_UMAP<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "UMAP", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_mosaic<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "mosaic", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_heatmap<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "heatmap", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_GO<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "GO", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_violin<-file.path("Output","DEG", format(Sys.Date(), "%y%m%d"), "violin", 
                               format(Sys.Date(), "%y%m%d"))

fname_prefix_scratch="scratch/"


okabe_pal=c("#E69F00","#56B4E9","#009E73","#f0E442","#0072B2","#D55E00","#CC79A7","#000000")


pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
         "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
         "#920000","#924900","#db6d00","#24ff24","#ffff6d")

library(pals)
cols_all <- pals::alphabet(20)#extended colorblind friendly palette
# amethyst      blue   caramel    damson     ebony    forest     green  honeydew      iron      jade     khaki      lime   magenta      navy    orange      pink 
#"#F0A0FF" "#0075DC" "#993F00" "#4C005C" "#191919" "#005C31" "#2BCE48" "#FFCC99" "#808080" "#94FFB5" "#8F7C00" "#9DCC00" "#C20088" "#003380" "#FFA405" "#FFA8BB" 
# quagmire       red       sky turquoise 
#"#426600" "#FF0010" "#5EF1F2" "#00998F"
```

########load data

```{r}

load(paste0(getwd(),"/scratch/monocle_seurat_cds_clustered_FINAL_AB017.RData")) #it was 251124_monocle_seurat_cds_clustered_FINAL.RData



#cl_pal=c("#000000","#56B4E9", "#009E73", "#CC79A7", "#D55E00", "#f0E442", "#0072B2")

#cl_pal=c("#000000","#E69F00", "#009E73", "#CC79A7", "#D55E00", "#f0E442")

cl_pal=c("#000000","#56B4E9", "#009E73","#004949","#24ff24", "#CC79A7", "#D55E00", "#f0E442", "#0072B2")

cluster_tab=data.frame(monocle_cluster=c("1","2","3","4","5","6","7","8","9"),
                       cluster_color=cl_pal)

cluster_tab$cluster_name=c("1","2","3","4","5","6","7","8","9")#need to change when defined
cluster_tab$cluster_name_short=c("1","2","3","4","5","6","7","8","9")#need to change when defined


#palettes for clusters on other datasets

# 5-cluster palette (colors NOT in cl_pal6) ####from iPS2seq
cl_pal5 <- c("#94FFB5", "#56B4E9", "#999999", "#332288", "#117733")

# 4-cluster palette (again distinct from both cl_pal6 and cl_pal5) ####from B8 dataset
cl_pal4 <- c("#88CCEE", "#CC6677", "#DDCC77", "#44AA99")
```

####DEG lists

```{r}

seurat=seurat_AB017
Idents(seurat) <- "monocle_clusters"
DEG_all_clusters <- FindAllMarkers(seurat,only.pos = FALSE,
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_all_clusters$genes=rownames(DEG_all_clusters)#more genes because each gene can be associated to more than one cluster

##other muscle clusters (progenitors?) vs not muscle
DEG_cl1vs2_7 <- FindMarkers(
  seurat,only.pos = FALSE,
  ident.1 = "1", ident.2 = c("2","7"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl1vs2_7$genes=rownames(DEG_cl1vs2_7)
DEG_cl1vs2_7=DEG_cl1vs2_7%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl1vs2_7) <- c("geneID", "q_value_1vs2_7", "logFC_1vs2_7")

###condizione OPTIOX vs CIRIs
DEG_cl6vs3_4_5_8=FindMarkers(seurat,only.pos = FALSE, ident.1 = "6", ident.2 = c("3","4","5","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl6vs3_4_5_8$genes=rownames(DEG_cl6vs3_4_5_8)
DEG_cl6vs3_4_5_8=DEG_cl6vs3_4_5_8%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl6vs3_4_5_8) <- c("geneID", "q_value_6vs3_4_5_8", "logFC_6vs3_4_5_8")

#OPTIOX vs CIRI old
DEG_cl6vs4=FindMarkers(seurat, only.pos = FALSE, ident.1 = "6", ident.2 = c("4"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl6vs4$genes=rownames(DEG_cl6vs4)
DEG_cl6vs4=DEG_cl6vs4%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl6vs4) <- c("geneID", "q_value_6vs4", "logFC_6vs4")

###condizione OPTIOX vs CIRIs clones
DEG_cl6vs3_5_8=FindMarkers(seurat,only.pos = FALSE, ident.1 = "6", ident.2 = c("3","5","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl6vs3_5_8$genes=rownames(DEG_cl6vs3_5_8)
DEG_cl6vs3_5_8=DEG_cl6vs3_5_8%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl6vs3_5_8) <- c("geneID", "q_value_6vs3_5_8", "logFC_6vs3_5_8")

#CIRI vs clones
DEG_cl4vs3_5_8=FindMarkers(seurat, only.pos = FALSE, ident.1 = "4", ident.2 = c("3","5","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl4vs3_5_8$genes=rownames(DEG_cl4vs3_5_8)
DEG_cl4vs3_5_8=DEG_cl4vs3_5_8%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl4vs3_5_8) <- c("geneID", "q_value_4vs3_5_8", "logFC_4vs3_5_8")

#CIRI clone JMJ vs BAF define cl3
DEG_cl3vs5_8=FindMarkers(seurat, only.pos = FALSE, ident.1 = "3", ident.2 = c("5","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl3vs5_8$genes=rownames(DEG_cl3vs5_8)
DEG_cl3vs5_8=DEG_cl3vs5_8%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl3vs5_8) <- c("geneID", "q_value_3vs5_8", "logFC_3vs5_8")

#CIRI clone JMJ vs BAF define cl5
DEG_cl5vs3_8=FindMarkers(seurat, only.pos = FALSE, ident.1 = "5", ident.2 = c("3","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl5vs3_8$genes=rownames(DEG_cl5vs3_8)
DEG_cl5vs3_8=DEG_cl5vs3_8%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl5vs3_8) <- c("geneID", "q_value_5vs3_8", "logFC_5vs3_8")

#CIRI clone JMJ and BAF common define cl8
DEG_cl8vs3_5=FindMarkers(seurat, only.pos = FALSE, ident.1 = "8", ident.2 = c("3","5"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl8vs3_5$genes=rownames(DEG_cl8vs3_5)
DEG_cl8vs3_5=DEG_cl8vs3_5%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl8vs3_5) <- c("geneID", "q_value_8vs3_5", "logFC_8vs3_5")

DEG_cl1vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "1", ident.2 = c("2","3","4","5","6","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl1vsall$genes=rownames(DEG_cl1vsall)
DEG_cl1vsall=DEG_cl1vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl1vsall) <- c("geneID", "q_value_1vsall", "logFC_1vsall")


DEG_cl7_2vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = c("7","2"), ident.2 = c("1","4","3","5","6","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl7_2vsall$genes=rownames(DEG_cl7_2vsall)
DEG_cl7_2vsall=DEG_cl7_2vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl7_2vsall) <- c("geneID", "q_value_7_2vsall", "logFC_7_2vsall")

DEG_cl7_2vsall_m=FindMarkers(seurat,only.pos = FALSE, ident.1 = c("7","2"), ident.2 = c("4","3","5","6","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl7_2vsall_m$genes=rownames(DEG_cl7_2vsall_m)
DEG_cl7_2vsall_m=DEG_cl7_2vsall_m%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl7_2vsall_m) <- c("geneID", "q_value_7_2vsall_m", "logFC_7_2vsall_m")


DEG_cl2vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "2", ident.2 = c("5","1","3","4","6","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl2vsall$genes=rownames(DEG_cl2vsall)
DEG_cl2vsall=DEG_cl2vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl2vsall) <- c("geneID", "q_value_2vsall", "logFC_2vsall")

DEG_cl3vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "3", ident.2 = c("5","1","2","4","6","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl3vsall$genes=rownames(DEG_cl3vsall)
DEG_cl3vsall=DEG_cl3vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl3vsall) <- c("geneID", "q_value_3vsall", "logFC_3vsall")

DEG_cl4vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "4", ident.2 = c("5","1","2","3","6","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl4vsall$genes=rownames(DEG_cl4vsall)
DEG_cl4vsall=DEG_cl4vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl4vsall) <- c("geneID", "q_value_4vsall", "logFC_4vsall")

DEG_cl5vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "5", ident.2 = c("1","2","3","4","6","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl5vsall$genes=rownames(DEG_cl5vsall)
DEG_cl5vsall=DEG_cl5vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl5vsall) <- c("geneID", "q_value_5vsall", "logFC_5vsall")

DEG_cl6vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "6", ident.2 = c("1","2","3","4","5","7","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl6vsall$genes=rownames(DEG_cl6vsall)
DEG_cl6vsall=DEG_cl6vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl6vsall) <- c("geneID", "q_value_6vsall", "logFC_6vsall")

DEG_cl7vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "7", ident.2 = c("1","2","3","4","5","6","8","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl7vsall$genes=rownames(DEG_cl7vsall)
DEG_cl7vsall=DEG_cl7vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl7vsall) <- c("geneID", "q_value_7vsall", "logFC_7vsall")

DEG_cl8vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "8", ident.2 = c("1","2","3","4","5","6","7","9"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl8vsall$genes=rownames(DEG_cl8vsall)
DEG_cl8vsall=DEG_cl8vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl8vsall) <- c("geneID", "q_value_8vsall", "logFC_8vsall")

DEG_cl9vsall=FindMarkers(seurat,only.pos = FALSE, ident.1 = "9", ident.2 = c("1","2","3","4","5","6","7","8"),
  test.use = "wilcox",
  logfc.threshold = 0,
  min.pct = 0,
  min.cells.feature = 1,
  min.cells.group  = 1,
  features = rownames(seurat))
DEG_cl9vsall$genes=rownames(DEG_cl9vsall)
DEG_cl9vsall=DEG_cl9vsall%>%dplyr::select(genes,p_val_adj,avg_log2FC)
colnames(DEG_cl9vsall) <- c("geneID", "q_value_9vsall", "logFC_9vsall")


de_stats <- DEG_cl1vs2_7 %>%
  left_join(DEG_cl6vs3_4_5_8, by = "geneID") %>%
  left_join(DEG_cl6vs4,       by = "geneID") %>%
  left_join(DEG_cl6vs3_5_8,   by = "geneID") %>%
  left_join(DEG_cl4vs3_5_8,   by = "geneID") %>%
  left_join(DEG_cl3vs5_8,     by = "geneID") %>%
  left_join(DEG_cl5vs3_8,     by = "geneID") %>%
  left_join(DEG_cl8vs3_5,     by = "geneID") %>%
  left_join(DEG_cl1vsall,     by = "geneID") %>%
  left_join(DEG_cl2vsall,     by = "geneID") %>%
  left_join(DEG_cl3vsall,     by = "geneID") %>%
  left_join(DEG_cl4vsall,     by = "geneID") %>%
  left_join(DEG_cl5vsall,     by = "geneID") %>%
  left_join(DEG_cl6vsall,     by = "geneID") %>%
  left_join(DEG_cl7vsall,     by = "geneID") %>%
  left_join(DEG_cl8vsall,     by = "geneID") %>%
  left_join(DEG_cl9vsall,     by = "geneID") %>%
  left_join(DEG_cl7_2vsall,   by = "geneID") %>%
  left_join(DEG_cl7_2vsall_m, by = "geneID")


de_stats$dup=duplicated(de_stats$geneID)
de_stats=subset(de_stats,de_stats$dup!=T)%>%dplyr::select(-dup)


#deg_list <- list(
 # DEG_cl1vs2_7,
  #DEG_cl6vs3_4_5_8,
  #DEG_cl6vs4,
  #DEG_cl6vs3_5_8,
  #DEG_cl4vs3_5_8,
  #DEG_cl3vs5_8,
  #DEG_cl5vs3_8,
  #DEG_cl8vs3_5,
  #DEG_cl1vsall,
  #DEG_cl2vsall,
  #DEG_cl3vsall,
  #DEG_cl4vsall,
  #DEG_cl5vsall,
  #DEG_cl6vsall,
  #DEG_cl7vsall,
  #DEG_cl8vsall,
  #DEG_cl9vsall,
  #DEG_cl7_2vsall,
  #DEG_cl7_2vsall_m)

#de_stats <- Reduce(function(x, y) left_join(x, y, by = "geneID"), deg_list)
#de_stats <- de_stats %>% distinct(geneID, .keep_all = TRUE)

analysis_clusters <- de_stats

analysis_clusters=de_stats

stat_filename = paste0(fname_prefix_csv, "_complete_gene_reg_by_clusters.csv")
write.csv(analysis_clusters, stat_filename, quote = F, row.names = T, sep=",")

```

####add directions

```{r}
#Set the significance level cutoff
sig_level <- 0.01
#Set a fold-change cutoff
fc_level <- log2(2)
fc_level2 <- (log2(2))*-1
###################################
N=20
list=c("1vs2_7",
  "6vs3_4_5_8",
  "6vs4",
  "6vs3_5_8",
  "4vs3_5_8",
  "3vs5_8",
  "5vs3_8",
  "8vs3_5",
  "1vsall",
  "2vsall",
  "3vsall",
  "4vsall",
  "5vsall",
  "6vsall",
  "7vsall",
  "8vsall",
  "9vsall",
  "7_2vsall",
  "7_2vsall_m")

for (i in list) {
  cluster_name <- paste0("clust", i)
  
  # Create column names dynamically
  logFC_col <- paste0("logFC_", i)
  q_value_col <- paste0("q_value_", i)
  treshold_col <- paste0("treshold_cl", i)
  Log10pvalue_col <- paste0("Log10pvalue_cl", i)
  
  # Apply thresholding logic
  de_stats[[treshold_col]] <- ifelse(de_stats[[logFC_col]] >= fc_level & de_stats[[q_value_col]] <= sig_level, "up",
                                      ifelse(de_stats[[logFC_col]] <= fc_level2 & de_stats[[q_value_col]] <= sig_level, "down", "other"))
  
  de_stats[[Log10pvalue_col]] <- -log10(de_stats[[q_value_col]]+1e-305)
}

N <- 25
contrasts <- c("1vs2_7",
  "6vs3_4_5_8",
  "6vs4",
  "6vs3_5_8",
  "4vs3_5_8",
  "3vs5_8",
  "5vs3_8",
  "8vs3_5",
  "1vsall",
  "2vsall",
  "3vsall",
  "4vsall",
  "5vsall",
  "6vsall",
  "7vsall",
  "8vsall",
  "9vsall",
  "7_2vsall",
  "7_2vsall_m")

for (i in contrasts) {
  logFC_col       <- paste0("logFC_", i)
  q_value_col     <- paste0("q_value_", i)
  Log10pvalue_col <- paste0("Log10pvalue_cl", i)
  to_show_col     <- paste0("to_show_cl", i)

  de_stats[[Log10pvalue_col]] <- -log10(de_stats[[q_value_col]] + 1e-305)

  up_idx <- !is.na(de_stats[[logFC_col]]) &
            !is.na(de_stats[[q_value_col]]) &
            de_stats[[logFC_col]] >= fc_level &
            de_stats[[q_value_col]] <= sig_level

  down_idx <- !is.na(de_stats[[logFC_col]]) &
              !is.na(de_stats[[q_value_col]]) &
              de_stats[[logFC_col]] <= fc_level2 &
              de_stats[[q_value_col]] <= sig_level

  up_candidates <- which(up_idx)
  down_candidates <- which(down_idx)

  up_top <- up_candidates[
    order(de_stats[[Log10pvalue_col]][up_candidates], decreasing = TRUE)
  ][seq_len(min(N, length(up_candidates)))]

  down_top <- down_candidates[
    order(de_stats[[Log10pvalue_col]][down_candidates], decreasing = TRUE)
  ][seq_len(min(N, length(down_candidates)))]

  show_idx <- logical(nrow(de_stats))
  show_idx[up_top] <- TRUE
  show_idx[down_top] <- TRUE

  de_stats[[to_show_col]] <- ifelse(show_idx, de_stats$geneID, "")
}
stat_filename = paste0(fname_prefix_csv, "_gene_reg_with_directions.csv")
write.csv(de_stats, stat_filename, quote = F, row.names = T, sep=",")

###########
de_stats=de_stats%>% distinct(geneID, .keep_all = TRUE)#eliminate duplicates

# Loop through clusters to make all the volcano plots in bulk than use the one below to make specific ones
contrasts <- c("1vs2_7",
  "6vs3_4_5_8",
  "6vs4",
  "6vs3_5_8",
  "4vs3_5_8",
  "3vs5_8",
  "5vs3_8",
  "8vs3_5",
  "1vsall",
  "2vsall",
  "3vsall",
  "4vsall",
  "5vsall",
  "6vsall",
  "7vsall",
  "8vsall",
  "9vsall",
  "7_2vsall",
  "7_2vsall_m")

for (i in contrasts)  {
  logFC_col <- paste0("logFC_", i)
  treshold_col <- paste0("treshold_cl", i)
  Log10pvalue_col <- paste0("Log10pvalue_cl", i)
  q_value_col <- paste0("q_value_", i)
  to_show_col <- paste0("to_show_cl", i)

  volcano_plot <- ggplot(data = de_stats, aes(x = !!rlang::sym(logFC_col), y = !!rlang::sym(Log10pvalue_col))) +
    geom_point(aes(colour = as.factor(!!rlang::sym(treshold_col))), size = 1.5) +
    scale_colour_manual(values = c("up" = "red", "down" = "blue", "other" = "grey")) +
     ggrepel::geom_text_repel(
  data = de_stats %>%
    filter(!!rlang::sym(q_value_col) < 0.05),
  aes(label = !!rlang::sym(to_show_col)),
  size = 3,
  box.padding = unit(0.2, "lines"),
  point.padding = unit(0.2, "lines"),
  segment.color = "grey",
  segment.size = 0.2,max.overlaps =100
)+
    theme(legend.position = "none") +
    geom_hline(yintercept = (log10(sig_level) * -1),linetype="dotted") +
    geom_vline(xintercept = c(fc_level, fc_level2),linetype="dotted") +
    xlim(c(-100, 80)) + ylim(c(0, 350)) +
    xlab("log2 fold change") + ylab("-Log10 adj p-value")+
  annotate("text", x = -60, y = 350, label = paste("comparison cl ", i), col = "black", size = 4, hjust = 1, vjust = 1)# +
    # Add "other clusters" on the top left for clusters other than the current one
   # annotate("text", x = -4, y = 200, label = "Other Clusters", col = "black", size = 4, hjust = 0, vjust = 1)


  ggsave(
    filename = paste0(fname_prefix_dotplot, "_cl", i, ".pdf"),
    plot = volcano_plot,
    width = 8,
    height = 6
  )
}
```

###run clusterprofiler
#GO analysis on clusters

```{r}

####make gene list to study

###run clusterprofiler
#GO analysis on clusters

####make gene list to study

library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)

genes_6vs3_4_5_8_cl6       <- subset(de_stats, de_stats$treshold_cl6vs3_4_5_8 == "up")
genes_6vs3_4_5_8_cl3_4_5_8 <- subset(de_stats, de_stats$treshold_cl6vs3_4_5_8 == "down")

genes_6vs4_cl6             <- subset(de_stats, de_stats$treshold_cl6vs4 == "up")
genes_6vs4_cl4             <- subset(de_stats, de_stats$treshold_cl6vs4 == "down")

genes_6vs3_5_8_cl6         <- subset(de_stats, de_stats$treshold_cl6vs3_5_8 == "up")
genes_6vs3_5_8_cl3_5_8     <- subset(de_stats, de_stats$treshold_cl6vs3_5_8 == "down")

genes_4vs3_5_8_cl4         <- subset(de_stats, de_stats$treshold_cl4vs3_5_8 == "up")
genes_4vs3_5_8_cl3_5_8     <- subset(de_stats, de_stats$treshold_cl4vs3_5_8 == "down")

genes_3vs5_8_cl3           <- subset(de_stats, de_stats$treshold_cl3vs5_8 == "up")
genes_3vs5_8_cl5_8         <- subset(de_stats, de_stats$treshold_cl3vs5_8 == "down")

genes_5vs3_8_cl5           <- subset(de_stats, de_stats$treshold_cl5vs3_8 == "up")
genes_5vs3_8_cl3_8         <- subset(de_stats, de_stats$treshold_cl5vs3_8 == "down")

genes_8vs3_5_cl8           <- subset(de_stats, de_stats$treshold_cl8vs3_5 == "up")
genes_8vs3_5_cl3_5         <- subset(de_stats, de_stats$treshold_cl8vs3_5 == "down")

genes_1vs2_7_cl1           <- subset(de_stats, de_stats$treshold_cl1vs2_7 == "up")
genes_1vs2_7_cl2_7         <- subset(de_stats, de_stats$treshold_cl1vs2_7 == "down")

genes_7_2vsall_cl7_2       <- subset(de_stats, de_stats$treshold_cl7_2vsall == "up")
genes_7_2vsall_all         <- subset(de_stats, de_stats$treshold_cl7_2vsall == "down")

genes_7_2vsall_m_cl7_2     <- subset(de_stats, de_stats$treshold_cl7_2vsall_m == "up")
genes_7_2vsall_m_all_m     <- subset(de_stats, de_stats$treshold_cl7_2vsall_m == "down")

genes_1vsall               <- subset(de_stats, de_stats$treshold_cl1vsall == "up")
genes_2vsall               <- subset(de_stats, de_stats$treshold_cl2vsall == "up")
genes_3vsall               <- subset(de_stats, de_stats$treshold_cl3vsall == "up")
genes_4vsall               <- subset(de_stats, de_stats$treshold_cl4vsall == "up")
genes_5vsall               <- subset(de_stats, de_stats$treshold_cl5vsall == "up")
genes_6vsall               <- subset(de_stats, de_stats$treshold_cl6vsall == "up")
genes_7vsall               <- subset(de_stats, de_stats$treshold_cl7vsall == "up")
genes_8vsall               <- subset(de_stats, de_stats$treshold_cl8vsall == "up")
genes_9vsall               <- subset(de_stats, de_stats$treshold_cl9vsall == "up")

# Convert each data frame into a named sorted vector of logFC values
gene_dfs <- list(
  genes_6vs3_4_5_8_cl6       = genes_6vs3_4_5_8_cl6,
  genes_6vs3_4_5_8_cl3_4_5_8 = genes_6vs3_4_5_8_cl3_4_5_8,
  genes_6vs4_cl6             = genes_6vs4_cl6,
  genes_6vs4_cl4             = genes_6vs4_cl4,
  genes_6vs3_5_8_cl6         = genes_6vs3_5_8_cl6,
  genes_6vs3_5_8_cl3_5_8     = genes_6vs3_5_8_cl3_5_8,
  genes_4vs3_5_8_cl4         = genes_4vs3_5_8_cl4,
  genes_4vs3_5_8_cl3_5_8     = genes_4vs3_5_8_cl3_5_8,
  genes_3vs5_8_cl3           = genes_3vs5_8_cl3,
  genes_3vs5_8_cl5_8         = genes_3vs5_8_cl5_8,
  genes_5vs3_8_cl5           = genes_5vs3_8_cl5,
  genes_5vs3_8_cl3_8         = genes_5vs3_8_cl3_8,
  genes_8vs3_5_cl8           = genes_8vs3_5_cl8,
  genes_8vs3_5_cl3_5         = genes_8vs3_5_cl3_5,
  genes_1vs2_7_cl1           = genes_1vs2_7_cl1,
  genes_1vs2_7_cl2_7         = genes_1vs2_7_cl2_7,
  genes_7_2vsall_cl7_2       = genes_7_2vsall_cl7_2,
  genes_7_2vsall_all         = genes_7_2vsall_all,
  genes_7_2vsall_m_cl7_2     = genes_7_2vsall_m_cl7_2,
  genes_7_2vsall_m_all_m     = genes_7_2vsall_m_all_m,
  genes_1vsall               = genes_1vsall,
  genes_2vsall               = genes_2vsall,
  genes_3vsall               = genes_3vsall,
  genes_4vsall               = genes_4vsall,
  genes_5vsall               = genes_5vsall,
  genes_6vsall               = genes_6vsall,
  genes_7vsall               = genes_7vsall,
  genes_8vsall               = genes_8vsall,
  genes_9vsall               = genes_9vsall
)

logFC_cols <- c(
  genes_6vs3_4_5_8_cl6       = "logFC_6vs3_4_5_8",
  genes_6vs3_4_5_8_cl3_4_5_8 = "logFC_6vs3_4_5_8",
  genes_6vs4_cl6             = "logFC_6vs4",
  genes_6vs4_cl4             = "logFC_6vs4",
  genes_6vs3_5_8_cl6         = "logFC_6vs3_5_8",
  genes_6vs3_5_8_cl3_5_8     = "logFC_6vs3_5_8",
  genes_4vs3_5_8_cl4         = "logFC_4vs3_5_8",
  genes_4vs3_5_8_cl3_5_8     = "logFC_4vs3_5_8",
  genes_3vs5_8_cl3           = "logFC_3vs5_8",
  genes_3vs5_8_cl5_8         = "logFC_3vs5_8",
  genes_5vs3_8_cl5           = "logFC_5vs3_8",
  genes_5vs3_8_cl3_8         = "logFC_5vs3_8",
  genes_8vs3_5_cl8           = "logFC_8vs3_5",
  genes_8vs3_5_cl3_5         = "logFC_8vs3_5",
  genes_1vs2_7_cl1           = "logFC_1vs2_7",
  genes_1vs2_7_cl2_7         = "logFC_1vs2_7",
  genes_7_2vsall_cl7_2       = "logFC_7_2vsall",
  genes_7_2vsall_all         = "logFC_7_2vsall",
  genes_7_2vsall_m_cl7_2     = "logFC_7_2vsall_m",
  genes_7_2vsall_m_all_m     = "logFC_7_2vsall_m",
  genes_1vsall               = "logFC_1vsall",
  genes_2vsall               = "logFC_2vsall",
  genes_3vsall               = "logFC_3vsall",
  genes_4vsall               = "logFC_4vsall",
  genes_5vsall               = "logFC_5vsall",
  genes_6vsall               = "logFC_6vsall",
  genes_7vsall               = "logFC_7vsall",
  genes_8vsall               = "logFC_8vsall",
  genes_9vsall               = "logFC_9vsall"
)

sort_decreasing <- c(
  genes_6vs3_4_5_8_cl6       = TRUE,
  genes_6vs3_4_5_8_cl3_4_5_8 = FALSE,
  genes_6vs4_cl6             = TRUE,
  genes_6vs4_cl4             = FALSE,
  genes_6vs3_5_8_cl6         = TRUE,
  genes_6vs3_5_8_cl3_5_8     = FALSE,
  genes_4vs3_5_8_cl4         = TRUE,
  genes_4vs3_5_8_cl3_5_8     = FALSE,
  genes_3vs5_8_cl3           = TRUE,
  genes_3vs5_8_cl5_8         = FALSE,
  genes_5vs3_8_cl5           = TRUE,
  genes_5vs3_8_cl3_8         = FALSE,
  genes_8vs3_5_cl8           = TRUE,
  genes_8vs3_5_cl3_5         = FALSE,
  genes_1vs2_7_cl1           = TRUE,
  genes_1vs2_7_cl2_7         = FALSE,
  genes_7_2vsall_cl7_2       = TRUE,
  genes_7_2vsall_all         = FALSE,
  genes_7_2vsall_m_cl7_2     = TRUE,
  genes_7_2vsall_m_all_m     = FALSE,
  genes_1vsall               = TRUE,
  genes_2vsall               = TRUE,
  genes_3vsall               = TRUE,
  genes_4vsall               = TRUE,
  genes_5vsall               = TRUE,
  genes_6vsall               = TRUE,
  genes_7vsall               = TRUE,
  genes_8vsall               = TRUE,
  genes_9vsall               = TRUE
)

for (nm in names(gene_dfs)) {
  df <- gene_dfs[[nm]]

  if (is.null(df) || nrow(df) == 0) {
    gene_dfs[[nm]] <- numeric(0)
    next
  }

  fc_col <- logFC_cols[[nm]]
  dec <- sort_decreasing[[nm]]

  genes <- df[[fc_col]]
  names(genes) <- df$geneID

  genes <- genes[!is.na(genes)]
  genes <- genes[!is.na(names(genes)) & names(genes) != ""]
  genes <- genes[!duplicated(names(genes))]
  genes <- sort(genes, decreasing = dec)

  gene_dfs[[nm]] <- genes
}

list2env(gene_dfs, envir = .GlobalEnv)

gene_sets <- list(
  "6vs3_4_5_8_cl6_up"         = genes_6vs3_4_5_8_cl6,
  "6vs3_4_5_8_cl3_4_5_8_down" = genes_6vs3_4_5_8_cl3_4_5_8,
  "6vs4_cl6_up"               = genes_6vs4_cl6,
  "6vs4_cl4_down"             = genes_6vs4_cl4,
  "6vs3_5_8_cl6_up"           = genes_6vs3_5_8_cl6,
  "6vs3_5_8_cl3_5_8_down"     = genes_6vs3_5_8_cl3_5_8,
  "4vs3_5_8_cl4_up"           = genes_4vs3_5_8_cl4,
  "4vs3_5_8_cl3_5_8_down"     = genes_4vs3_5_8_cl3_5_8,
  "3vs5_8_cl3_up"             = genes_3vs5_8_cl3,
  "3vs5_8_cl5_8_down"         = genes_3vs5_8_cl5_8,
  "5vs3_8_cl5_up"             = genes_5vs3_8_cl5,
  "5vs3_8_cl3_8_down"         = genes_5vs3_8_cl3_8,
  "8vs3_5_cl8_up"             = genes_8vs3_5_cl8,
  "8vs3_5_cl3_5_down"         = genes_8vs3_5_cl3_5,
  "1vs2_7_cl1_up"             = genes_1vs2_7_cl1,
  "1vs2_7_cl2_7_down"         = genes_1vs2_7_cl2_7,
  "7_2vsall_cl7_2_up"         = genes_7_2vsall_cl7_2,
  "7_2vsall_all_down"         = genes_7_2vsall_all,
  "7_2vsall_m_cl7_2_up"       = genes_7_2vsall_m_cl7_2,
  "7_2vsall_m_all_m_down"     = genes_7_2vsall_m_all_m,
  "1vsall_up"                 = genes_1vsall,
  "2vsall_up"                 = genes_2vsall,
  "3vsall_up"                 = genes_3vsall,
  "4vsall_up"                 = genes_4vsall,
  "5vsall_up"                 = genes_5vsall,
  "6vsall_up"                 = genes_6vsall,
  "7vsall_up"                 = genes_7vsall,
  "8vsall_up"                 = genes_8vsall,
  "9vsall_up"                 = genes_9vsall
)

# background universe = all detectable symbols in de_stats
univ_map <- suppressMessages(
  bitr(unique(na.omit(de_stats$geneID)),
       fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
)
universe_entrez <- unique(univ_map$ENTREZID)

# set list ego and GO_list to store results
ego_list <- list()
GO_list  <- list()

for (nm in names(gene_sets)) {
  syms <- unique(names(gene_sets[[nm]]))
  if (!length(syms)) next

  id_map <- suppressMessages(
    bitr(syms, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  )
  entrez <- unique(id_map$ENTREZID)
  if (!length(entrez)) next

  ego <- enrichGO(
    gene          = entrez,
    OrgDb         = org.Hs.eg.db,
    keyType       = "ENTREZID",
    ont           = "BP",
    universe      = universe_entrez,
    pAdjustMethod = "BH",
    pvalueCutoff  = 1,
    readable      = TRUE
  )

  ego_list[[nm]] <- ego
  GO_list[[nm]]  <- as.data.frame(ego)
}

# Write every GO table to CSV
if (length(GO_list)) {
  for (nm in names(GO_list)) {
    if (is.null(GO_list[[nm]]) || !nrow(GO_list[[nm]])) next
    file <- paste0(fname_prefix_csv, "_GO_", nm, ".csv")
    write.csv(GO_list[[nm]], file, row.names = FALSE)
    message("Wrote: ", file)
  }
}

# Helper: safe save for ggplot objects
save_plot <- function(plot_obj, filename, width = 8, height = 6) {
  if (inherits(plot_obj, "ggplot")) {
    ggsave(filename = filename, plot = plot_obj, width = width, height = height)
    message("Saved plot: ", filename)
  } else {
    message("Skipping (not a ggplot): ", filename)
  }
}

# Make dotplots and barplots for each ego in ego_list
if (length(ego_list)) {
  for (nm in names(ego_list)) {
    ego <- ego_list[[nm]]
    if (is.null(ego)) next

    res_df <- tryCatch(as.data.frame(ego), error = function(e) NULL)
    if (is.null(res_df) || !nrow(res_df)) next

    p_dot <- tryCatch(
      dotplot(ego, showCategory = 15, x = "Count"),
      error = function(e) NULL
    )
    save_plot(
      p_dot,
      paste0(fname_prefix_GO, "_dotplot_GO_", nm, ".pdf"),
      width = 8, height = 8
    )

    p_bar <- tryCatch(
      barplot(ego, x = "FoldEnrichment", showCategory = 15),
      error = function(e) NULL
    )
    save_plot(
      p_bar,
      paste0(fname_prefix_GO, "_bar_GO_", nm, ".pdf"),
      width = 8, height = 4
    )
  }
}

Rdata_filename = paste0(fname_prefix_GO, "_GO_analysis_clusters.RData")
save(gene_sets, ego_list, GO_list, de_stats, file = Rdata_filename)

```
###run clusterprofiler
#GSEA analysis on clusters

```{r}
####make gene list to study

###run clusterprofiler
#GSEA analysis on clusters

library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(DOSE)

set_min_n_genes <- 5

# Build ranked gene lists from ALL genes in each contrast
contrast_cols <- c(
  "1vs2_7",
  "6vs3_4_5_8",
  "6vs4",
  "6vs3_5_8",
  "4vs3_5_8",
  "3vs5_8",
  "5vs3_8",
  "8vs3_5",
  "1vsall",
  "2vsall",
  "3vsall",
  "4vsall",
  "5vsall",
  "6vsall",
  "7vsall",
  "8vsall",
  "9vsall",
  "7_2vsall",
  "7_2vsall_m"
)

gene_sets_GSEA <- list()

for (ctr in contrast_cols) {
  fc_col <- paste0("logFC_", ctr)

  if (!fc_col %in% colnames(de_stats)) next

  df <- de_stats[, c("geneID", fc_col)]
  colnames(df) <- c("geneID", "logFC")

  df <- df[!is.na(df$geneID) & df$geneID != "", ]
  df <- df[!is.na(df$logFC), ]

  # remove duplicate genes, keep the row with largest absolute logFC
  df <- df[order(-abs(df$logFC)), ]
  df <- df[!duplicated(df$geneID), ]

  genes <- df$logFC
  names(genes) <- df$geneID
  genes <- sort(genes, decreasing = TRUE)

  gene_sets_GSEA[[ctr]] <- genes
}

# sanity check
stopifnot(all(sapply(gene_sets_GSEA, function(x) is.numeric(x) && !is.null(names(x)))))

# SYMBOL -> ENTREZ mapping
symbol_to_entrez <- suppressMessages(
  bitr(
    unique(unlist(lapply(gene_sets_GSEA, names))),
    fromType = "SYMBOL",
    toType   = "ENTREZID",
    OrgDb    = org.Hs.eg.db
  )
)

gsea_list   <- list()
GSEA_tables <- list()

for (nm in names(gene_sets_GSEA)) {
  ranks_sym <- gene_sets_GSEA[[nm]]

  df <- data.frame(
    SYMBOL = names(ranks_sym),
    rank   = as.numeric(ranks_sym),
    stringsAsFactors = FALSE
  )

  df2 <- merge(df, symbol_to_entrez, by = "SYMBOL")
  if (!nrow(df2)) next

  # deduplicate ENTREZ: keep the entry with the highest absolute rank
  df2 <- df2[order(-abs(df2$rank)), ]
  df2 <- df2[!duplicated(df2$ENTREZID), ]

  ranks_entrez <- df2$rank
  names(ranks_entrez) <- df2$ENTREZID
  ranks_entrez <- sort(ranks_entrez, decreasing = TRUE)

  if (length(ranks_entrez) < set_min_n_genes) next

  gsea <- tryCatch(
    gseGO(
      geneList      = ranks_entrez,
      OrgDb         = org.Hs.eg.db,
      keyType       = "ENTREZID",
      ont           = "BP",
      minGSSize     = set_min_n_genes,
      maxGSSize     = 500,
      pAdjustMethod = "BH",
      pvalueCutoff  = 1,
      nPermSimple   = 10000,
      verbose       = FALSE,
      by            = "fgsea"
    ),
    error = function(e) {
      message("GSEA failed for ", nm, ": ", e$message)
      NULL
    }
  )

  if (is.null(gsea)) next

  gsea_readable <- tryCatch(
    setReadable(gsea, OrgDb = org.Hs.eg.db, keyType = "ENTREZID"),
    error = function(e) gsea
  )

  gsea_list[[nm]]   <- gsea_readable
  GSEA_tables[[nm]] <- as.data.frame(gsea_readable)
}

# significant-only version for plotting
gsea_sig_list <- lapply(gsea_list, function(gsea_obj) {
  if (is.null(gsea_obj)) return(NULL)

  df <- tryCatch(as.data.frame(gsea_obj), error = function(e) NULL)
  if (is.null(df) || !nrow(df)) return(NULL)

  df_sig <- dplyr::filter(df, p.adjust < 0.05)
  if (!nrow(df_sig)) return(NULL)

  gsea_obj@result <- df_sig
  gsea_obj
})

# write GSEA tables
if (length(GSEA_tables)) {
  for (nm in names(GSEA_tables)) {
    if (is.null(GSEA_tables[[nm]]) || !nrow(GSEA_tables[[nm]])) next
    file <- paste0(fname_prefix_csv, "_GSEA_", nm, ".csv")
    write.csv(GSEA_tables[[nm]], file, row.names = FALSE)
    message("Wrote: ", file)
  }
}

# helper: save ggplot safely
.save_plot <- function(p, file, w = 8, h = 10) {
  if (!is.null(p) && inherits(p, "ggplot")) {
    ggsave(filename = file, plot = p, width = w, height = h)
    message("Saved: ", file)
  }
}

# dotplots for all GSEA results
plot_gsea_dotplots <- function(gsea_list, fname_prefix, show_k = 10, width = 8, height = 10) {
  if (!length(gsea_list)) {
    message("gsea_list is empty.")
    return(invisible(NULL))
  }

  for (nm in names(gsea_list)) {
    g <- gsea_list[[nm]]
    df <- tryCatch(as.data.frame(g), error = function(e) NULL)

    if (is.null(df) || !nrow(df)) {
      message("Skipping ", nm, " (no terms).")
      next
    }

    k <- min(show_k, nrow(df))

    p <- tryCatch(
      dotplot(g, showCategory = k, split = ".sign", orderBy = "NES") +
        facet_grid(. ~ .sign),
      error = function(e) {
        message("Plot error in ", nm, ": ", e$message)
        NULL
      }
    )

    out_pdf <- paste0(fname_prefix, "_GSEA_", nm, "_dotplot.pdf")
    .save_plot(p, out_pdf, w = width, h = height)
  }
}

plot_gsea_dotplots(gsea_list, fname_prefix = fname_prefix_GO, show_k = 15)

# helper: save enrichment plots for selected GO IDs
save_gsea_plots <- function(gsea_res, contrast_name, go_ids, prefix, width = 8, height = 6) {
  if (is.null(gsea_res)) return(invisible(NULL))

  res_df <- tryCatch(as.data.frame(gsea_res), error = function(e) NULL)
  if (is.null(res_df) || !nrow(res_df)) return(invisible(NULL))

  for (go_id in go_ids) {
    if (go_id %in% res_df$ID) {
      p <- tryCatch(
        gseaplot2(gsea_res, geneSetID = go_id, title = paste0(contrast_name, " - ", go_id)),
        error = function(e) NULL
      )

      if (!is.null(p)) {
        out_file <- paste0(prefix, "_", contrast_name, "_", gsub(":", "-", go_id), ".pdf")
        ggsave(out_file, plot = p, width = width, height = height)
        message("Saved: ", out_file)
      }
    } else {
      message("GO term ", go_id, " not found in ", contrast_name)
    }
  }
}

# automatic enrichment plots:
# one for top positively enriched term and one for top negatively enriched term
save_top_bottom_gsea_plots <- function(gsea_list, prefix, padj_cutoff = 0.05, width = 8, height = 6) {
  for (nm in names(gsea_list)) {
    gsea_res <- gsea_list[[nm]]
    if (is.null(gsea_res)) next

    df <- tryCatch(as.data.frame(gsea_res), error = function(e) NULL)
    if (is.null(df) || !nrow(df)) next

    df <- df[!is.na(df$NES) & !is.na(df$p.adjust), ]
    if (!nrow(df)) next

    df_sig <- df[df$p.adjust < padj_cutoff, , drop = FALSE]
    if (nrow(df_sig) == 0) {
      message("No significant GSEA terms for ", nm)
      next
    }

    pos_df <- df_sig[df_sig$NES > 0, , drop = FALSE]
    neg_df <- df_sig[df_sig$NES < 0, , drop = FALSE]

    if (nrow(pos_df) > 0) {
      pos_df <- pos_df[order(-pos_df$NES, pos_df$p.adjust), , drop = FALSE]
      top_pos_id <- pos_df$ID[1]

      p_pos <- tryCatch(
        gseaplot2(gsea_res, geneSetID = top_pos_id,
                  title = paste0(nm, " top positive: ", top_pos_id)),
        error = function(e) NULL
      )

      if (!is.null(p_pos)) {
        out_pos <- paste0(prefix, "_GSEA_", nm, "_top_positive_", gsub(":", "-", top_pos_id), ".pdf")
        ggsave(out_pos, plot = p_pos, width = width, height = height)
        message("Saved: ", out_pos)
      }
    }

    if (nrow(neg_df) > 0) {
      neg_df <- neg_df[order(neg_df$NES, neg_df$p.adjust), , drop = FALSE]
      top_neg_id <- neg_df$ID[1]

      p_neg <- tryCatch(
        gseaplot2(gsea_res, geneSetID = top_neg_id,
                  title = paste0(nm, " top negative: ", top_neg_id)),
        error = function(e) NULL
      )

      if (!is.null(p_neg)) {
        out_neg <- paste0(prefix, "_GSEA_", nm, "_top_negative_", gsub(":", "-", top_neg_id), ".pdf")
        ggsave(out_neg, plot = p_neg, width = width, height = height)
        message("Saved: ", out_neg)
      }
    }
  }
}

save_top_bottom_gsea_plots(gsea_list, prefix = fname_prefix_GO, padj_cutoff = 0.05)

# optional custom GO IDs from previous analysis
# go_ids <- c("GO:0042551","GO:0007492","GO:0007498")
# for (nm in names(gsea_list)) {
#   save_gsea_plots(gsea_list[[nm]], nm, go_ids, fname_prefix_GO)
# }

# optional: significant-only dotplots
save_sig_gsea_dotplots <- function(gsea_sig_list, prefix, show_k = 10, width = 8, height = 8) {
  for (nm in names(gsea_sig_list)) {
    g <- gsea_sig_list[[nm]]
    if (is.null(g)) next

    df <- tryCatch(as.data.frame(g), error = function(e) NULL)
    if (is.null(df) || !nrow(df)) next

    k <- min(show_k, nrow(df))

    p <- tryCatch(
      dotplot(g, showCategory = k, split = ".sign", orderBy = "NES") +
        facet_grid(. ~ .sign),
      error = function(e) NULL
    )

    if (!is.null(p)) {
      ggsave(
        filename = paste0(prefix, "_", nm, "_GSEA_sig.pdf"),
        plot = p,
        width = width,
        height = height
      )
      message("Saved: ", paste0(prefix, "_", nm, "_GSEA_sig.pdf"))
    }
  }
}

save_sig_gsea_dotplots(gsea_sig_list, prefix = fname_prefix_GO, show_k = 10)

Rdata_filename <- paste0(fname_prefix_GO, "_GSEA_analysis_clusters.RData")
save(gene_sets_GSEA, gsea_list, gsea_sig_list, GSEA_tables, de_stats,
     file = Rdata_filename)


# Vector of target GO IDs ###ID studied before
#go_ids <- c("GO:0042551","GO:0007492","GO:0007498") # 

# Function: loop through GO IDs for a given contrast
#save_gsea_plots <- function(gsea_res, contrast_name, go_ids, prefix, width=8, height=6) {
 # for (go_id in go_ids) {
  #  if (go_id %in% gsea_res@result$ID) {
   #   p <- gseaplot2(gsea_res, geneSetID = go_id, title = go_id)
    #  out_file <- paste0(prefix, "_", contrast_name, "_", gsub(":", "-", go_id), ".pdf")
     # ggsave(out_file, plot = p, width = width, height = height)
    #  message("Saved: ", out_file)
    #} else {
     # message("GO term ", go_id, " not found in ", contrast_name)
    #}
#  }
#}


# Example usage:
#save_gsea_plots(gsea_list[["4vsall"]], "4vsall", go_ids, fname_prefix_GO)
#save_gsea_plots(gsea_list[["3vsall"]], "3vsall", go_ids, fname_prefix_GO)



# Vector of target GO IDs for plotting
#go_ids <- c("GO:0042551","GO:0007492","GO:0007498")#neuron migration

# Loop over and plot#
#for (go_id in go_ids) {
 # if (go_id %in% gsea_list[["3vsall"]]@result$ID) {
  #  print(gseaplot2(gsea_list[["3vsall"]], geneSetID = go_id, title = go_id))
  #} else {
   # message(paste("GO term", go_id, "not found in results."))
  #}
#}


#dotplot(gsea, showCategory = 10, split = ".sign",orderBy = "enrichmentScore") + facet_grid(.~.sign)
#ggsave(filename = paste0(fname_prefix_ATAC_GO, "_", "_GSEA.pdf"),
 #    width = 8, height = 8)

### Plotting GSEA results
#dotplot(gsea_list[["3vsall"]], showCategory = 10, split = ".sign",orderBy = "enrichmentScore") + facet_grid(.~.sign)
#ggsave(filename = paste0(fname_prefix_GO, "_", "_GSEA.pdf"),
 #    width = 8, height = 8)

#dotplot(gsea_sig_list[["1vs2"]], showCategory = 10, split = ".sign", orderBy = "NES") +
 # facet_grid(. ~ .sign)
#ggsave(filename = paste0(fname_prefix_GO, "_1vs2_GSEA_sig.pdf"),
 #      width = 8, height = 8)


```
####cluster annotation

```{r}
cluster_tab$cluster_name = c(
  "1" = "Mesenchymal intermediates",
  "2" = "Non-myogenic (neural-like-2)",
  "3" = "Mature myotubes",
  "4" = "Early myotubes",
  "5" = "BAF smooth-like myotubes",
  "6" = "Myoblasts",
  "7" = "Non-myogenic (neural-like)",
  "8" = "Intermediate Myotubes",
  "9" = "Slow fiber / innervation-ready"
)#need to change when defined

cluster_tab$cluster_name_short=c("Mes","Neur-2","Late Myot","Erly Myot","BAF Myot","Myob","Neur","Int Myot","9")#need to change when defined

cluster_markers_plot <-list(cl1 = c("VIM", "NPM1", "GLUL", "RPS8","RPL10"),#pre myogenic progenitors Ribosome-hi, VIM+, GLUL+; no sarcomeric genes by DEG
  cl2 = c("MKI67", "TOP2A","PCNA","TTN","MYH3","MYOD1"), #proliferating progenitors, early sarcomeric commitment compared to cl1
  cl3 = c("MYH2", "MYH7", "TNNI1", "ACTN2", "MYOT","NES","COL2A1"), #highest sarcomere score
  cl4 = c("DES", "CA3", "TAGLN", "DLK1","MYL7","MYL9","TAGLN"),#CIRI
  cl5 = c("MYH11", "ACTG2", "ACTG1","MEIS2", "KCNMA1","HCN1","HCN2","HCN3","HCN4","CACNA1H","CACNA1G"), #BAF specific
  cl6 = c("MYOD1", "CKM", "CDKN1C", "CRABP1"),#OPTIOX
  cl7 = c("PDGFRA","DCN","LUM","COL1A1","THY1","PAX7","VIM","GABRG3","TMEM108","CADM1","PLCB4","TNIK"),#synaptic genes
  cl8 = c("MYOG", "MYH3","MYMK","MYMX","NLGN1","ENDOU", "FAM246C", "PDC", "PROKR2", "IFIT1B", "CIB3", "LRRTM3"),#Weakly defined transitional progenitor #MYMX is nearly absent in cl8 (This means cl8 is not a fusing myocyte)
  cl9 = c("DMD", "SOX6", "RYR3", "DLG2","CSMD1","KCND2","CNTN5","AKAP6"))


####make plots

for (sig_name in names(cluster_markers_plot)) {
  
  genes_use <- intersect(cluster_markers_plot[[sig_name]], rowData(cds)$gene_short_name)
  
  message("Plotting clust set: ", sig_name, " | genes found: ", length(genes_use))
  
  if (length(genes_use) == 0) next
  
  p <- plot_cells(
    cds,
    genes = genes_use,
    label_cell_groups = FALSE,
    show_trajectory_graph = FALSE
  ) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = sig_name
    ) +
    scale_color_viridis(option = "E", discrete = FALSE)
  
  print(p)
  
  ggsave(
    filename = paste0(fname_prefix_dotplot, "_UMAP_", sig_name, ".pdf"),
    plot = p,
    width = 8,
    height = 6
  )
}

# ── violin plots for all module scores ─────────────────────────────────────

# Colour palette (Okabe-Ito, 4 samples)
samp_pal <- c(
  "JMJ"     = "#009E73",
  "BAF_cl14"= "#0072B2",
  "CIRI"    = "#E69F00",
  "OPTIOX"  = "#D55E00"
)

# Nice labels for x axis
cluster_labels <- c(
  "1" = "Mesenchymal intermediates",
  "2" = "Non-myogenic (neural-like-2)",
  "3" = "Mature myotubes",
  "4" = "Early myotubes",
  "5" = "BAF smooth-like myotubes",
  "6" = "Myoblasts",
  "7" = "Non-myogenic (neural-like)",
  "8" = "Intermediate Myotubes",
  "9" = "Slow fiber / innervation-ready")


score_labels <- c(
  sarcomere_core                 = "sarcomere core",
  developmental_embryonic_perinatal = "developmental embryonic perinatal",
  myotube_differentiation_fusion = "myotube differentiation & fusion",
  myoblast_proliferation_cycle   = "myoblast proliferation cycle",
  fiber_type_I_slow              = "fiber type I slow",
  fiber_type_IIa_fast            = "fiber type IIa fast",
  pluripotency_core              = "pluripotency core",
  glycolytic_fast                = "glycolytic fast",
  ribosome_tone                  = "ribosome tone",
  calcium_excitation_contraction = "calcium excitation-contraction",
  oxidative_mitochondrial        = "oxidative mitochondrial",
  satellite_cell_core            = "satellite cell core",
  myogenic_regulators_axis       = "myogenic regulators axis",
  neuromuscular_junction         = "neuromuscular junction",
  cellcycle_s                    = "cell cycle S phase",
  cellcycle_g2m                  = "cell cycle G2/M"
)

# ── data ─────────────────────────────────────────────────────────────────────
meta <- as.data.frame(seurat_AB017@meta.data)
meta$monocle_clusters <- factor(meta$monocle_clusters, levels = as.character(1:9))

# merge cluster colours into meta for violin fill
meta <- meta %>%
  left_join(cluster_tab, by = c("monocle_clusters" = "monocle_cluster"))

# per-sample medians (for the coloured dots)
sample_medians <- meta %>%
  group_by(monocle_clusters, sample_id) %>%
  summarise(across(all_of(names(score_labels)), median, na.rm = TRUE),
            .groups = "drop")

# named vector for scale_fill_manual: cluster -> colour
violin_fill_pal  <- setNames(cluster_tab$cluster_color, cluster_tab$monocle_cluster)

# ── plot loop ─────────────────────────────────────────────────────────────────
for (module in names(score_labels)) {

  if (!module %in% colnames(meta)) next
  message("Plotting: ", module)

  med_long <- sample_medians %>%
    dplyr::select(monocle_clusters, sample_id, value = all_of(module))

  p <- ggplot(meta, aes(x = monocle_clusters, y = .data[[module]])) +

    # violin: fill = cluster colour, white interior, coloured outline
    geom_violin(
      aes(fill = monocle_clusters, color = monocle_clusters),
      alpha     = 0.25,       # light fill so dots remain visible
      trim      = FALSE,
      scale     = "width",
      linewidth = 0.5
    ) +

    # coloured dots = per-sample median
    geom_point(
      data  = med_long,
      aes(x = monocle_clusters, y = value, color = sample_id),
      size     = 3,
      shape    = 16,
      position = position_jitter(width = 0.05, seed = 42)
    ) +

    # violin fill = cluster colours (light, alpha already set above)
    scale_fill_manual(
      values = violin_fill_pal,
      guide  = "none"          # don't show fill in legend
    ) +

    # colour scale shared by violin outline AND sample dots
    # we need two separate scales: one for violin outline, one for dots
    # easiest: use fill for violin, colour for dots only
    scale_color_manual(
      values = samp_pal,
      name   = "Condition",
      breaks = names(samp_pal)   # only show sample colours in legend
    ) +

    scale_x_discrete(labels = cluster_labels) +

    labs(
      x     = "",
      y     = "AddModuleScore",
      title = score_labels[[module]]
    ) +

    theme_classic(base_size = 11) +
    theme(
      plot.background    = element_rect(fill = "white", color = NA),
      panel.background   = element_rect(fill = "white", color = NA),
      panel.grid.major.y = element_line(color = "#eeeeee", linewidth = 0.4),
      panel.grid.major.x = element_blank(),
      axis.line          = element_line(color = "#111111"),
      axis.ticks         = element_line(color = "#111111"),
      axis.text          = element_text(color = "#333333"),
      axis.text.x        = element_text(angle = 40, hjust = 1, size = 9),
      axis.title.y       = element_text(color = "#333333"),
      plot.title         = element_text(color = "#111111", face = "bold", hjust = 0.5),
      legend.background  = element_rect(fill = "white", color = NA),
      legend.text        = element_text(color = "#333333"),
      legend.title       = element_text(color = "#333333"),
      legend.key         = element_rect(fill = "white", color = NA)
    )

  ggsave(
    filename = paste0(fname_prefix_violin, "_violin_", module, ".pdf"),
    plot     = p,
    width    = 6,
    height   = 6
  )
}

```
  
####violin plots for genes selected for validation

```{r}
# Define your colour mapping (unchanged)

my_colors <- c(
  "1" = "#000000",
  "2" = "#56B4E9",
  "3" = "#009E73",
  "4" = "#004949",
  "5" = "#24ff24",
  "6" = "#CC79A7",
  "7" = "#D55E00",
  "8" = "#F0E442",
  "9" = "#0072B2"
)

gene_list_violin <- c("VIM","MYOD1","CRABP1","CDKN1C","TNN1","MYH3","MYOG","MYL9","PAX7","CA3","MYH11","ACTG1","MKI67")

plot_violin_gene <- function(gene) {
  if (!gene %in% rownames(normalized_counts(cds))) {
    message("Gene ", gene, " not found in normalized_counts(cds); skipping.")
    return(invisible(NULL))
  }

  plot_df <- as.data.frame(colData(cds))
  plot_df$cell_id <- rownames(plot_df)

  norm_expr <- normalized_counts(cds)[gene, ]
  plot_df$expr <- as.numeric(norm_expr[plot_df$cell_id])

  # make sure clusters are characters/factors matching color names
  plot_df$monocle_clusters <- as.character(plot_df$monocle_clusters)
  plot_df$monocle_clusters <- factor(
    plot_df$monocle_clusters,
    levels = names(my_colors)
  )

  p <- ggplot(
    plot_df,
    aes(x = monocle_clusters, y = expr, fill = monocle_clusters)
  ) +
    geom_violin(trim = FALSE) +
    stat_summary(
      fun = median,
      geom = "crossbar",
      width = 0.5,
      color = "black"
    ) +
    scale_fill_manual(values = my_colors, drop = FALSE) +
    labs(
      x = "Monocle cluster",
      y = "Normalized expression",
      title = paste0(gene, " gene expression")
    ) +
    theme_classic()

  print(p)

  ggsave(
    filename = paste0(fname_prefix_violin, "_", gene, ".pdf"),
    plot = p,
    width = 6,
    height = 5
  )
}

for (g in gene_list_violin) {
  plot_violin_gene(g)
}

####compare samples

samp_pal <- c(
  "CIRI"     = "#E69F00",
  "OPTIOX"   = "#D55E00",
  "JMJ"      = "#009E73",
  "BAF_cl14" = "#0072B2"
  )

gene_list_violin <- c("VIM","MYOD1","CRABP1","CDKN1C","TNN1","MYH3","MYOG","MYL9","PAX7","CA3","MYH11","ACTG1","MKI67")

plot_violin_sample <- function(gene) {
  if (!gene %in% rownames(normalized_counts(cds))) {
    message("Gene ", gene, " not found in normalized_counts(cds); skipping.")
    return(invisible(NULL))
  }

  plot_df <- as.data.frame(colData(cds))
  plot_df$cell_id <- rownames(plot_df)

  norm_expr <- normalized_counts(cds)[gene, ]
  plot_df$expr <- as.numeric(norm_expr[plot_df$cell_id])

  plot_df$sample_id <- as.character(plot_df$sample_id)
  plot_df$sample_id <- factor(plot_df$sample_id, levels = names(samp_pal))

  p <- ggplot(plot_df, aes(x = sample_id, y = expr, fill = sample_id)) +
    geom_violin(trim = FALSE) +
    stat_summary(
      fun = median,
      geom = "crossbar",
      width = 0.5,
      color = "black"
    ) +
    scale_fill_manual(values = samp_pal, drop = FALSE) +
    labs(
      x = "Sample",
      y = "Normalized expression",
      title = paste0(gene, " expression by sample")
    ) +
    theme_classic()

  print(p)

  ggsave(
    filename = paste0(fname_prefix_violin, "_sample_", gene, ".pdf"),
    plot = p,
    width = 6,
    height = 5
  )
}

for (g in gene_list_violin) {
  plot_violin_sample(g)
}


plot_cells(cds,
           genes = gene_list_violin,
            label_cell_groups = FALSE,
            show_trajectory_graph = FALSE)+ 
     labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     scale_color_viridis(option="E", discrete=F)
 
 ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_validation.pdf"),
     width = 8, height = 6)

```

#load dataset to compare

```{r}
load(paste0(getwd(),"/scratch/seurat_AB07.RData"))
```


######################################################
#####combine seurat UMAPS and lables between AB07 and AB017 ###after
######ideally we can use annotated published datasets
######################################################

```{r}
####combine UMAPS
seurat_AB017 <- FindVariableFeatures(seurat_AB017, selection.method = "vst", nfeatures = 8000)
seurat_AB017 =ScaleData(seurat_AB017)
seurat_AB017 =RunPCA(seurat_AB017 , npcs = 30)
DimPlot(seurat_AB017, group.by = "monocle_clusters", label = F)

seurat_AB07=FindVariableFeatures(seurat_AB07, selection.method = "vst", nfeatures = 8000)
seurat_AB07 =ScaleData(seurat_AB07)
seurat_AB07 =RunPCA(seurat_AB07 , npcs = 30)
DimPlot(seurat_AB07, group.by = "monocle_clusters", label = F)


# Find transfer anchors
anchors <- FindTransferAnchors(reference = seurat_AB07, 
                               query = seurat_AB017, 
                               normalization.method = "LogNormalize", 
                               #features = VariableFeatures(object = seurat_AB07),
    reference.assay = "originalexp", query.assay = "originalexp", reduction = "pcaproject",
                               #features="PCA",
                               dims = 1:30)

# Transfer labels from reference to query
predictions <- TransferData(anchorset = anchors, 
                            refdata = seurat_AB07$monocle_clusters, 
                            dims = 1:30)
predictions$libID=rownames(predictions)

predictions=predictions%>%dplyr::select(predicted.id,libID)%>%dplyr::rename(predicted_AB07=predicted.id)

cell_metadata=cell_metadata%>%right_join(predictions, by="libID")
cds@colData$predicted_AB07=predictions$predicted_AB07

AB07_clust=meta_AB07%>%
  group_by(monocle_clusters,cluster) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(monocle_clusters, desc(n))

AB07_clust=AB07_clust%>%rename(predicted_AB07=monocle_clusters,cell_type_AB07=cluster)%>%
  dplyr::select(-n)

cell_metadata=cell_metadata%>%left_join(AB07_clust, by="predicted_AB07")
cds@colData$cell_type_AB07=cell_metadata$cell_type_AB07
cds@colData$cell_type_short_AB07=cell_metadata$cell_type_short_AB07

meta=as.data.frame(cds@colData@listData)
stat_filename = paste0(fname_prefix_R, "_meta.csv")
write.csv(meta, stat_filename, quote = F, row.names = T, sep=",")

# Add the predicted cluster annotations to the query metadata
seurat_AB017 <- AddMetaData(object = seurat_AB017, metadata = cell_metadata)

DimPlot(seurat_AB017, group.by = "monocle_clusters", label = T)
DimPlot(seurat_AB017, group.by = "predicted_AB07", label = F)
DimPlot(seurat_AB017, group.by = "cell_type_AB07", label = T)

cluster_col_AB07 <- c(
  "Mesenchymal Progenitors"="#CC79A7",
  "Myoblasts"="#000000",
  "Myocytes"="#E69F00",
  "Posterior neurectoderm"="#56B4E9")

cl_col_AB07 <- c(
  "1"="#000000",
  "2"="#E69F00",
  "3"="#56B4E9",
  "4"="#CC79A7")

a=DimPlot(seurat_AB07, group.by = "monocle_clusters", label = F,cols=cl_col_AB07)
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", "AB07",".pdf"),
    plot = a,
    width = 6,
    height = 4)

p=ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle,color = predicted_AB07)) +
    geom_point(size = 0.8,alpha=1) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = "predicted_AB07"
    ) +
    scale_color_manual(name = "monocle",
                       values=cl_col_AB07) +
                      #values=cl_pal5) +
    facet_grid(~(factor(sample_id, levels = c("OPTIOX", "CIRI","JMJ","BAF_cl14"))))
  
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", "predicted_AB07", "_predictions_AB07.pdf"),
    plot = p,
    width = 8,
    height = 4)
  
p=ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle,color = cell_type_AB07)) +
    geom_point(size = 0.8,alpha=1) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = "cell_type_AB07"
    ) +
    scale_color_manual(name = "monocle",
                       values=cluster_col_AB07) +
                      #values=cl_pal5) +
    facet_grid(~(factor(sample_id, levels = c("OPTIOX", "CIRI","JMJ","BAF_cl14"))))
  
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", "cell_type_AB07", "_predictions_AB07.pdf"),
    plot = p,
    width = 8,
    height = 4)


Rdata_filename = paste0(fname_prefix_R, "_monocle_seurat_cds_lable_prediction_AB07.RData")
save(cell_metadata,counts,cds,meta,original_counts,seurat_AB017,
     file = Rdata_filename)
```


####calculate % of cells

```{r}
meta_to_use=meta

tot=table(meta_to_use$sample_id) #total of the cluster
df=table(meta_to_use$monocle_clusters,meta_to_use$sample_id)
df = as.data.frame(df)
colnames(df)=c("monocle_cluster","sample_id","freq")

#mdf2 = melt(df)
mdf2=df#11664 rows
total=as.data.frame(tot)
colnames(total)=c("sample_id","tot")
mdf2=mdf2%>%left_join(total, by="sample_id")
mdf2$tot = as.numeric(mdf2$tot)
mdf2$frxn = (mdf2$freq/mdf2$tot)*100
m=subset(mdf2,mdf2$freq!=0)

m=m%>%left_join(cluster_tab, by="monocle_cluster")

percent_cl=m


ggplot(percent_cl, aes(x = factor(sample_id, levels = names(samp_pal)),
    y = frxn,
    fill = factor(cluster_name, levels = cluster_tab$cluster_name)     # se vuoi anche qui un ordine preciso, basta aggiungere levels = c(...)
  ))+
  geom_bar(stat = "identity") +theme(legend.position='none')+
  #geom_col(position = "stack") +  # equivalente a geom_bar(stat = "identity")
  scale_fill_manual(values =c(cluster_tab$cluster_color))+
  #facet_wrap(
   # vars(factor(media, levels = c("E8", "E6_FGF")))   # ordine dei facet
  #) +
  labs(
    x = "",
    y = "% cells per medium and cell clone",
    fill = "monocle cluster"
  ) +
  theme_bw()

ggsave(filename = paste0(fname_prefix_mosaic, "_", "percent_cells_type.pdf"),
     width = 8, height = 5)


#ggplot(m, aes(x = factor(assigned_cell_type_short,levels = cell_type_short_levels), y = frxn, fill=factor(assigned_cell_type_short,levels = cell_type_short_levels))) + geom_bar(stat = "identity") +theme(legend.position='none')+
 # scale_fill_manual(values =c(col_cl))+
#  labs(ylab = "% gene by treatment",
#xlab = "cluster")+
 # facet_grid(~factor(medium,levels=c("E8","B8","M8")))
  

#percent_cl=percent_cl%>%right_join(clust_tab,by=c("monocle_cluster"))

stat_filename = paste0(fname_prefix_stat, "_percent_cells_AB017.csv")
write.csv(percent_cl, stat_filename, quote = F, row.names = T, sep=",")


```
###fisher exact test for AB017 clusters

```{r}
library(dplyr)
library(purrr)
library(tibble)

pairwise_cluster_fisher <- function(meta,
                                    cluster_col = "monocle_clusters",
                                    sample_col = "sample_id",
                                    pairs = list(
                                      c("BAF_cl14", "CIRI"),
                                      c("BAF_cl14", "OPTIOX"),
                                      c("JMJ", "CIRI"),
                                      c("JMJ", "OPTIOX"),
                                      c("BAF_cl14", "JMJ")
                                    )) {
  
  clusters <- sort(unique(as.character(meta[[cluster_col]])))
  
  res <- map_dfr(pairs, function(pair) {
    s1 <- pair[1]
    s2 <- pair[2]
    
    meta_sub <- meta %>%
      filter(.data[[sample_col]] %in% c(s1, s2)) %>%
      mutate(
        cluster_chr = as.character(.data[[cluster_col]]),
        sample_chr = as.character(.data[[sample_col]])
      )
    
    n1_total <- sum(meta_sub$sample_chr == s1)
    n2_total <- sum(meta_sub$sample_chr == s2)
    
    map_dfr(clusters, function(cl) {
      n1_in <- sum(meta_sub$sample_chr == s1 & meta_sub$cluster_chr == cl)
      n2_in <- sum(meta_sub$sample_chr == s2 & meta_sub$cluster_chr == cl)
      
      n1_out <- n1_total - n1_in
      n2_out <- n2_total - n2_in
      
      mat <- matrix(
        c(n1_in, n1_out,
          n2_in, n2_out),
        nrow = 2,
        byrow = FALSE,
        dimnames = list(
          c("in_cluster", "out_cluster"),
          c(s1, s2)
        )
      )
      
      ft <- fisher.test(mat)
      
      tibble(
        comparison = paste0(s1, "_vs_", s2),
        sample_1 = s1,
        sample_2 = s2,
        monocle_cluster = cl,
        count_1 = n1_in,
        total_1 = n1_total,
        frac_1 = 100 * n1_in / n1_total,
        count_2 = n2_in,
        total_2 = n2_total,
        frac_2 = 100 * n2_in / n2_total,
        diff_frac = (100 * n1_in / n1_total) - (100 * n2_in / n2_total),
        odds_ratio = unname(ft$estimate),
        conf_low = ft$conf.int[1],
        conf_high = ft$conf.int[2],
        p_value = ft$p.value
      )
    }) %>%
      mutate(p_adj_BH = p.adjust(p_value, method = "BH"))
  })
  
  res
}

stat_pairs <- pairwise_cluster_fisher(meta)

stat_pairs=stat_pairs%>%mutate(
  comparison = factor(stat_pairs$comparison, levels = c(
  "BAF_cl14_vs_CIRI",
  "BAF_cl14_vs_OPTIOX",
  "JMJ_vs_CIRI",
  "JMJ_vs_OPTIOX",
  "BAF_cl14_vs_JMJ"
)),
significant = p_adj_BH <= 0.01,
low_count_flag = (count_1 + count_2) < 20,
prop_1 = count_1 / total_1,
prop_2 = count_2 / total_2,
log2_prop_ratio = log2((prop_1 + 1e-6) / (prop_2 + 1e-6)),
neglog10_padj = -log10(p_adj_BH + 1e-300)
)%>%
  left_join(cluster_tab, by = "monocle_cluster") %>%
  mutate(
    prop_1 = count_1 / total_1,
    prop_2 = count_2 / total_2,
    log2_prop_ratio = log2((prop_1 + 1e-6) / (prop_2 + 1e-6)),
    neglog10_padj = -log10(p_adj_BH + 1e-300),
    plot_color = ifelse(significant & !low_count_flag, cluster_color, "grey75"),
    label_me = significant & !low_count_flag
  )

write.csv(
  stat_pairs,
  file = paste0(fname_prefix_stat, "_pairwise_cluster_fisher.csv"),
  row.names = FALSE
)
```

##volcano plot to show statistics of fisher exact test

```{r}
library(ggplot2)
library(ggrepel)

lab_dat <- stat_pairs %>%
  filter(label_me) %>%
  group_by(comparison) %>%
  arrange(p_adj_BH, desc(abs(diff_frac))) %>%
  slice_head(n = 3) %>%
  ungroup()

p_volcano <- ggplot(
  stat_pairs,
  aes(x = diff_frac, y = neglog10_padj)
) +
  geom_hline(yintercept = -log10(0.01), linetype = 2, color = "grey50") +
  geom_vline(xintercept = 0, linetype = 2, color = "grey50") +
  geom_point(
    aes(fill = plot_color),
    shape = 21,
    color = "black",
    size = 3,
    stroke = 0.2,
    alpha = 0.95
  ) +
  scale_fill_identity() +
  ggrepel::geom_text_repel(
    data = lab_dat,
    aes(label = cluster_name),
    size = 3,
    max.overlaps = Inf
  ) +
  facet_wrap(~comparison, scales = "free_x") +
  labs(
    x = "Difference in cluster fraction (%)",
    y = expression(-log[10](adjusted~pvalue)),
    title = "Exploratory pairwise cluster-composition differences"
  ) +
  theme_bw()

ggsave(
  filename = paste0(fname_prefix_stat, "_pairwise_cluster_volcano_clusterColors.pdf"),
  plot = p_volcano,
  width = 10,
  height = 7
)
```




