---
title: "n1 AB007: QC and monocle analysis"
author: "Elisa Balmas PhD"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment

```{r}
library(monocle3)
library(tidyr)
library(stringr)
library(grid)
library(viridis)
library(viridisLite)
library(ggthemes)
library(dplyr)
library(tidyverse)
library(ggsignif)
library(umap)
library(heatmap3)
library(plyr)
library(edgeR)
library(compareGroups)
library(dbscan)
library(MAST)
library(geosphere)
library(RColorBrewer)
library(Seurat)
library(htmlwidgets)
library(Matrix)
library(tidyseurat)
library(SeuratWrappers)
library(scMAGeCK)
library(readxl)
#library(rCASC)
#library(deepToolsUtils)
#library(R.utils)


library(ggplot2); library(reshape2); theme_set(theme_bw(26) + theme(panel.grid.major = element_blank(), 
                                                                    panel.grid.minor = element_blank()) +
                                                 theme(legend.key = element_blank()))
update_geom_defaults("point", aes(size = 4))

#setwd("/home/ebalmas/scpHUB/projects/H001AS7_iKDseq_10X_4tp_monocle/analysis_EB")

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))


# Set up filename prefixes and output folders with the data
dir.create("Output")
#dir.create(file.path("Output","Fig5_input_folder"))
dir.create(file.path("Output","QC"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito"))

dir.create(file.path("Output","monocle"))#create forders MONOCLE
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots","single"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "stat"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "UMAP"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "mosaic"))

dir.create(file.path("Output","Seurat"))#create forders Seurat
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "dotplots"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "PCA"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "UMAP"))


fname_Fig5_input_folder <- file.path("Output","Fig5_input_folder")
#set up folder path QC
fname_prefix_R_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_csv_QC<-file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_ribomito_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito", 
                               format(Sys.Date(), "%y%m%d"))

#set up forder path MONOCLE
fname_prefix_R <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot_single <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", "single", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_csv<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_stat<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "stat", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_UMAP<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "UMAP", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_mosaic<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "mosaic", 
                               format(Sys.Date(), "%y%m%d"))


#set up forder path Seurat
fname_prefix_R_Seurat <- file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot_Seurat <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_PCA_Seurat<-file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "PCA", 
                               format(Sys.Date(), "%y%m%d"))

fname_prefix_UMAP_Seurat<-file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "UMAP", 
                               format(Sys.Date(), "%y%m%d"))

okabe_tab=read_excel("Fig5_input_folder/color_scale.xlsx")
okabe_pal=okabe_tab$hex

okabe_pal=c("#E69F00","#56B4E9","#009E73","#f0E442","#0072B2","#D55E00","#CC79A7","#000000")


pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
         "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
         "#920000","#924900","#db6d00","#24ff24","#ffff6d")
```

######create Seurat Object

```{r}
matrix_dir ="Fig5_input_folder/filtered_feature_bc_matrix"

barcode.path <- paste0(matrix_dir, "/barcodes.tsv.gz")
features.path <- paste0(matrix_dir, "/features.tsv.gz")
matrix.path <- paste0(matrix_dir, "/matrix.mtx.gz")

#h5=paste0(matrix_dir, "/filtered_feature_bc_matrix.h5")

data=Read10X(
  data.dir =matrix_dir,
  gene.column = 2,
  cell.column = 1,
  unique.features = TRUE,
  strip.suffix = FALSE
)

#d=Read10X_h5(h5, use.names = TRUE, unique.features = TRUE)

#create Seurat object
data_seurat <- CreateSeuratObject(counts = data, project = "H001AS9_AB03", min.cells = 3, names.delim = "-", names.field = 2)

meta_original=data_seurat[[]]

#sample_info=read.csv("Fig5_input_folder/aggr_all_config.csv")
sample_info=read_delim("Fig5_input_folder/aggr_all_config.csv", 
    delim = ";", escape_double = FALSE, trim_ws = TRUE)
#clones_info=read.csv("/home/ebalmas/scpHUB/projects/AB03/analisis_EB/clones.csv")
sample.names=sample_info$sample_id
colnames(sample_info)=c("sample","replicate","orig.ident")

sample_info$orig.ident <- as.factor(sample_info$orig.ident)
meta_original=meta_original%>%left_join(sample_info, by="orig.ident")
pre_QCcounts=meta_original %>%dplyr::group_by(replicate,sample)%>%dplyr::summarise(count = n())
write.csv(pre_QCcounts, file= paste0(fname_prefix_csv_QC, "_", "pre_QCcounts.csv"), row.names=T)

```

###make a seurat object
#create metadata and filter Seurar object (count and meta) with the barcode data

```{r}
#build sample name file to add metadata in Seurat
#use orig.ident to make sure you are matching the right sample
sample_info$orig.ident <- as.factor(sample_info$orig.ident)
OBJ_meta <- data_seurat@meta.data %>% 
  dplyr::select(orig.ident)
r=rownames((OBJ_meta))
OBJ_meta$libID=r
OBJ_meta=OBJ_meta%>%
  dplyr::left_join(sample_info, by="orig.ident")%>%
  dplyr:: distinct(libID, .keep_all=T)
row.names(OBJ_meta)<-r
OBJ_meta$test=OBJ_meta$libID==rownames((OBJ_meta))
summary(as.factor(OBJ_meta$test))
OBJ_meta=OBJ_meta%>%
  dplyr::select(-test)

OBJ_meta$lib_ID=sapply(strsplit(as.character(OBJ_meta$libID), "-"), "[[", 1)


rownames(OBJ_meta)=OBJ_meta$libID

########################
###filter only for cells with a barcode in Seurat Object
########################
cell_metadata=OBJ_meta

####add metadata after filtering
data_seurat<- AddMetaData(data_seurat, cell_metadata, col.name = NULL)

####################
####CREATE EXP MATRIX
####################
#expression_data=data[["Gene Expression"]]
expression_data=LayerData(data_seurat, assay = "RNA", layer = "counts")
expression_matrix=as.matrix(expression_data)

write.csv(cell_metadata, file= paste0(fname_prefix_csv_QC, "_", "full_meta.csv"), row.names=T)
write.csv(expression_matrix, file= paste0(fname_prefix_csv_QC, "_", "full_counts.csv"), row.names=T)


Rdata_filename = paste0(fname_prefix_R_QC, "_preQC.RData")
save(cell_metadata, data_seurat,expression_matrix,
     file = Rdata_filename)

```
###Get QC quantifications (10X data pipeline)

```{r}
#########calculate % of mitocondiral reads and ribosomal reads
data_seurat[["percent.mt"]] <- PercentageFeatureSet(data_seurat, pattern = "^MT-")
data_seurat[["percent.ribo"]] <- PercentageFeatureSet(data_seurat, pattern = "RPS")

#make a QC dataframe to plot the QC metrics and then use it as meta table for monocle
QC=data_seurat[[c("nFeature_RNA","percent.mt","nCount_RNA","percent.ribo")]]
QC <- QC %>%
        dplyr::rename(nUMI = nCount_RNA,
                      nGene = nFeature_RNA)
QC$libID=rownames(QC)
QC=QC%>%right_join(cell_metadata, by="libID")
QC$log10GenesPerUMI <- log10(QC$nGene) / log10(QC$nUMI)

VlnPlot(data_seurat, features = c("nFeature_RNA", "nCount_RNA"),group.by = "sample",
     pt.size=0.1, ncol = 2)
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "feature_QC.pdf"),
     width = 10, height = 8)
###n UMI
QC %>% 
  	ggplot(aes(color=sample, x=nUMI, fill= sample)) + 
  	geom_density(alpha = 0.2) + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = 200)
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "UMI_QC.pdf"),
     width = 10, height = 8)
###n genes detected
QC %>% 
  	ggplot(aes(color=sample, x=nGene, fill= sample)) + 
  	geom_density(alpha = 0.2) + 
  	theme_classic() +
  	scale_x_log10() + 
  	geom_vline(xintercept = 300)
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_det_QC.pdf"),
     width = 10, height = 8)
QC %>% 
  	ggplot(aes(x=sample, y=log10(nGene), fill=sample)) + 
  	geom_boxplot() + 
  	theme_classic() +
  	theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  	theme(plot.title = element_text(hjust=0.5, face="bold")) +
  	ggtitle("NCells vs NGenes")
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_QC.pdf"),
     width = 10, height = 8)
QC %>% 
  	ggplot(aes(x=nUMI, y=nGene, color=percent.mt)) + 
  	geom_point() + 
	scale_colour_gradient(low = "gray90", high = "black") +
  	stat_smooth(method=lm) +
  	scale_x_log10() + 
  	scale_y_log10() + 
  	theme_classic() +
  	geom_vline(xintercept = 200) +
  	geom_hline(yintercept = 300) 
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "gene_UMI_QC.pdf"),
     width = 10, height = 8)
QC %>%
  	ggplot(aes(x=log10GenesPerUMI, color = sample, fill=sample)) +
  	geom_density(alpha = 0.2) +
  	theme_classic() +
  	geom_vline(xintercept = 1.07)
rownames(QC)=QC$info
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "log10GenesPerUMI_QC.pdf"),
     width = 10, height = 8)

plot1 <- FeatureScatter(data_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt",group.by = "sample")+
  	geom_vline(xintercept = 200) +
  	geom_hline(yintercept = 15) 
plot2 <- FeatureScatter(data_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by = "sample")+geom_vline(xintercept = 200) +
  	geom_hline(yintercept = 300) 
plot3 <- FeatureScatter(data_seurat, feature1 = "percent.ribo", feature2 = "percent.mt",group.by = "sample")+
  	geom_hline(yintercept = 15)+
  geom_vline(xintercept = 18)
plot1 + plot2+plot3
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "all_QC.pdf"),
     width = 20, height = 8)

######################################
#FILTER CELLS by QC metrics #NB. 
filtered_seurat <- subset(data_seurat, subset = nFeature_RNA>300 & nCount_RNA>200 & percent.mt<15 &percent.ribo<18)
cells_to_keep=colnames(filtered_seurat)
rownames(QC)=QC$libID
QC=QC[rownames(QC) %in% cells_to_keep,]
######################################
#FILTER COUNTS by QC metrics
#get counts QC info and calculate the gene detection rate and drop the non expressed ones
counts <- GetAssayData(object = filtered_seurat, slot = "counts")
nonzero <- counts > 0 # Output a logical vector for every gene on whether the more than zero counts per cell
# Sums all TRUE values and returns TRUE if more than 10 TRUE values per gene
keep_genes <- Matrix::rowSums(nonzero) >= 10
filtered_seurat <- filtered_seurat[rownames(filtered_seurat) %in% keep_genes,]
genes_to_keep=rownames(filtered_seurat)

###########
#post QC plots
plot1 <- FeatureScatter(filtered_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt",group.by = "sample")+
  	geom_vline(xintercept = 200) +
  	geom_hline(yintercept = 15) 
plot2 <- FeatureScatter(filtered_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",group.by = "sample")+geom_vline(xintercept = 200) +
  	geom_hline(yintercept = 300) 
plot3 <- FeatureScatter(filtered_seurat, feature1 = "percent.ribo", feature2 = "percent.mt",group.by = "sample")+
  	geom_hline(yintercept = 15)+
  geom_vline(xintercept = 15)
plot1 + plot2+plot3
ggsave(filename = paste0(fname_prefix_ribomito_QC, "_", "all_post_QC.pdf"),
     width = 20, height = 8)

#######################################
####################
####CREATE EXP MATRIX and cell metadata post QC
####################
cell_metadata=QC
expression_matrix <- filtered_seurat@assays[["RNA"]]@layers[["counts"]]
counts <- GetAssayData(object = filtered_seurat, slot = "counts")

# get gene annotations
gene_annotation=data.frame(rownames(counts))
colnames(gene_annotation) <- "gene_short_name"
rownames(gene_annotation)=gene_annotation$gene_short_name

write.csv(expression_data, file= paste0(fname_prefix_csv_QC, "_", "full_counts_post_QC.csv"), row.names=T)
write.csv(cell_metadata, file= paste0(fname_prefix_csv_QC, "_", "full_meta_post_QC.csv"), row.names=T)

#######################################
#make a Seurat Object normalized and scaled
#######################################
#Lognorm
filtered_seurat <- NormalizeData(filtered_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
# set seed and put two plots in one figure
set.seed(123)
par(mfrow=c(1,2))
# original expression distribution
raw_geneExp = as.vector(filtered_seurat[['RNA']]@layers[["counts"]]) %>% sample(10000)
raw_geneExp = raw_geneExp[raw_geneExp != 0]
hist(raw_geneExp)
# expression distribution after normalization
logNorm_geneExp = as.vector(filtered_seurat[['RNA']]@layers[["data"]]) %>% sample(10000)
logNorm_geneExp = logNorm_geneExp[logNorm_geneExp != 0]
hist(logNorm_geneExp)

ggsave(filename = paste0(fname_prefix_dotplot_Seurat, "_", "hist_QC_normalization.pdf"),
     width = 12, height = 8)

####find variable features
filtered_seurat <- FindVariableFeatures(filtered_seurat, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(filtered_seurat), 10)
# plot variable features with and without labels
plot1 <- VariableFeaturePlot(filtered_seurat) + 
  theme(legend.position="top")
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE) + 
  theme(legend.position="none")
plot1 + plot2

ggsave(filename = paste0(fname_prefix_dotplot_Seurat, "_", "DT_QC_HVG.png"),
     width = 12, height = 8)

#scale before performing dim reduction
## scale data
all.genes=rownames(filtered_seurat)
filtered_seurat <- NormalizeData(filtered_seurat, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_seurat <- ScaleData(filtered_seurat, features = all.genes)

#PCA
filtered_seurat <- FindVariableFeatures(filtered_seurat, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
filtered_seurat<- RunPCA(filtered_seurat, features = VariableFeatures(object = filtered_seurat))

# Examine and visualize PCA results 
#define meaningful PCAs
ElbowPlot(filtered_seurat) #main variation is captured from PCA 1 to 15
ggsave(filename = paste0(fname_prefix_PCA_Seurat, "_", "PCA_QC_elbow.png"),
     width = 8, height = 5)

print(filtered_seurat[["pca"]], dims = 1:30, nfeatures = 5)

VizDimLoadings(filtered_seurat,dims = 1:6, nfeatures = 30, col = "blue", reduction = "pca", projected = FALSE, balanced = FALSE, ncol = NULL, combine = TRUE)
ggsave(filename = paste0(fname_prefix_PCA_Seurat, "_", "PCA_QC_genes_definingPCs.png"),
     width = 12, height = 15)
png(file= paste0(fname_prefix_PCA_Seurat, "_", "PCA_QC_heatmaps_PCs.png"),
width=800, height=1200)
DimHeatmap(filtered_seurat, dims = 1:30, cells = 500, balanced = TRUE)
dev.off()
#no clear differences between replicates, mainly between D4 and D7
DimPlot(filtered_seurat, dims = c(1,2), reduction = "pca", group.by = "sample")
ggsave(filename = paste0(fname_prefix_PCA_Seurat, "_", "PCA_QC_PCAplots_1_2.png"),
     width = 8, height = 8)

#clustering and UMAP

filtered_seurat <- FindNeighbors(filtered_seurat, dims = 1:30)#first 15 seem to be  meaningful
filtered_seurat <- FindClusters(filtered_seurat, resolution = 0.4)
summary(as.factor(Idents(filtered_seurat)))

filtered_seurat <- RunUMAP(filtered_seurat, dims = 1:30)
DimPlot(filtered_seurat, reduction = "umap", split.by = "sample")

ggsave(filename = paste0(fname_prefix_UMAP_Seurat, "_", "UMAP.png"),
     width = 5, height = 3)


Rdata_filename = paste0(fname_prefix_R_QC, "_postQC.RData")
save(cell_metadata, expression_matrix,filtered_seurat,
     file = Rdata_filename)


rm(data_exp3,QC,matrix_Transpose,keep_genes,raw_geneExp,logNorm_geneExp,filtered_seurat, data_seurat, nonzero,sparse_Matrix, plot1, plot2, plot3)

post_QCcounts=cell_metadata %>%dplyr::group_by(replicate,sample)%>%dplyr::summarise(count = n())
write.csv(pre_QCcounts, file= paste0(fname_prefix_csv_QC, "_", "post_QCcounts.csv"), row.names=T)

```


####preprocess monocle data

```{r}

# get cell information

###create cds file
cds <- new_cell_data_set(counts,
                                      cell_metadata = cell_metadata,
                                      gene_metadata = gene_annotation)
cds <- preprocess_cds(cds, num_dim = 30)
#cds <- align_cds(cds, num_dim = 100, alignment_group = "sample")
plot_pc_variance_explained(cds)

ggsave(filename = paste0(fname_prefix_dotplot, "_", "variance.png"),
     width = 12, height = 8)

#cds <- align_cds(cds, alignment_group = "experiment")
cds = reduce_dimension(cds,reduction_method="tSNE")
cds = reduce_dimension(cds,umap.min_dist = 0.5,reduction_method="UMAP")

#monocle.object = order_cells(monocle.object, reduction_method = "UMAP")
set.seed(12345)
cds_1 = cluster_cells(cds, resolution = 1e-2)
colData(cds_1)$monocle_clusters = as.character(monocle3::clusters(cds_1))
set.seed(123479870)
cds_2 = cluster_cells(cds, resolution = 1e-4)
colData(cds_2)$monocle_clusters = as.character(monocle3::clusters(cds_2))
set.seed(123457768)
cds_3 = cluster_cells(cds, resolution = 8e-4)
colData(cds_3)$monocle_clusters = as.character(monocle3::clusters(cds_3))
set.seed(123458980)
cds_4 = cluster_cells(cds, resolution = 2e-4)
colData(cds_4)$monocle_clusters = as.character(monocle3::clusters(cds_4))
set.seed(1234597698)
cds_5 = cluster_cells(cds, resolution = 5e-4)
colData(cds_5)$monocle_clusters = as.character(monocle3::clusters(cds_5))


plot_cells(cds_1, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_2, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_4, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_5, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)

Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered.RData")
save(cell_metadata, expression_matrix,cds, cds_1,
     cds_2, cds_3, cds_4, cds_5,
     file = Rdata_filename)
```

###chose the cluster iteration with better results and clean environment

```{r}
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)#map by seurat clusters generated
last_plot()+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP.pdf"),
     width = 18, height = 8)

#map by seurat clusters generated
cds_3 <- learn_graph(cds_3)#pochi ma ok
cds=cds_3#this is the iteration we decided to use

#cds_cardio= cds[,cds@colData@listData[["monocle_clusters"]]=="5"|cds@colData@listData[["monocle_clusters"]]=="4"|cds@colData@listData[["monocle_clusters"]]=="3"|cds@colData@listData[["monocle_clusters"]]=="2"]


Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered_FINAL.RData")
save(cell_metadata, expression_matrix,cds,
     file = Rdata_filename)

#clean environment
rm(combined_exp,data_exp5,expression_data,gene_annotation,expression_matrix,cds_1,cds_2,cds_3,cds_4,cds_5)
```

###select cds with protein coding genes
#add umap data on cds meta
#add pseudotime info

```{r}

library("biomaRt")
   mart <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
   all_coding_genes <- getBM(attributes = c( "hgnc_symbol"), filters = c("biotype"), values = list(biotype="protein_coding"), mart = mart)
protein_coding_genes=all_coding_genes$hgnc_symbol

Rdata_filename = paste0(fname_Fig5_input_folder, "_protein_coding_genes.RData")
save(protein_coding_genes, 
     file = Rdata_filename)

#m=fData%>%filter(gene_short_name %in% protein_coding_genes)
cds <- cds[rowData(cds)$gene_short_name %in% protein_coding_genes,]#here I selected only protein coding genes

pbuild <- plot_cells(cds, color_cells_by = "monocle_clusters", label_cell_groups=T, show_trajectory_graph = F, label_leaves=F, label_branch_points=F, graph_label_size=3, group_label_size = 4, cell_size = 1) 
UmapCDS <- data.frame(pbuild$data$sample_name,UMAP1_monocle = pbuild$data$data_dim_1, UMAP2_monocle = pbuild$data$data_dim_2, row.names = 1) 
colData(cds)$UMAP_1_monocle=pbuild$data$data_dim_1 #add umap data for further analysis outside monocle
colData(cds)$UMAP_2_monocle=pbuild$data$data_dim_2 #add umap data for further analysis outside monocle

#counts=as.data.frame(exprs(cds))
counts=as.data.frame(normalized_counts(cds))
meta=as.data.frame(cds@colData@listData)

#cds<- order_cells(cds,reduction_method = "UMAP")#selected around cluster 3

#plot_cells(cds,
 #          color_cells_by = "pseudotime",
    #       label_cell_groups=FALSE,
   #        label_leaves=FALSE,
  #         label_branch_points=FALSE,
 #          graph_label_size=1.5)
#last_plot()+facet_grid(~sample)
#ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_pseudotime.pdf"),
 #    width = 18, height = 8)

Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered_FINAL_PROTEINCODING.RData")
save(counts,meta,cds,
     file = Rdata_filename)


```
###make gene lists

```{r}

#load gene lists
file=paste0("Fig5_input_folder/List_MuscleGenes.xlsx")
library(readxl)
m_markers_full_gene_list <- read_excel(file, 
    sheet = "A")

gene_lists=list()

for (c in unique(m_markers_full_gene_list$celltype)) {
 
tmp=subset(m_markers_full_gene_list, m_markers_full_gene_list$celltype==c)
gene_list=tmp$gene

gene_lists[[c]]=gene_list

}

# Create a data frame for the GMT format
gmt_data <- data.frame(
  Name = names(gene_lists),  # Use cell type names as gene set names
  Description = rep("https://doi.org/10.1016/j.cell.2022.11.028", length(gene_lists)),
  Gene_Symbols = lapply(gene_lists, paste, collapse = "\t")
)

# Write the data frame to a GMT file
write.table(gmt_data, file= paste0(fname_prefix_csv, "_", "gene_list_muscle.txt"), sep = "\t", quote = FALSE, row.names = FALSE, col.names = FALSE)
Rdata_filename = paste0(fname_prefix_R, "_gene_lists.RData")
save(gene_lists,
     file = Rdata_filename)

```

#make graphs to visualize genes
```{r}

for (gene_set_name in names(gene_lists)) {
  
  genes <- gene_lists[[gene_set_name]] # Extract the genes from list
  
  p <- plot_cells(
    cds,
    genes = genes,
    label_cell_groups = FALSE,
    show_trajectory_graph = FALSE
  ) + 
    labs(
      x = "UMAP 1", 
      y = "UMAP 2", 
      title = paste("UMAP Plot -", gene_set_name)
    ) +
    scale_color_viridis(option = "E", discrete = FALSE)
  
  ggsave(
    filename = paste0(fname_prefix_dotplot, "_", gene_set_name, ".pdf"),
    plot = p,
    width = 5, 
    height = 5
  )
}

#single plot
# plot_cells(cds,
 #           genes = c("SOX2","TBXT"),
  #          label_cell_groups = FALSE,
   #         show_trajectory_graph = FALSE)+ 
    # labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     #scale_color_viridis(option="E", discrete=F)
 
 #ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_Mesoderm_progenitors.pdf"),
   #  width = 10, height = 10)

ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = monocle_clusters),
               size=0.8)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clusters")+
  scale_color_manual(name = "monocle",
                        values=c(okabe_pal))+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_monocle_clusters.pdf"),
     width = 8, height = 6)

plot_feature=c("percent.mt", "percent.ribo","nUMI","nGene")

#ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
 #   geom_point(aes(color = nGene),
  #             size=0.8)+
  #labs(x = "UMAP 1", y = "UMAP 2", title = "nGene")+
  #scale_color_viridis(option="E", discrete=F)+facet_grid(~sample)
#ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_nGene.pdf"),
 #    width = 8, height = 6)

for (feature in plot_feature) {
  
  p <- ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes_string(color = feature), size = 0.8) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = feature
    ) +
    scale_color_viridis(option = "E", discrete = FALSE) +
    facet_grid(~sample)
  
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", feature, ".pdf"),
    plot = p,
    width = 8,
    height = 4
  )
}

```

##get top 25 genes per cluster

```{r}

marker_test_res = top_markers(cds, group_cells_by="monocle_clusters", reference_cells=NULL, cores=16)
marker_test_res <- top_markers(cds, group_cells_by="monocle_clusters", 
                               reference_cells=10000, cores=8)

top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(10, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t10=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t10)=c("ids", "cluster")

write.csv(t10, file= paste0(fname_prefix_csv, "_", "top_10genes_per_cluster.CSV"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="monocle_clusters",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_dotplot, "_", "top10_heatmap.pdf"),
     width = 8, height = 8)



top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(25, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t25=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t25)=c("ids", "cluster")

write.csv(t25, file= paste0(fname_prefix_csv, "_", "top_25genes_per_cluster.CSV"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="monocle_clusters",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_dotplot, "_", "top25_heatmap.pdf"),
     width = 8, height = 15)

```

####filter out cluster 2

```{r}
cds_clean=cds[,cds$monocle_clusters!="2"]

post_filter=as.data.frame(cds_clean@colData@listData) %>%dplyr::group_by(replicate,sample)%>%dplyr::summarise(count = n())
write.csv(post_filter, file= paste0(fname_prefix_csv_QC, "_", "post_filter.csv"), row.names=T)

cds_clean = reduce_dimension(cds_clean,reduction_method="UMAP",umap.metric = "cosine",umap.n_neighbors = 35L)

set.seed(1234589)
cds_1 = cluster_cells(cds_clean, resolution = 1e-2)
colData(cds_1)$monocle_clusters = as.character(monocle3::clusters(cds_1))
set.seed(1234798708)
cds_2 = cluster_cells(cds_clean, resolution = 1e-4)
colData(cds_2)$monocle_clusters = as.character(monocle3::clusters(cds_2))
set.seed(1234577688)
cds_3 = cluster_cells(cds_clean, resolution = 8e-4)
colData(cds_3)$monocle_clusters = as.character(monocle3::clusters(cds_3))
set.seed(1234589809)
cds_4 = cluster_cells(cds_clean, resolution = 2e-4)
colData(cds_4)$monocle_clusters = as.character(monocle3::clusters(cds_4))
set.seed(1234596989)
cds_5 = cluster_cells(cds_clean, resolution = 5e-4)
colData(cds_5)$monocle_clusters = as.character(monocle3::clusters(cds_5))


plot_cells(cds_1, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_2, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_4, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
plot_cells(cds_5, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)
last_plot()+facet_grid(~sample)
```
####chose the better permuation of UMAP
```{r}
plot_cells(cds_3, color_cells_by="cluster", group_cells_by="cluster",group_label_size = 7, cell_size=0.8)#map by seurat clusters generated
last_plot()+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clean.pdf"),
     width = 18, height = 8)

#map by seurat clusters generated
cds_3 <- learn_graph(cds_3)#pochi ma ok
cds=cds_3#this is the iteration we decided to use

#cds_cardio= cds[,cds@colData@listData[["monocle_clusters"]]=="5"|cds@colData@listData[["monocle_clusters"]]=="4"|cds@colData@listData[["monocle_clusters"]]=="3"|cds@colData@listData[["monocle_clusters"]]=="2"]

pbuild <- plot_cells(cds, color_cells_by = "monocle_clusters", label_cell_groups=T, show_trajectory_graph = F, label_leaves=F, label_branch_points=F, graph_label_size=3, group_label_size = 4, cell_size = 1) 
UmapCDS <- data.frame(pbuild$data$sample_name,UMAP1_monocle = pbuild$data$data_dim_1, UMAP2_monocle = pbuild$data$data_dim_2, row.names = 1) 
colData(cds)$UMAP_1_monocle=pbuild$data$data_dim_1 #add umap data for further analysis outside monocle
colData(cds)$UMAP_2_monocle=pbuild$data$data_dim_2 #add umap data for further analysis outside monocle

#counts=as.data.frame(exprs(cds))
counts=as.data.frame(normalized_counts(cds))
meta=as.data.frame(cds@colData@listData)


Rdata_filename = paste0(fname_prefix_R, "_monocle_cds_clustered_FINAL_PROTEINCODING_CLEAN.RData")
save(meta,counts,cds,
     file = Rdata_filename)

#clean environment
rm(combined_exp,data_exp5,expression_data,gene_annotation,expression_matrix,cds_1,cds_2,cds_3,cds_4,cds_5)
```
#make graphs to visualize genes
```{r}

for (gene_set_name in names(gene_lists)) {
  
  genes <- gene_lists[[gene_set_name]] # Extract the genes from list
  
  p <- plot_cells(
    cds,
    genes = genes,
    label_cell_groups = FALSE,
    show_trajectory_graph = FALSE
  ) + 
    labs(
      x = "UMAP 1", 
      y = "UMAP 2", 
      title = paste("UMAP Plot -", gene_set_name)
    ) +
    scale_color_viridis(option = "E", discrete = FALSE)
  
  ggsave(
    filename = paste0(fname_prefix_dotplot, "_", gene_set_name, "_clean.pdf"),
    plot = p,
    width = 5, 
    height = 5
  )
}

#single plot
# plot_cells(cds,
 #           genes = c("SOX2","TBXT"),
  #          label_cell_groups = FALSE,
   #         show_trajectory_graph = FALSE)+ 
    # labs(x = "UMAP 1", y = "UMAP 2", title = "")+
     #scale_color_viridis(option="E", discrete=F)
 
 #ggsave(filename = paste0(fname_prefix_dotplot, "_", "UMAP_Mesoderm_progenitors.pdf"),
   #  width = 10, height = 10)

ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes(color = monocle_clusters),
               size=0.8)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clusters")+
  scale_color_manual(name = "monocle",
                        values=c(okabe_pal))+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_monocle_clusters_clean.pdf"),
     width = 8, height = 6)

plot_feature=c("percent.mt", "percent.ribo","nUMI","nGene")

#ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
 #   geom_point(aes(color = nGene),
  #             size=0.8)+
  #labs(x = "UMAP 1", y = "UMAP 2", title = "nGene")+
  #scale_color_viridis(option="E", discrete=F)+facet_grid(~sample)
#ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_nGene.pdf"),
 #    width = 8, height = 6)

for (feature in plot_feature) {
  
  p <- ggplot(meta, aes(x = UMAP_1_monocle, y = UMAP_2_monocle)) +
    geom_point(aes_string(color = feature), size = 0.8) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      title = feature
    ) +
    scale_color_viridis(option = "E", discrete = FALSE) +
    facet_grid(~sample)
  
  ggsave(
    filename = paste0(fname_prefix_UMAP, "_", "UMAP_", feature, "_clean.pdf"),
    plot = p,
    width = 8,
    height = 4
  )
}

```
##get top 25 genes per cluster

```{r}

marker_test_res = top_markers(cds, group_cells_by="monocle_clusters", reference_cells=NULL, cores=16)
marker_test_res <- top_markers(cds, group_cells_by="monocle_clusters", 
                               reference_cells=10000, cores=8)

top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(10, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t10=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t10)=c("ids", "cluster")

write.csv(t10, file= paste0(fname_prefix_csv, "_", "top_10genes_per_cluster_clean.csv"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="monocle_clusters",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_dotplot, "_", "top10_heatmap_clean.pdf"),
     width = 8, height = 8)



top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(25, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t25=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t25)=c("ids", "cluster")

write.csv(t25, file= paste0(fname_prefix_csv, "_", "top_25genes_per_cluster_clean.csv"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="monocle_clusters",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_dotplot, "_", "top25_heatmap_clean.pdf"),
     width = 8, height = 15)

```