---
title: "n1 2D analysis exp5 exp7 combined: QC and monocle analysis"
author: "Elisa Balmas PhD"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up environment

```{r}
library(monocle3)
library(tidyr)
library(stringr)
library(grid)
library(viridis)
library(viridisLite)
library(ggthemes)
library(dplyr)
library(tidyverse)
library(ggsignif)
library(umap)
library(heatmap3)
library(plyr)
library(edgeR)
library(compareGroups)
library(dbscan)
library(MAST)
library(geosphere)
library(RColorBrewer)
library(Seurat)
library(htmlwidgets)
library(Matrix)
library(tidyseurat)
library(SeuratWrappers)
library(scMAGeCK)
library(readxl)
library(rCASC)
library(deepToolsUtils)
library(R.utils)


library(ggplot2); library(reshape2); theme_set(theme_bw(26) + theme(panel.grid.major = element_blank(), 
                                                                    panel.grid.minor = element_blank()) +
                                                 theme(legend.key = element_blank()))
update_geom_defaults("point", aes(size = 4))

#setwd("/home/ebalmas/scpHUB/projects/H001AS7_iKDseq_10X_4tp_monocle/analysis_EB")

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))


# Set up filename prefixes and output folders with the data
dir.create("Output")
dir.create(file.path("Output","Fig5_input_folder"))
dir.create(file.path("Output","QC"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito"))

dir.create(file.path("Output","monocle"))#create forders MONOCLE
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots","single"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "stat"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "UMAP"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "mosaic"))
dir.create(file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "heatmap"))

dir.create(file.path("Output","Seurat"))#create forders Seurat
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "dotplots"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "PCA"))
dir.create(file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "UMAP"))


fname_Fig5_input_folder <- file.path("Output","Fig5_input_folder")
#set up folder path QC
fname_prefix_R_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_csv_QC<-file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_ribomito_QC <- file.path("Output","QC", format(Sys.Date(), "%y%m%d"), "ribomito", 
                               format(Sys.Date(), "%y%m%d"))

#set up forder path MONOCLE
fname_prefix_R <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot_single <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", "single", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_csv<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "csv", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_stat<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "stat", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_UMAP<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "UMAP", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_mosaic<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "mosaic", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_heatmap<-file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "heatmap", 
                               format(Sys.Date(), "%y%m%d"))

#set up forder path Seurat
fname_prefix_R_Seurat <- file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_dotplot_Seurat <- file.path("Output","monocle", format(Sys.Date(), "%y%m%d"), "dotplots", 
                                                 format(Sys.Date(), "%y%m%d"))
fname_prefix_PCA_Seurat<-file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "PCA", 
                               format(Sys.Date(), "%y%m%d"))

fname_prefix_UMAP_Seurat<-file.path("Output","Seurat", format(Sys.Date(), "%y%m%d"), "UMAP", 
                               format(Sys.Date(), "%y%m%d"))

okabe_tab=read_excel("Output/Fig5_input_folder/color_scale.xlsx")
okabe_pal=okabe_tab$hex

okabe_pal=c("#E69F00","#56B4E9","#009E73","#f0E442","#0072B2","#D55E00","#CC79A7","#000000")
gray_pal=c("#000000","#666666","#cccccc","#ffffff")

pal <- c("#000000","#004949","#009292","#ff6db6","#ffb6db",
         "#490092","#006ddb","#b66dff","#6db6ff","#b6dbff",
         "#920000","#924900","#db6d00","#24ff24","#ffff6d")
```

########load data

```{r}
load("Output/Fig5_input_folder/241126_monocle_cds_clustered_FINAL_PROTEINCODING_CLEAN.RData")

#write.csv(counts, file= paste0(fname_prefix_csv, "_", "normalized_counts.csv"), row.names=T)
#write.csv(meta, file= paste0(fname_prefix_csv, "_", "cell_metadata.csv"), row.names=T)
#raw_counts <- as.data.frame(counts(cds, norm=F))
#write.csv(raw_counts, file= paste0(fname_prefix_csv, "_", "raw_counts.csv"), row.names=T)

```

####make polished plots

```{r}
colData(cds)$assigned_cell_type <- as.character(clusters(cds))
colData(cds)$assigned_cell_type_short <- as.character(clusters(cds))

colData(cds)$assigned_cell_type <- dplyr::recode(colData(cds)$assigned_cell_type,
                                                 "1"="Myoblasts",
                                                 "2"="Myocytes",
                                                 "3"="Posterior neuroectoderm",
                                                 "4"="Mesenchimal progenitors")

colData(cds)$assigned_cell_type_short <- dplyr::recode(colData(cds)$assigned_cell_type_short,
                                                 "1"="Myob",
                                                 "2"="Myoc",
                                                 "3"="P-N",
                                                 "4"="Mes-P")

###prepare color/time order vectors
cell_order=c("Myocytes","Myoblasts","Mesenchimal progenitors","Posterior neuroectoderm")
cell_order_short=c("Myoc","Myob","Mes-P","P-N")
####prepare table with cluster assignment
ct=c("Myoblasts","Myocytes","Posterior neuroectoderm","Mesenchimal progenitors")

ct_sh=c("Myob","Myoc","P-N","Mes-P")

clust_df=data.frame(monocle_cluster=c(1:4),cell_types=ct,cell_types_abbreviations=ct_sh)

clust_df$colors=pal=c("#000000","#E69F00","#56B4E9","#CC79A7")
clust_df$colors_g=gray_pal

###prepare color/time order vectors

monocle_long=data.frame(cell_order=c("Mesenchimal progenitors","Myoblasts","Myocytes","Posterior neuroectoderm"),order_LONG=c(1:4))

monocle_short=data.frame(cell_order_short=c("Mes-P","Myob","Myoc","P-N"),order_SHORT=c(1:4))

color_df=data.frame(cell_order_short=cell_order_short,cell_order=cell_order)
color_df$cell_types=color_df$cell_order
color_df$cell_types_abbreviations=color_df$cell_order_short
color_df=color_df%>%left_join(clust_df,by=c("cell_types","cell_types_abbreviations"))

color_df=color_df%>%left_join(monocle_short,by="cell_order_short")%>%left_join(monocle_long,by="cell_order")

color_df$cl_col=color_df$colors
color_df$order=rownames(color_df)

color_df_l=color_df[order(color_df$order_LONG),]
color_df_s=color_df[order(color_df$order_SHORT),]

#cds_subset=cds[,cds$libID%in%libID_downsample]

plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8,show_trajectory_graph=F)+
  scale_color_manual(name="cell types",values=color_df_l$cl_col)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering",colour="cell types")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters.pdf"),
     width = 8, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type_short", group_cells_by="assigned_cell_type_short",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8,show_trajectory_graph=F)+
  scale_color_manual(name="cell types",values=color_df_s$cl_col)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering",colour="cell types")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_sh.pdf"),
     width = 8, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8,show_trajectory_graph=F)+
  scale_color_manual(name="cell types",values=color_df_l$colors_g)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering",colour="cell types")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_gray.pdf"),
     width = 8, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type_short", group_cells_by="assigned_cell_type_short",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8,show_trajectory_graph=F)+
  scale_color_manual(name="cell types",values=color_df_s$colors_g)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering",colour="cell types")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_sh_gray.pdf"),
     width = 8, height = 5)
```



####make dot plots with gene expression

```{r}

show=c("POU5F1", "SOX2", "CDX2", "PTPRZ1", "PLIN2","UGCG","PAX7", "CCND2", "MYOD1", "MYH7B", "MB", "MYLK")
plot_cells(cds,
           genes=show,
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE)+ scale_color_viridis(option="E", discrete=F)

ggsave(filename = paste0(fname_prefix_dotplot, "_", "myo_general.pdf"),
     width = 5, height = 4)



```


##get top 25 genes per cluster

```{r}

marker_test_res = top_markers(cds, group_cells_by="monocle_clusters", reference_cells=NULL, cores=16)
marker_test_res <- top_markers(cds, group_cells_by="monocle_clusters", 
                               reference_cells=10000, cores=8)

top_specific_markers <- marker_test_res %>%
                            filter(fraction_expressing >= 0.10) %>%
                            group_by(cell_group) %>%
                            top_n(25, pseudo_R2)

top_specific_marker_ids <- unique(top_specific_markers %>% dplyr::pull(gene_id))
t25=data.frame(top_specific_markers$gene_id, top_specific_markers$cell_group, stringsAsFactors = F)
colnames(t25)=c("ids", "cluster")

write.csv(t25, file= paste0(fname_prefix_csv, "_", "top_25genes_per_cluster.CSV"), row.names=FALSE)

plot_genes_by_group(cds,
                    top_specific_marker_ids,
                    group_cells_by="assigned_cell_type",
                    ordering_type="maximal_on_diag",
                    max.size=3)+ scale_color_viridis(option="E", discrete=F)
ggsave(filename = paste0(fname_prefix_dotplot, "_", "top25_heatmap.pdf"),
     width = 8, height = 20)

```


###calculate % cells of one gene in each cluster

```{r}

meta_to_use=as.data.frame(cds@colData@listData)

tot=table(meta_to_use$sample) #total of the cluster
df=table(meta_to_use$sample,meta_to_use$monocle_clusters)
df = as.data.frame(df)
colnames(df)=c("sample","monocle_cluster","freq")

#mdf2 = melt(df)
mdf2=df#11664 rows
total=as.data.frame(tot)
colnames(total)=c("sample","tot")
mdf2=mdf2%>%left_join(total, by="sample")
mdf2$tot = as.numeric(mdf2$tot)
mdf2$frxn = (mdf2$freq/mdf2$tot)*100
m=subset(mdf2,mdf2$freq!=0)
#m <- m %>%
 # mutate(samp = ifelse(grepl("n3_", sampleID), "n3", "n4"))

m$monocle_cluster=as.integer(as.character(m$monocle_cluster))

m=m%>%dplyr::left_join(color_df, by="monocle_cluster")
m$monocle_cluster=as.character(m$monocle_cluster)

#####graph overview

ggplot(m, aes(x = factor(cell_types,levels=color_df_l$cell_order), y = frxn, fill=cell_types)) + geom_bar(stat = "identity") +theme(legend.position='none')+
  scale_fill_manual(values =color_df_l$cl_col)+
  facet_grid(~sample)
  
ylab = "%  by cell type"
xlab = "cluster"

last_plot() + labs(x = xlab, y = ylab)
ggsave(filename = paste0(fname_prefix_mosaic, "_", "percent_cells.pdf"),
     width = 16, height = 15)

####graph by gene

ggplot(m, aes(x = sample, y = frxn, fill=cell_types)) + geom_bar(stat = "identity") +
  scale_fill_manual(name="cell type",values =color_df_l$cl_col)
  
ylab = "% by cell type"
xlab = "cell line"
last_plot() + labs(x = xlab, y = ylab)
output_file <- paste0(fname_prefix_mosaic, "_", "percent_cell_type_mosaic.pdf")
ggsave(filename = output_file, width = 5, height = 5)

######calculate ratio KD vs CTR
CIRI=subset(m,m$sample=="CIRI")%>%select(-sample)%>%dplyr::rename(n_cells_CIRI=freq,tot_CIRI=tot,percent_CIRI=frxn)

CRISPRa=subset(m,m$sample=="CRISPRa")%>%dplyr::rename(n_cells_CRISPRa=freq,tot_CRISPRa=tot,percent_CRISPRa=frxn)%>%select(-sample)### NB. one cluster missing

m_ratio=CIRI%>%dplyr::left_join(CRISPRa,by=c(colnames(color_df)))
m_ratio=m_ratio%>%mutate(FC=percent_CIRI/percent_CRISPRa)
m_ratio$monocle_cluster=as.integer(as.character(m_ratio$monocle_cluster))


#####graph overview FC
ggplot(m_ratio, aes(x = factor(cell_types,levels=color_df_l$cell_order), y = log2(FC), fill=cell_types)) + geom_bar(stat = "identity") +theme(legend.position='none')+
  geom_hline(yintercept=1, linetype="dashed", color = "red")+
  geom_hline(yintercept=-1, linetype="dashed", color = "red")+
  scale_fill_manual(values =color_df_l$cl_col)#+
  #facet_grid(~factor(gene, levels = gene_order))
  
ylab = "log2 (% CIRI vs CRISPRa)"
xlab = ""

last_plot() + labs(x = xlab, y = ylab)
ggsave(filename = paste0(fname_prefix_mosaic, "_", "FC_cells_clust.pdf"),
     width = 5, height = 4)

###save data
stat_filename = paste0(fname_prefix_stat, "_nb_percent_in_each_cluster_polished.csv")
write.csv(m, stat_filename, quote = F, row.names = T, sep=",")
stat_filename = paste0(fname_prefix_stat, "_nb_percent_in_each_cluster_FC_polished.csv")
write.csv(m_ratio, stat_filename, quote = F, row.names = T, sep=",")

Rdata_filename = paste0(fname_prefix_R, "_calculations_nb_percent_polished.RData")
save(m,m_ratio,file = Rdata_filename)

```
###graphs in gray


```{r}
#####graph overview

ggplot(m, aes(x = factor(cell_types,levels=color_df_l$cell_order), y = frxn, fill=cell_types)) + geom_bar(stat = "identity", colour="black") +theme(legend.position='none')+
  scale_fill_manual(values =color_df_l$colors_g)+
  facet_grid(~sample)
  
ylab = "%  by cell type"
xlab = "cluster"

last_plot() + labs(x = xlab, y = ylab)
ggsave(filename = paste0(fname_prefix_mosaic, "_", "percent_cells_gray.pdf"),
     width = 16, height = 15)

####graph by gene

ggplot(m, aes(x = sample, y = frxn, fill=cell_types)) + geom_bar(stat = "identity", colour="black") +
  scale_fill_manual(name="cell type",values =color_df_l$colors_g)
  
ylab = "% by cell type"
xlab = "cell line"
last_plot() + labs(x = xlab, y = ylab)
output_file <- paste0(fname_prefix_mosaic, "_", "percent_cell_type_mosaic_gray.pdf")
ggsave(filename = output_file, width = 5, height = 5)

#####graph overview FC
ggplot(m_ratio, aes(x = factor(cell_types,levels=color_df_l$cell_order), y = log2(FC), fill=cell_types)) + geom_bar(stat = "identity", colour="black") +theme(legend.position='none')+
  geom_hline(yintercept=1, linetype="dashed", color = "black")+
  geom_hline(yintercept=-1, linetype="dashed", color = "black")+
  scale_fill_manual(values =color_df_l$colors_g)#+
  #facet_grid(~factor(gene, levels = gene_order))
  
ylab = "log2 (% CIRI vs CRISPRa)"
xlab = ""

last_plot() + labs(x = xlab, y = ylab)
ggsave(filename = paste0(fname_prefix_mosaic, "_", "FC_cells_clust_gray.pdf"),
     width = 5, height = 4)
```


#####calculate fisher stats

```{r}
#make true false annotations per cluster to make the contingency tables

colData(cds)$clust1 = cds$monocle_clusters=="1"
colData(cds)$clust2 = cds$monocle_clusters=="2"
colData(cds)$clust3 = cds$monocle_clusters=="3"
colData(cds)$clust4 = cds$monocle_clusters=="4"

meta=as.data.frame(cds@colData@listData)

#meta=subset(meta,meta$has_70_counts==T)

#####clusters to consider are 1 to 4
#make a loop to create the statistics
################## start without subsetting for cell populations


meta_sh=meta%>%select(sample,clust1,clust2,clust3,clust4)

gene_list <- c("SMAD2")
result_list <- list()


for (cluster in paste0("clust", 1:4)) {
  p_values <- numeric()

  for (gene in gene_list) {
    subset_data <- meta_sh
    
    if (nrow(subset_data) > 0) {
      # Perform chi-squared test
      fisher_result <- fisher.test(subset_data$sample, subset_data[[cluster]], simulate.p.value = FALSE)
      
      # Extract p-value
      p_value <- fisher_result[["p.value"]]
      
      p_values <- c(p_values, p_value)
    } else {
      # If no data for the gene in the current cluster, set p-value to NA
      p_values <- c(p_values, NA)
    }
  }

  result_df <- data.frame(cluster = cluster, p_value = p_values, stringsAsFactors = FALSE)
  
  # Add a q-value column using p.adjust
  result_df$q_value <- p.adjust(result_df$p_value, method = "fdr")
  
  # Add a column indicating significance at a 5% level
  result_df$sig <- result_df$q_value <= 0.05

  result_list[[cluster]] <- result_df
}

combined_result_df <- do.call(rbind, result_list)
combined_result_df$q_value=p.adjust(combined_result_df$p_value, method = "fdr")

m_ratio$cluster=paste0("clust",m_ratio$monocle_cluster)
FC=m_ratio
combined_result_df=combined_result_df%>%left_join(FC, by=c("cluster"))

stat_filename = paste0(fname_prefix_stat, "_fisher_stat.csv")
write.csv(combined_result_df, stat_filename, quote = F, row.names = T, sep=",")


ggplot(data=combined_result_df, aes(x=log2(as.numeric(FC)), y=log10(q_value)*-1)) +
  geom_point(aes(colour = factor(cell_types,levels=color_df_l$cell_order)), size=1.5) +scale_colour_manual(values=color_df_l$colors)+
  ggrepel::geom_text_repel(
    data = subset(combined_result_df,sig==T&FC>1|sig==T&FC<0.5),
    aes(label = cell_types_abbreviations),
    size = 3,
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")
  )+
  #geom_vline(xintercept = log2(1.5),linetype="dotted")+
  #geom_vline(xintercept = -log2(1.5),linetype="dotted")+
  geom_hline(yintercept = -log10(0.05),linetype="dotted")+
  ggtitle("fisher test statistics in each cluster compared to SMAD2 CT")+
  theme_minimal() +  # You can change the theme as needed
  labs(
    x = "Log2FC compared to CTR",
    y = "-Log10 adj p-value",
    color = "cell type"  # Legend title
  ) 

ggsave(filename = paste0(fname_prefix_dotplot, "_volcano_fisher_stats.pdf"),
       width = 5, height = 4)



plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8, show_trajectory_graph =F)+
  scale_color_manual(name="monocle_cluster",values=color_df_l$cl_col)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering")
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters.pdf"),
     width = 6, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8, show_trajectory_graph =F)+
  scale_color_manual(name="monocle_cluster",values=color_df_l$cl_col)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_sample.pdf"),
     width = 8, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8, show_trajectory_graph =F)+
  scale_color_manual(name="monocle_cluster",values=color_df_l$colors_g)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering")
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_gray.pdf"),
     width = 6, height = 5)

plot_cells(cds, color_cells_by="assigned_cell_type", group_cells_by="assigned_cell_type",label_cell_groups = F,group_label_size = 7,label_roots=F,label_leaves=F,label_branch_points=F,cell_size=0.8, show_trajectory_graph =F)+
  scale_color_manual(name="monocle_cluster",values=color_df_l$colors_g)+
  labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle clustering")+facet_grid(~sample)
ggsave(filename = paste0(fname_prefix_UMAP, "_", "UMAP_clusters_sample_gray.pdf"),
     width = 8, height = 5)
```


####add regression analysis

```{r}
seurat<-as.Seurat(cds, assay = NULL)
Idents(seurat) <- "assigned_cell_type"
c <- FindAllMarkers(seurat, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)


colData(cds)$clust1 = cds$monocle_clusters=="1"
colData(cds)$clust2 = cds$monocle_clusters=="2"
colData(cds)$clust3 = cds$monocle_clusters=="3"
colData(cds)$clust4 = cds$monocle_clusters=="4"


# clust1 to clust9 are the cluster names, make a list with it
cluster_names <- paste0("clust", 1:4)

# Initialize empty lists to store results
gene_fits_list <- list()
fit_coefs_list <- list()
fit_coefs_sig_list <- list()

# Loop through each cluster
for (cluster_name in cluster_names) {
  # Fit models
  gene_fits <- fit_models(cds, model_formula_str = paste0("~", cluster_name))
  gene_fits_list[[cluster_name]] <- gene_fits
  
  # Extract coefficient tables
  fit_coefs <- coefficient_table(gene_fits)
  fit_coefs_list[[cluster_name]] <- fit_coefs
  
  # Filter significant coefficients
  fit_coefs_sig <- fit_coefs %>%
    filter(term == paste0(cluster_name, "TRUE")) %>%
    filter(q_value < 0.05) %>%
    select(gene_short_name, term, q_value, estimate, num_cells_expressed)
  
  fit_coefs_sig_list[[cluster_name]] <- fit_coefs_sig
}


# assign cluster names to a vector
cluster_names <- paste0("clust", 1:4)

# Initialize an empty list to store fit_coefs_renamed for each cluster
fit_coefs_list_renamed <- list()

# Loop through each cluster and rename the fit_coef dataframes for each cluster in order to then merge them into a single dataframe
for (i in 1:4) {
  cluster_name <- cluster_names[i]
  
  # Perform operations for each cluster
  fit_coefs <- fit_coefs_list[[cluster_name]] %>%
    dplyr::rename(logFC = estimate, q_value = q_value) %>%
    filter(term == paste0(cluster_name, "TRUE")) %>%
    select(gene_short_name, q_value, logFC, num_cells_expressed)
  
  # Rename columns consistently
  col_rename <- c("geneID", paste0("q_value_", i), paste0("logFC_", i), "num_cells_expressed")
  
  # Rename columns in fit_coefs
  fit_coefs_renamed <- fit_coefs %>%
    setNames(col_rename)
  
  # Store fit_coefs_renamed in the list
  fit_coefs_list_renamed[[cluster_name]] <- fit_coefs_renamed
}

de_stats<-fit_coefs_list_renamed[["clust1"]]%>%
  left_join(fit_coefs_list_renamed[["clust2"]], by =c("geneID","num_cells_expressed"))%>%
  left_join(fit_coefs_list_renamed[["clust3"]], by =c("geneID","num_cells_expressed"))%>%
  left_join(fit_coefs_list_renamed[["clust4"]], by =c("geneID","num_cells_expressed"))

de_stats$dup=duplicated(de_stats$geneID)
de_stats=subset(de_stats,de_stats$dup!=T)%>%dplyr::select(-dup)

analysis_clusters=de_stats

stat_filename = paste0(fname_prefix_csv, "_complete_gene_reg_by_clusters.csv")
write.csv(analysis_clusters, stat_filename, quote = F, row.names = T, sep=",")
```

#######add direction

```{r}
#Set the significance level cutoff
sig_level <- 0.05
#Set a fold-change cutoff
fc_level <- log2(1.5)
fc_level2 <- (log2(1.5))*-1
###################################

# add thresholds (up= upregulated; down= downregulated, other= neighter)
for (i in 1:9) {
  cluster_name <- paste0("clust", i)
  
  # Create column names dynamically
  logFC_col <- paste0("logFC_", i)
  q_value_col <- paste0("q_value_", i)
  treshold_col <- paste0("treshold_cl", i)
  Log10pvalue_col <- paste0("Log10pvalue_cl", i)
  
  # Apply thresholding logic
  de_stats[[treshold_col]] <- ifelse(de_stats[[logFC_col]] >= fc_level & de_stats[[q_value_col]] <= sig_level, "up",
                                      ifelse(de_stats[[logFC_col]] <= fc_level2 & de_stats[[q_value_col]] <= sig_level, "down", "other"))
  
  de_stats[[Log10pvalue_col]] <- -log10(de_stats[[q_value_col]])
}

####create to_show columns
####here are the specific upregulated genes for each cluster and not by others
# Loop through each cluster
for (i in 1:9) {
  cluster_name <- paste0("clust", i)
  
  # Create column names dynamically
  treshold_col <- paste0("treshold_cl", i)
  to_show_col <- paste0("to_show_cl", i)
  
  # Create condition dynamically
  condition <- paste0(
    "de_stats$", to_show_col, "<-ifelse(",
    paste(
      sprintf("de_stats$treshold_cl%d == 'up' & (", i),
      paste(sprintf("de_stats$treshold_cl%d == 'down' | de_stats$treshold_cl%d == 'other'", 1:9, 1:9), collapse = " | "),
      ")"
    ),
    ", de_stats$geneID, \"\")"
  )
  
  # Evaluate the condition
  eval(parse(text = condition))
}

stat_filename = paste0(fname_prefix_csv, "_gene_reg_with_directions.csv")
write.csv(de_stats, stat_filename, quote = F, row.names = T, sep=",")

Rdata_filename = paste0(fname_prefix_csv, "_gene_regression_every_clust_against_eachother.RData")
save(de_stats,
     file = Rdata_filename)
```

####make volcano plots

```{r}
###########
de_stats=de_stats%>% distinct(geneID, .keep_all = TRUE)#eliminate duplicates

# Loop through clusters to make all the volcano plots in bulk than use the one below to make specific ones
for (i in 1:9) {
  logFC_col <- paste0("logFC_", i)
  treshold_col <- paste0("treshold_cl", i)
  Log10pvalue_col <- paste0("Log10pvalue_cl", i)
  q_value_col <- paste0("q_value_", i)
  to_show_col <- paste0("to_show_cl", i)

  volcano_plot <- ggplot(data = de_stats, aes(x = !!rlang::sym(logFC_col), y = !!rlang::sym(Log10pvalue_col))) +
    geom_point(aes(colour = as.factor(!!rlang::sym(treshold_col))), size = 1.5) +
    scale_colour_manual(values = c("up" = "red", "down" = "blue", "other" = "grey")) +
     ggrepel::geom_text_repel(
  data = de_stats %>%
    filter(!!rlang::sym(q_value_col) < 0.05),
  aes(label = !!rlang::sym(to_show_col)),
  size = 3,
  box.padding = unit(0.2, "lines"),
  point.padding = unit(0.2, "lines"),
  segment.color = "grey",
  segment.size = 0.2
)+
    theme(legend.position = "none") +
    geom_hline(yintercept = (log10(sig_level) * -1)) +
    geom_vline(xintercept = c(fc_level, fc_level2)) +
    xlim(c(-4, 4)) + ylim(c(0, 200)) +
    xlab("log2 fold change") + ylab("-Log10 adj p-value")+
  annotate("text", x = 4, y = 200, label = paste("Cluster ", i), col = "black", size = 4, hjust = 1, vjust = 1) +
    # Add "other clusters" on the top left for clusters other than the current one
    annotate("text", x = -4, y = 200, label = "Other Clusters", col = "black", size = 4, hjust = 0, vjust = 1)


  ggsave(
    filename = paste0(fname_prefix_dotplot, "_cl", i, ".png"),
    plot = volcano_plot,
    width = 8,
    height = 6
  )
}
```

##### reaname clusters

```{r}
colData(cds)$assigned_cell_type <- as.character(clusters(cds))
colData(cds)$assigned_cell_type_short <- as.character(clusters(cds))

colData(cds)$assigned_cell_type <- dplyr::recode(colData(cds)$assigned_cell_type,
                                                 "1"="Cardiac fibroblasts",
                                                 "2"="Late cardiomyocytes",
                                                 "3"="Early cardiomyocytes",
                                                 "4"="Cardiac progenitors",
                                                 "5"="Lateral plate mesoderm",
                                                 "6"="Proliferating cardiomyocytes",
                                                 "7"="Endothelial cells",
                                                 "8"="Mesendoderm",
                                                 "9"="Endoderm derivatives")

colData(cds)$assigned_cell_type_short <- dplyr::recode(colData(cds)$assigned_cell_type_short,
                                                 "1"="CF",
                                                 "2"="L-CM",
                                                 "3"="E-CM",
                                                 "4"="CP",
                                                 "5"="LPM",
                                                 "6"="P-CM",
                                                 "7"="EC",
                                                 "8"="MES",
                                                 "9"="ENDO")

cds_cardio= cds[,cds@colData@listData[["monocle_clusters"]]=="4"|cds@colData@listData[["monocle_clusters"]]=="3"|cds@colData@listData[["monocle_clusters"]]=="2"]

meta=as.data.frame(cds@colData@listData)
meta_cardio=as.data.frame(cds_cardio@colData@listData)

Rdata_filename = paste0(fname_prefix_csv, "_updated_cds_with_modules.RData")
save(meta,meta_cardio,cds,cds_cardio,
     file = Rdata_filename)
```


###get gene modules shanging along pseudotime

```{r}

cds@colData$group=paste0(cds@colData$shRNA_gene,"_",cds@colData$knockdown)
cds_cardio@colData$group=paste0(cds_cardio@colData$shRNA_gene,"_",cds_cardio@colData$knockdown)

#########find gene modules along trajectory
cardio_cds_pr_test_res <- graph_test(cds, neighbor_graph="principal_graph", cores=4)#score genes along the trajectory
genes_mod_pseudotime <- row.names(subset(cardio_cds_pr_test_res, q_value < 0.05))

gene_module_df <- find_gene_modules(cds[genes_mod_pseudotime,], reduction_method = c("UMAP"),resolution=0.01)#organize genes into gene modules

cell_group_df <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$monocle_clusters)###group modules by cluster
cell_group_df_test <- tibble::tibble(cell=row.names(colData(cds)), 
                                cell_group=colData(cds)$group)###group modules by shRNA hene and KD (no selection yet)

agg_mat <- aggregate_gene_expression(cds, gene_module_df)###every lib ID has a value per each module

modules=as.data.frame(t(agg_mat))
modules <- modules[, order(names(modules))]
modules=as.data.frame(modules)

cds@colData[, paste0("module_", 1:74)] <- modules[, 1:74]###added 74 modules
meta[, paste0("module_", 1:74)] <- modules[, 1:74]

#colnames(cds_cardio)

agg_mat_cluster <- aggregate_gene_expression(cds, gene_module_df, cell_group_df)
row.names(agg_mat_cluster) <- stringr::str_c("Module ", row.names(agg_mat))
hp1=pheatmap::pheatmap(agg_mat_cluster,
                   scale="row", clustering_method="ward.D2")

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_cl.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp1
dev.off()

agg_mat_test <- aggregate_gene_expression(cds, gene_module_df, cell_group_df_test)
row.names(agg_mat_test) <- stringr::str_c("Module ", row.names(agg_mat_test))
hp2=pheatmap::pheatmap(agg_mat_test,
                   scale="row", clustering_method="ward.D2",main="row normalized modules all",color = viridis(n = 256, alpha = 1,begin = 0, end = 1, option = "H"))

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_all_modules.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp2
dev.off()

Rdata_filename = paste0(fname_prefix_R, "_gene_modules_pseudotime.RData")
save(cardio_cds_pr_test_res,genes_mod_pseudotime, gene_module_df, modules,agg_mat,agg_mat_cluster,agg_mat_test,
     file = Rdata_filename)

#########find gene modules along trajectory for only cardiac clusters
subset_pr_test_res <- graph_test(cds_cardio, neighbor_graph="principal_graph", cores=4)#score genes along the trajectory
genes_mod_pseudotime_cardio <- row.names(subset(subset_pr_test_res, q_value < 0.05))

gene_module_df_cardio <- find_gene_modules(cds_cardio[genes_mod_pseudotime_cardio,], resolution=0.01)#organize genes into gene modules

cell_group_df_c <- tibble::tibble(cell=row.names(colData(cds_cardio)), 
                                cell_group=colData(cds_cardio)$monocle_clusters)###group modules by cluster
cell_group_df_test_c <- tibble::tibble(cell=row.names(colData(cds_cardio)), 
                                cell_group=colData(cds_cardio)$group)###group modules by shRNA hene and KD (no selection yet)

agg_mat_cardio <- aggregate_gene_expression(cds_cardio, gene_module_df_cardio)###every lib ID has a value per each module
modules_cardio=as.data.frame(t(agg_mat_cardio))
#modules=as.data.frame(t(agg_mat))

cds_cardio@colData[, paste0("module_", 1:59,"_cardio")] <- modules_cardio[, 1:59]###added 40 modules

agg_mat_cardio_all <- aggregate_gene_expression(cds, gene_module_df_cardio)###every lib ID has a value per each module
modules_cardio_all=as.data.frame(t(agg_mat_cardio_all))
cds@colData[, paste0("module_", 1:59,"_cardio")] <- modules_cardio_all[, 1:59]###added 59 modules

module_dendro <- hclust(dist(agg_mat_cardio))
gene_module_df_cardio$module <- factor(gene_module_df_cardio$module, 
                                levels = row.names(agg_mat_cardio)[module_dendro$order])

agg_mat_cardio_cluster <- aggregate_gene_expression(cds_cardio, gene_module_df_cardio, cell_group_df_c)
row.names(agg_mat_cardio_cluster) <- stringr::str_c("Module ", row.names(agg_mat_cardio_cluster))

hp3=pheatmap::pheatmap(agg_mat_cardio_cluster,
                   scale="column", clustering_method="ward.D2")

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_cardio_modules_cl.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp3
dev.off()

agg_mat_cardio_test <- aggregate_gene_expression(cds_cardio, gene_module_df_cardio, cell_group_df_test_c)
row.names(agg_mat_cardio_test) <- stringr::str_c("Module ", row.names(agg_mat_cardio_test))

hp4=pheatmap::pheatmap(agg_mat_cardio_test,
                   scale="row", clustering_method="ward.D2")

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_cardio_modules.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp4
dev.off()

agg_mat_cardio_test_all <- aggregate_gene_expression(cds, gene_module_df_cardio, cell_group_df_test_c)
row.names(agg_mat_cardio_test_all) <- stringr::str_c("Module ", row.names(agg_mat_cardio_test_all))

hp5=pheatmap::pheatmap(agg_mat_cardio_test_all,
                   scale="row", clustering_method="ward.D2")

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_cardio_modules_all.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp5
dev.off()

Rdata_filename = paste0(fname_prefix_R, "_gene_modules_pseudotime_cardio.RData")
save(subset_pr_test_res,genes_mod_pseudotime_cardio, gene_module_df_cardio, modules_cardio,agg_mat_cardio,agg_mat_cardio_all,agg_mat_cardio_cluster,agg_mat_cardio_test,
     file = Rdata_filename)

#cds_cardio= cds[,cds@colData@listData[["monocle_clusters"]]=="4"|cds@colData@listData[["monocle_clusters"]]=="3"|cds@colData@listData[["monocle_clusters"]]=="2"]

cds@colData@listData <- replace(cds@colData@listData, is.na(cds@colData@listData), "empty")

meta=as.data.frame(cds@colData@listData)
meta <- replace(meta, is.na(meta), "empty")
meta_cardio=as.data.frame(cds_cardio@colData@listData)
meta_cardio <- replace(meta_cardio, is.na(meta_cardio), "empty")

Rdata_filename = paste0(fname_prefix_R, "_updated_cds_with_modules.RData")
save(meta,meta_cardio,cds,cds_cardio,
     file = Rdata_filename)

```

####plot modules

```{r}
plot_cells(cds,
           genes=gene_module_df_cardio %>% filter(module %in% c(1:59)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE,cell_size=1)+
labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle top modules cardio clusters")
last_plot() + scale_color_viridis(option="E", discrete=F)

ggsave(filename = paste0(fname_prefix_UMAP, "_", "Monocle_cardio_modules.pdf"),
     width = 21, height = 30)

plot_cells(cds,
           genes=gene_module_df %>% filter(module %in% c(1:74)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE,cell_size=1)+
labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle top modules")
last_plot() + scale_color_viridis(option="E", discrete=F)

ggsave(filename = paste0(fname_prefix_UMAP, "_", "Monocle_modules.pdf"),
     width = 21, height = 30)


```
#########module selection
cardio

```{r}
####eliminate the modules with expression in only few cells abd general low expression
select_modules=as.data.frame(agg_mat_cardio_test)
select_modules=select_modules%>%dplyr::filter(rownames(select_modules)!="Module 59")#####select modules with decent amount of cells expressing genes

hp1=pheatmap::pheatmap(select_modules,
                   scale="row", clustering_method="ward.D2",main="row normalized modules filtered")#scaled on rows (modules) >>>to compare modules between perturbations
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_filtered.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp1
dev.off()
hp2=pheatmap::pheatmap(agg_mat_cardio_test,
                   scale="row", clustering_method="ward.D2")#scaled on row (modules) ###original data >>>to compare modules between perturbations

hp1=pheatmap::pheatmap(select_modules,
                   scale="row", clustering_method="ward.D2",main="row normalized modules filtered",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on rows (modules) >>>to compare modules between perturbations
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_filtered_TURBO.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp1
dev.off()


#######get z score of the modules (the values of the heatmap hp1 and hp2)
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

data_subset_norm <- t(apply(select_modules, 1, cal_z_score)) ###get z scored data
hp3=pheatmap::pheatmap(data_subset_norm, clustering_method="ward.D2")###check hp4 is the same as hp2

data_subset_norm=as.data.frame(data_subset_norm)
data_subset_norm$min<-apply(data_subset_norm,1,FUN=min)
data_subset_norm$max<-apply(data_subset_norm,1,FUN=max)
data_subset_norm$delta=data_subset_norm$max-data_subset_norm$min

top=data_subset_norm[order(-data_subset_norm$delta),] ####select top 10 based on delta
top=head(top, n=16)

select_modules=select_modules%>%dplyr::filter(rownames(select_modules)%in%rownames(top))

rownames(top)
# "Module 5"  "Module 55" "Module 9"  "Module 21" "Module 33" "Module 34" "Module 57" "Module 16" "Module 56" "Module 14" "Module 26" "Module 46" "Module 18" "Module 23" "Module 54""Module 49"

plot_cells(cds,
           genes=gene_module_df_cardio %>% filter(module %in% c(5,55,9,21,33,34,57,16,56,14,26,46,18,23,54,49)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE,cell_size=1)+
labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle top 16 modules")
last_plot() + scale_color_viridis(option="E", discrete=F)

ggsave(filename = paste0(fname_prefix_UMAP, "_", "Monocle_top_16_modules.pdf"),
     width = 10, height = 10)

##make the same hm as before but flipped
select_modules=t(select_modules)#flip data do have genes in rows and modules in columns

hp3=pheatmap::pheatmap(select_modules,
                   scale="column", clustering_method="ward.D2",main="Top 16 modules")#scaled on row (modules)
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_top16.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp3
dev.off()

hp3=pheatmap::pheatmap(select_modules,
                   scale="column", clustering_method="ward.D2",main="Top 16 modules",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on row (modules)
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_top16_Turbo.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp3
dev.off()

module_final_selection=c("Module 5","Module 55","Module 9","Module 21","Module 33","Module 34","Module 57","Module 16","Module 56","Module 14","Module 26","Module 46","Module 18","Module 23","Module 54","Module 49")

select_modules_FINAL=as.data.frame(agg_mat_cardio_test)
select_modules_FINAL=select_modules_FINAL%>%dplyr::filter(rownames(select_modules_FINAL)%in%module_final_selection)
select_modules_FINAL=t(select_modules_FINAL)
hp2=pheatmap::pheatmap(select_modules_FINAL,
                   scale="column", clustering_method="ward.D",main="Top selected modules")#scaled on row (modules)

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_selected_FINAL_norm_ROW.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp2
dev.off()

hp2=pheatmap::pheatmap(select_modules_FINAL,
                   scale="column", clustering_method="ward.D",main="Top selected modules",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on row (modules)

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_selected_FINAL_norm_ROW_TURBO.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp2
dev.off()


Rdata_filename = paste0(fname_prefix_R, "_updated_cds_with_modules_SELECTION.RData")
save(meta,cds,cds_cardio,meta_cardio,agg_mat,agg_mat_test,modules,cardio_cds_pr_test_res,genes_mod_pseudotime,gene_module_df,cell_group_df_test,select_modules,module_final_selection,select_modules_FINAL,
     file = Rdata_filename) 

side=read.csv(paste0(fname_Fig5_input_folder,"/clone_side.csv"))

clones=side$clone
side=side%>%distinct(clone,side)
meta=meta%>%left_join(side, by="clone")#5858
meta$side=meta$side%>% replace_na("unknown")
cds@colData$side=meta$side

Rdata_filename = paste0(fname_prefix_R, "_updated_cds_with_modules_SELECTION_side.RData")
save(meta,cds,cds_cardio,meta_cardio,agg_mat,agg_mat_test,modules,cardio_cds_pr_test_res,genes_mod_pseudotime,gene_module_df,cell_group_df_test,select_modules,module_final_selection,select_modules_FINAL,
     file = Rdata_filename)

```
#########module selection
all

```{r}
####eliminate the modules with expression in only few cells abd general low expression
select_modules=as.data.frame(agg_mat_test)
select_modules=select_modules%>%dplyr::filter(rownames(select_modules)!="Module 73"&rownames(select_modules)!="Module 74"&rownames(select_modules)!="Module 58")#####select modules with decent amount of cells expressing genes

hp1=pheatmap::pheatmap(select_modules,
                   scale="row", clustering_method="ward.D2",main="row normalized modules filtered",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on rows (modules) >>>to compare modules between perturbations
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_filtered.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp1
dev.off()
hp2=pheatmap::pheatmap(agg_mat_cardio_test,
                   scale="row", clustering_method="ward.D2")#scaled on row (modules) ###original data >>>to compare modules between perturbations

hp1=pheatmap::pheatmap(select_modules,
                   scale="row", clustering_method="ward.D2",main="row normalized modules filtered",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on rows (modules) >>>to compare modules between perturbations
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_filtered_TURBO.pdf")
pdf(Rdata_filename, height = 10, width = 8)
hp1
dev.off()


#######get z score of the modules (the values of the heatmap hp1 and hp2)
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

data_subset_norm <- t(apply(select_modules, 1, cal_z_score)) ###get z scored data
hp3=pheatmap::pheatmap(data_subset_norm, clustering_method="ward.D2")###check hp4 is the same as hp2

data_subset_norm=as.data.frame(data_subset_norm)
data_subset_norm$min<-apply(data_subset_norm,1,FUN=min)
data_subset_norm$max<-apply(data_subset_norm,1,FUN=max)
data_subset_norm$delta=data_subset_norm$max-data_subset_norm$min

top=data_subset_norm[order(-data_subset_norm$delta),] ####select top 10 based on delta
top=head(top, n=16)

select_modules=select_modules%>%dplyr::filter(rownames(select_modules)%in%rownames(top))

rownames(top)
# "Module 5"  "Module 55" "Module 9"  "Module 21" "Module 33" "Module 34" "Module 57" "Module 16" "Module 56" "Module 14" "Module 26" "Module 46" "Module 18" "Module 23" "Module 54""Module 49"

plot_cells(cds,
           genes=gene_module_df_cardio %>% filter(module %in% c(5,55,9,21,33,34,57,16,56,14,26,46,18,23,54,49)),
           label_cell_groups=FALSE,
           show_trajectory_graph=FALSE,cell_size=1)+
labs(x = "UMAP 1", y = "UMAP 2", title = "Monocle top 16 modules")
last_plot() + scale_color_viridis(option="E", discrete=F)

ggsave(filename = paste0(fname_prefix_UMAP, "_", "Monocle_top_16_modules.pdf"),
     width = 10, height = 10)

##make the same hm as before but flipped
select_modules=t(select_modules)#flip data do have genes in rows and modules in columns

hp3=pheatmap::pheatmap(select_modules,
                   scale="column", clustering_method="ward.D2",main="Top 16 modules")#scaled on row (modules)
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_top16.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp3
dev.off()

hp3=pheatmap::pheatmap(select_modules,
                   scale="column", clustering_method="ward.D2",main="Top 16 modules",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on row (modules)
Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_top16_Turbo.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp3
dev.off()

module_final_selection=c("Module 5","Module 55","Module 9","Module 21","Module 33","Module 34","Module 57","Module 16","Module 56","Module 14","Module 26","Module 46","Module 18","Module 23","Module 54","Module 49")

select_modules_FINAL=as.data.frame(agg_mat_cardio_test)
select_modules_FINAL=select_modules_FINAL%>%dplyr::filter(rownames(select_modules_FINAL)%in%module_final_selection)
select_modules_FINAL=t(select_modules_FINAL)
hp2=pheatmap::pheatmap(select_modules_FINAL,
                   scale="column", clustering_method="ward.D",main="Top selected modules")#scaled on row (modules)

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_selected_FINAL_norm_ROW.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp2
dev.off()

hp2=pheatmap::pheatmap(select_modules_FINAL,
                   scale="column", clustering_method="ward.D",main="Top selected modules",color = viridis(n = 256, alpha = 1, 
                                   begin = 0, end = 1, option = "H"))#scaled on row (modules)

Rdata_filename = paste0(fname_prefix_heatmap, "_hp_modules_selected_FINAL_norm_ROW_TURBO.pdf")
pdf(Rdata_filename, height = 4, width = 4)
hp2
dev.off()


Rdata_filename = paste0(fname_prefix_R, "_updated_cds_with_modules_SELECTION.RData")
save(meta,cds,cds_cardio,meta_cardio,agg_mat,agg_mat_test,modules,cardio_cds_pr_test_res,genes_mod_pseudotime,gene_module_df,cell_group_df_test,select_modules,module_final_selection,select_modules_FINAL,
     file = Rdata_filename) 

side=read.csv(paste0(fname_Fig5_input_folder,"/clone_side.csv"))

clones=side$clone
side=side%>%distinct(clone,side)
meta=meta%>%left_join(side, by="clone")#5858
meta$side=meta$side%>% replace_na("unknown")
cds@colData$side=meta$side

Rdata_filename = paste0(fname_prefix_R, "_updated_cds_with_modules_SELECTION_side.RData")
save(meta,cds,cds_cardio,meta_cardio,agg_mat,agg_mat_test,modules,cardio_cds_pr_test_res,genes_mod_pseudotime,gene_module_df,cell_group_df_test,select_modules,module_final_selection,select_modules_FINAL,
     file = Rdata_filename)

```


###plot genes in pseudotime

```{r}
maturation=c("TTN","TNNT2", "TNNC1")
cardiac_lineage_cds <- cds[rowData(cds)$gene_short_name %in% maturation,]

plot_genes_in_pseudotime(cardiac_lineage_cds,
                         color_cells_by="monocle_clusters",
                         min_expr=0.5)

ggsave(filename = paste0(fname_prefix_dotplot, "_", "_maturation_pseudotime.pdf"),
     width = 8, height = 8)
```



